---
title: "Applying PhenoMapR to facioscapulohumeral muscular dystrophy (FSHD) datasets"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_collapsed: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  out.width = "100%"
  # cache = TRUE,
  # cache.path = "cache/fshd-vignette/"
)
# Always run this vignette when knitting.
vignette_run <- TRUE

# Seurat's FindNeighbors / FindClusters use future::future_lapply. A multisession
# or multicore plan (e.g. from .Rprofile) can trigger FutureLaunchError when
# workers see a different 'future' version than the parent, or when unloading
# 'future' fails because Seurat imports it. Sequential plan avoids forked workers.
if (requireNamespace("future", quietly = TRUE)) {
  future::plan("sequential", gc = FALSE)
}
```

## Overview

This vignette shows how to use **`PhenoMapR`** to mirror the **FSHD (facioscapulohumeral muscular dystrophy)** analysis in **Scissor** ([Sun *et al.*, *Nat. Biotechnol.* 2022](https://doi.org/10.1038/s41587-021-01091-3); [Nature article](https://www.nature.com/articles/s41587-021-01091-3)), where bulk expression and a binary phenotype are integrated with single-cell data to surface phenotype-associated cells. **Figure 5** of that paper applies Scissor to FSHD using public GEO data.

We follow the paper’s description of the GEO inputs:

- **FSHD scRNA-seq ([s](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE122873))** — 7,047 cells from muscle biopsy–derived myocyte cultures (four FSHD patients, two controls); **148 cells** with high mitochondrial expression (≥10% of total) are removed, leaving **6,899** cells for analysis (Scissor’s reported numbers).
- **FSHD bulk ([GSE140261](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE140261))** — TPM from **27** follow-up FSHD biopsies and **eight** control biopsies in the processed **year 2** supplementary files; Ensembl IDs are collapsed to **gene symbols** by **summing** expression when multiple Ensembl IDs map to the same symbol (here we use the **mean TPM** across duplicate Ensembl IDs after mapping, which is equivalent for proportional summaries).

**PhenoMapR** workflow:

1. Derive a **custom z-score reference** from bulk TPM with **`derive_reference_from_bulk()`** (FSHD vs. control).
2. Score all single cells with **`PhenoMap()`** and visualize on **UMAP**.
3. Define **adverse / favorable** tails with **`define_phenotype_groups()`** and run **`find_phenotype_markers()`** (`marker_scope = "phenotype_groups"` for global markers).
4. Draw a **global marker heatmap** with **`plot_phenotype_markers(..., heatmap_type = "global")`**.
5. **Validation** — apply the **GSE140261**-derived bulk signature (`ref_fshd`) to three independent GEO cohorts (**GSE115650**, **GSE56787** muscle biopsies only, **GSE26852** FSHD vs. normal controls) and test **FSHD vs. control** stratification with **`PhenoMap()`** (Section 6).

## Packages

```{r libraries}
suppressPackageStartupMessages({
  library(PhenoMapR)
  library(ggplot2)
  library(dplyr)
    library(Seurat)
    library(AnnotationDbi)
    library(org.Hs.eg.db)
  })
```

## 1. Bulk TPM and phenotype (GSE140261)

Processed **year 2** supplementary files ([GEO FTP](https://ftp.ncbi.nlm.nih.gov/geo/series/GSE140nnn/GSE140261/suppl/)):

- `GSE140261_year2_TPM.csv.gz` — TPM with **Ensembl IDs** (with version) as rows.
- `GSE140261_year2_column_data.csv.gz` — metadata; we use **`pheno_type`** (`FSHD` vs `Control`).

```{r bulk-download}
ftp_base <- "https://ftp.ncbi.nlm.nih.gov/geo/series/GSE140nnn/GSE140261/suppl/"
tpm_url <- paste0(ftp_base, "GSE140261_year2_TPM.csv.gz")
meta_url <- paste0(ftp_base, "GSE140261_year2_column_data.csv.gz")

#' Try several strategies: NCBI HTTPS often fails with default download.file (proxy/SSL/RStudio).
#' Falls back to GEOquery supplementary download, then system curl if available.
download_geo_file <- function(url, dest, geo_acc = NULL, supp_filename = NULL) {
  if (file.exists(dest) && file.info(dest)$size > 1000L) {
    return(normalizePath(dest, winslash = "/"))
  }
  dir.create(dirname(dest), showWarnings = FALSE, recursive = TRUE)

  # 1) curl package: most reliable for HTTPS (often works when download.file / RStudio fails)
  if (requireNamespace("curl", quietly = TRUE)) {
    st <- tryCatch(
      {
        curl::curl_download(url, dest, quiet = TRUE)
        0L
      },
      error = function(e) 1L
    )
    if (st == 0L && file.exists(dest) && file.info(dest)$size > 1000L) {
      return(normalizePath(dest, winslash = "/"))
    }
  }

  try_download_file <- function() {
    methods <- c("libcurl", "wininet", "auto")
    for (m in methods) {
      if (m == "wininet" && .Platform$OS.type != "windows") next
      st <- tryCatch(
        download.file(url, dest, mode = "wb", quiet = TRUE, method = m),
        error = function(e) 1L
      )
      if (isTRUE(st == 0L) && file.exists(dest) && file.info(dest)$size > 1000L) {
        return(TRUE)
      }
    }
    FALSE
  }

  if (try_download_file()) {
    return(normalizePath(dest, winslash = "/"))
  }

  if (!is.null(geo_acc) && !is.null(supp_filename) && requireNamespace("GEOquery", quietly = TRUE)) {
    base <- tempfile("geo_supp")
    dir.create(base)
    tryCatch(
      suppressMessages(
        GEOquery::getGEOSuppFiles(geo_acc, makeDirectory = TRUE, baseDir = base)
      ),
      error = function(e) invisible(NULL)
    )
    hits <- list.files(
      base,
      pattern = supp_filename,
      fixed = TRUE,
      recursive = TRUE,
      full.names = TRUE
    )
    if (length(hits) >= 1L && file.copy(hits[1], dest, overwrite = TRUE)) {
      if (file.exists(dest) && file.info(dest)$size > 1000L) {
        return(normalizePath(dest, winslash = "/"))
      }
    }
  }

  curl_bin <- Sys.which("curl")
  if (nzchar(curl_bin)) {
    st <- system2(curl_bin, c("-fsSL", "-o", dest, shQuote(url)), stdout = FALSE, stderr = FALSE)
    if (st == 0L && file.exists(dest) && file.info(dest)$size > 1000L) {
      return(normalizePath(dest, winslash = "/"))
    }
  }

  stop(
    "Download failed for:\n  ", url,
    "\nTry: install.packages('curl'); BiocManager::install('GEOquery'); ",
    "options(download.file.method = 'libcurl'); ",
    "or download the file in a browser and set PHENOMAPR_GSE122873_TAR to the local path."
  )
}

download_if_needed <- function(url, dest, geo_acc = NULL, supp_filename = basename(url)) {
  download_geo_file(url, dest, geo_acc = geo_acc, supp_filename = supp_filename)
}

#' Collapse Ensembl rows to HGNC symbols by mean TPM (duplicate Ensembl → one symbol).
collapse_ensembl_tpm_to_symbol <- function(tpm_ens_rows_samples_cols) {
  ens <- rownames(tpm_ens_rows_samples_cols)
  ens_base <- sub("\\.[0-9]+$", "", ens)
  sym <- AnnotationDbi::mapIds(
    org.Hs.eg.db,
    keys = ens_base,
    column = "SYMBOL",
    keytype = "ENSEMBL",
    multiVals = "first"
  )
  df <- as.data.frame(tpm_ens_rows_samples_cols)
  df$gene_symbol <- sym
  df <- df[!is.na(df$gene_symbol) & nzchar(df$gene_symbol), , drop = FALSE]
  sample_cols <- setdiff(names(df), "gene_symbol")
  agg <- stats::aggregate(
    df[, sample_cols, drop = FALSE],
    by = list(gene_symbol = df$gene_symbol),
    FUN = mean,
    na.rm = TRUE
  )
  mat <- as.matrix(agg[, -1, drop = FALSE])
  rownames(mat) <- agg$gene_symbol
  mat
}


  td <- tempdir()
  tpm_path <- download_if_needed(
    tpm_url,
    file.path(td, "GSE140261_year2_TPM.csv.gz"),
    geo_acc = "GSE140261",
    supp_filename = "GSE140261_year2_TPM.csv.gz"
  )
  meta_path <- download_if_needed(
    meta_url,
    file.path(td, "GSE140261_year2_column_data.csv.gz"),
    geo_acc = "GSE140261",
    supp_filename = "GSE140261_year2_column_data.csv.gz"
  )

  tpm_wide <- read.csv(gzfile(tpm_path), row.names = 1, check.names = FALSE, stringsAsFactors = FALSE)
  tpm_sym <- collapse_ensembl_tpm_to_symbol(tpm_wide)

  coldata <- read.csv(gzfile(meta_path), stringsAsFactors = FALSE, check.names = FALSE)
  coldata <- coldata[tolower(coldata$pheno_type) %in% c("fshd", "control"), , drop = FALSE]
  coldata$group <- ifelse(tolower(coldata$pheno_type) == "fshd", "FSHD", "Control")
  coldata$sample_id <- coldata$sample_name

  use_samples <- intersect(colnames(tpm_sym), coldata$sample_id)
  tpm_sym <- tpm_sym[, use_samples, drop = FALSE]
  pheno_bulk <- coldata[match(use_samples, coldata$sample_id), c("sample_id", "group")]
  pheno_bulk$group <- factor(pheno_bulk$group, levels = c("FSHD", "Control"))

  ref_fshd <- derive_reference_from_bulk(
    bulk_expression = tpm_sym,
    phenotype = pheno_bulk,
    sample_id_column = "sample_id",
    phenotype_column = "group",
    phenotype_type = "binary",
    normalize = FALSE,
    binary_positive_reference = "first",
    verbose = TRUE
  )
```

With **`levels = c("FSHD", "Control")`** and **`binary_positive_reference = "first"`**, genes with **positive** z-scores are **higher in FSHD** bulk samples.

```{r bulk-ref-plot, fig.width=6, fig.height=3.5}
if (vignette_run) {
  plot_score_distribution(ref_fshd, main = "GSE140261: FSHD vs control (gene-level z-scores)")
}
```

## 2. Single-cell data (GSE122873) and mitochondrial QC

[GSE122873](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE122873) provides **`GSE122873_RAW.tar`** with six **`GSM*_*.txt.gz`** tables (myocyte cultures: four FSHD + two control). Each table is **genes × cells** with a header row of cell barcodes ([Kendall *et al.*, *Genome Res.* 2019](https://doi.org/10.1101/gr.243734.118); [PMC6423425](https://pmc.ncbi.nlm.nih.gov/articles/PMC6423425/)). **Gene identifiers are Ensembl IDs** (with optional version suffix); **`PhenoMapR`** matches the custom reference by **HUGO symbols**, so we **map Ensembl → symbol** and **sum counts** per symbol **before** building the Seurat object.

**If downloads fail** (common behind strict firewalls or in RStudio: `download.file` to NCBI can error), install **`GEOquery`** (`BiocManager::install("GEOquery")`) so the code can fetch supplementary files via Bioconductor, or set **`PHENOMAPR_GSE122873_TAR`** to the **full path** to a manually downloaded `GSE122873_RAW.tar`.

```{r sc-download}
sc_tar_url <- "https://ftp.ncbi.nlm.nih.gov/geo/series/GSE122nnn/GSE122873/suppl/GSE122873_RAW.tar"

read_gse122873_sample <- function(path) {
  m <- as.matrix(read.delim(gzfile(path), row.names = 1, check.names = FALSE))
  storage.mode(m) <- "double"
  m
}

merge_gse122873 <- function(dir_untar) {
  fs <- list.files(dir_untar, pattern = "^GSM.*\\.txt\\.gz$", full.names = TRUE)
  if (length(fs) < 1L) stop("No GSM*.txt.gz after extracting GSE122873_RAW.tar")
  mats <- lapply(fs, read_gse122873_sample)
  genes <- Reduce(intersect, lapply(mats, rownames))
  combined <- do.call(cbind, lapply(mats, function(m) m[genes, , drop = FALSE]))
  colnames(combined) <- make.unique(colnames(combined), sep = "_")
  sn <- basename(fs)
  lab <- sub("\\.txt\\.gz$", "", sn)
  lab <- sub("^GSM[0-9]+_", "", lab)
  rep_lab <- rep(lab, times = vapply(mats, ncol, 1L))
  list(counts = combined, donor_batch = rep_lab)
}

#' Genes x cells count matrix: Ensembl rownames -> HUGO symbols (sum counts per symbol).
#' Rows already looking like symbols (not ENSG...) are kept as-is.
collapse_counts_ensembl_to_hugo <- function(counts_mat) {
  if (!requireNamespace("org.Hs.eg.db", quietly = TRUE)) {
    stop("Package 'org.Hs.eg.db' is required to map Ensembl IDs to symbols.")
  }
  x <- as.matrix(counts_mat)
  genes <- rownames(x)
  ens_base <- sub("\\.[0-9]+$", "", genes)
  is_ens <- grepl("^ENSG[0-9]+", ens_base, ignore.case = TRUE)
  sym <- genes
  if (any(is_ens)) {
    mapped <- AnnotationDbi::mapIds(
      org.Hs.eg.db,
      keys = ens_base[is_ens],
      column = "SYMBOL",
      keytype = "ENSEMBL",
      multiVals = "first"
    )
    sym[is_ens] <- as.character(mapped)
  }
  ok <- !is.na(sym) & nzchar(sym) & sym != "NA"
  x <- x[ok, , drop = FALSE]
  sym <- sym[ok]
  if (length(unique(sym)) < length(sym)) {
    x <- rowsum(x, group = sym, reorder = FALSE)
  } else {
    rownames(x) <- sym
  }
  x
}

  td_sc <- tempfile("gse122873")
  dir.create(td_sc, showWarnings = FALSE)
  tar_path <- file.path(td_sc, "GSE122873_RAW.tar")
  env_tar <- Sys.getenv("PHENOMAPR_GSE122873_TAR", unset = "")
  if (nzchar(env_tar) && file.exists(env_tar)) {
    file.copy(env_tar, tar_path, overwrite = TRUE)
  } else {
    download_geo_file(
      sc_tar_url,
      tar_path,
      geo_acc = "GSE122873",
      supp_filename = "GSE122873_RAW.tar"
    )
  }
  untar(tar_path, exdir = td_sc)
  sc_list <- merge_gse122873(td_sc)
  counts <- collapse_counts_ensembl_to_hugo(sc_list$counts)
  meta_cells <- data.frame(
    cell_id = colnames(counts),
    donor_batch = sc_list$donor_batch,
    stringsAsFactors = FALSE
  )
  meta_cells$disease <- ifelse(grepl("^CTRL", meta_cells$donor_batch), "Control", "FSHD")
  rownames(meta_cells) <- meta_cells$cell_id

  seu <- CreateSeuratObject(counts = counts, meta.data = meta_cells)
  seu[["percent.mt"]] <- PercentageFeatureSet(seu, pattern = "^MT-")
  seu <- subset(seu, subset = percent.mt < 10)
  seu <- NormalizeData(seu, verbose = FALSE)
  seu <- FindVariableFeatures(seu, verbose = FALSE)
  all.genes <- rownames(seu)
seu <- ScaleData(seu, features = all.genes)
  seu <- RunPCA(seu, verbose = FALSE)

# Cluster the cells and run UMAP #####
seu <- FindNeighbors(seu, dims = 1:10)
seu <- FindClusters(seu, resolution = 0.5)
  seu <- RunUMAP(seu, dims = 1:30, verbose = FALSE)
  seu.markers <- FindAllMarkers(object = seu, only.pos = TRUE)
```

After filtering, you should obtain **about 6,899** cells (exact counts can differ slightly with software versions).

## 3. Score single cells with the custom reference

```{r score-sc}
if (vignette_run) {
  scores_sc <- PhenoMap(
    seu,
    reference = ref_fshd,
    assay = "RNA",
    slot = "data",
    z_score_cutoff = 2,
    verbose = FALSE
  )
  seu <- add_scores_to_seurat(seu, scores_sc)
  score_col <- grep("weighted_sum_score", colnames(scores_sc), value = TRUE)[1]
}
```

```{r umap-score, fig.width=12, fig.height=4.5}
if (vignette_run) {
  pal <- c(Control = "#2166AC", FSHD = "#B2182B")
  p1 <- DimPlot(seu, group.by = "disease", cols = pal) +
    ggtitle("GSE122873: culture (Control vs FSHD)")
  plot_data_sc <- data.frame(
    disease = seu$disease,
    score = seu@meta.data[[score_col]],
    stringsAsFactors = FALSE
  )
  p_violin <- ggplot(plot_data_sc, aes(x = disease, y = score, fill = disease)) +
    geom_violin(alpha = 0.6) +
    geom_jitter(width = 0.1, alpha = 0.15, size = 0.3) +
    scale_fill_manual(values = pal) +
    theme_minimal(base_size = 12) +
    theme(legend.position = "none") +
    labs(
      title = "Custom FSHD bulk signature (single-cell scores)",
      y = score_col,
      x = NULL
    )
  p2 <- FeaturePlot(seu, features = score_col, order = TRUE) +
    scale_color_gradientn(colors = c("#2166AC", "#F7F7F7", "#B2182B")) +
    ggtitle("PhenoMap score on UMAP")
  if (requireNamespace("patchwork", quietly = TRUE)) {
    suppressPackageStartupMessages(library(patchwork))
    print(p1 + p_violin + p2 + patchwork::plot_layout(widths = c(1, 1, 1)))
  } else {
    print(p1)
    print(p_violin)
    print(p2)
  }
}
```

## Cell type annotations (CyteTypeR results)

This dataset does not provide cell type labels. Here we use pre-computed **CyteTypeR** annotations (saved alongside this vignette) and attach them to the Seurat object metadata as `cell_type` and `coarse_cell_type`. The next figure shows those labels on the **UMAP** (`label = TRUE`).

```{r cytetype-annotations}
if (vignette_run) {
  ann_path <- "cytetype_annotations_ce748d9d-6191-42bc-a212-4bace7e97e4b.csv"
  if (!file.exists(ann_path)) {
    stop("Missing CyteTypeR annotations CSV: ", ann_path)
  }

  ann <- utils::read.csv(ann_path, stringsAsFactors = FALSE, check.names = FALSE)
  req_cols <- c("cluster_id", "cell_type", "coarse_cell_type")
  if (!all(req_cols %in% names(ann))) {
    stop("CyteTypeR annotations missing required columns: ", paste(setdiff(req_cols, names(ann)), collapse = ", "))
  }

  # Map Seurat clusters to CyteTypeR cluster_id.
  # Convention used when these results were generated: cluster_id = as.integer(seurat_clusters) + 1
  cl <- as.character(seu$seurat_clusters)
  if (anyNA(cl)) stop("Missing seurat_clusters in Seurat metadata.")
  if (!all(grepl("^[0-9]+$", cl))) {
    stop("Expected numeric seurat_clusters for CyteTypeR mapping; got non-numeric cluster labels.")
  }
  seu$cluster_id <- as.integer(cl) + 1L

  # Join per-cell annotations
  ann_sub <- ann[, req_cols, drop = FALSE]
  key <- match(seu$cluster_id, ann_sub$cluster_id)
  seu$cell_type <- ann_sub$cell_type[key]
  seu$coarse_cell_type <- ann_sub$coarse_cell_type[key]

  # Basic sanity check
  if (all(is.na(seu$cell_type))) {
    stop("CyteTypeR mapping produced all-NA cell_type; check cluster_id alignment.")
  }
}
```

```{r cytetype-umap, fig.width=9, fig.height=5}
if (vignette_run && "cell_type" %in% colnames(seu@meta.data)) {
  ct_levels <- sort(unique(as.character(stats::na.omit(seu$cell_type))))
  pal_ct <- PhenoMapR::get_celltype_palette(ct_levels)
  DimPlot(
    seu,
    group.by = "cell_type",
    cols = pal_ct,
    label = TRUE,
    repel = TRUE,
    label.size = 3.2
  ) +
    ggtitle("CyteTypeR: cell types on UMAP")
}
```

## 4. Phenotype tails and marker genes (single-cell)

We label **Most Adverse** / **Most Favorable** / **Other** tails from the custom score (**higher** = more FSHD-like under this reference), then run:

- Global markers (`marker_scope = "phenotype_groups"`)
- Cell-type-specific markers (`marker_scope = "cell_type_specific"`) using **CyteTypeR** `cell_type`
- Coarse cell-type-specific markers using **CyteTypeR** `coarse_cell_type`

```{r markers-sc}
if (vignette_run) {
  groups <- define_phenotype_groups(scores_sc, percentile = 0.05, score_columns = score_col)
  group_col <- grep("^phenotype_group_", names(groups), value = TRUE)[1]

  mk <- find_phenotype_markers(
    seu,
    group_labels = groups,
    group_column = group_col,
    cell_id_column = "cell_id",
    marker_scope = "phenotype_groups",
    assay = "RNA",
    slot = "data",
    verbose = FALSE
  )
  message("Adverse markers (top 8):")
  print(utils::head(mk$adverse_markers[order(mk$adverse_markers$p_adj), ], 8))
  message("Favorable markers (top 8):")
  print(utils::head(mk$favorable_markers[order(mk$favorable_markers$p_adj), ], 8))

  # Add CyteTypeR labels into group_labels for cell-type-specific contrasts
  groups_ct <- as.data.frame(groups)
  groups_ct$cell_type <- seu$cell_type[match(groups_ct$cell_id, colnames(seu))]
  groups_ct$coarse_cell_type <- seu$coarse_cell_type[match(groups_ct$cell_id, colnames(seu))]

  mk_celltype <- tryCatch(
    find_phenotype_markers(
      seu,
      group_labels = groups_ct,
      group_column = group_col,
      cell_id_column = "cell_id",
      marker_scope = "cell_type_specific",
      cell_type_column = "cell_type",
      assay = "RNA",
      slot = "data",
      verbose = FALSE
    ),
    error = function(e) {
      message("cell_type_specific markers (cell_type) failed: ", conditionMessage(e))
      NULL
    }
  )

  mk_coarse <- tryCatch(
    find_phenotype_markers(
      seu,
      group_labels = groups_ct,
      group_column = group_col,
      cell_id_column = "cell_id",
      marker_scope = "cell_type_specific",
      cell_type_column = "coarse_cell_type",
      assay = "RNA",
      slot = "data",
      verbose = FALSE
    ),
    error = function(e) {
      message("cell_type_specific markers (coarse_cell_type) failed: ", conditionMessage(e))
      NULL
    }
  )
}
```

## 5. Phenotype marker heatmap (`plot_phenotype_markers`)

After **`find_phenotype_markers()`**, **`plot_phenotype_markers()`** builds a **ComplexHeatmap** of marker expression with column annotations for **phenotype group**, **cell type**, and the continuous **PhenoMapR score**. Install **`ComplexHeatmap`** and **`circlize`** if needed.

```{r plot-phenotype-markers-fshd, fig.align='center', fig.width=13, fig.height=9}
if (vignette_run && exists("mk") && exists("groups") && exists("group_col")) {
  if (!requireNamespace("ComplexHeatmap", quietly = TRUE) || !requireNamespace("circlize", quietly = TRUE)) {
    message("Install packages 'ComplexHeatmap' and 'circlize' to draw plot_phenotype_markers().")
  } else {
    expr_mat <- tryCatch(
      Seurat::GetAssayData(seu, layer = "data", assay = "RNA"),
      error = function(e) Seurat::GetAssayData(seu, slot = "data", assay = "RNA")
    )
    meta_hm <- as.data.frame(seu@meta.data)
    meta_hm$cell_id <- colnames(seu)
    meta_hm[[group_col]] <- groups[[group_col]][match(meta_hm$cell_id, groups$cell_id)]
    pal_ct <- PhenoMapR::get_celltype_palette(levels(factor(meta_hm$coarse_cell_type)))

    plot_phenotype_markers(
      markers = mk,
      expr_mat = expr_mat,
      meta = meta_hm,
      cell_id_col = "cell_id",
      group_col = group_col,
      score_col = score_col,
      celltype_col = "coarse_cell_type",
      celltype_palette = pal_ct,
      heatmap_type = "global",
      top_n_markers = 20L,
      n_mark_labels = 5L,
      p_adj_threshold = 0.05,
      column_title = "FSHD bulk-derived signature: single-cell marker heatmap"
    )

    if (exists("mk_celltype") && !is.null(mk_celltype) && (nrow(mk_celltype$adverse_markers) > 0 || nrow(mk_celltype$favorable_markers) > 0)) {
      plot_phenotype_markers(
        markers = mk_celltype,
        expr_mat = expr_mat,
        meta = meta_hm,
        cell_id_col = "cell_id",
        group_col = group_col,
        score_col = score_col,
        celltype_col = "cell_type",
        celltype_palette = PhenoMapR::get_celltype_palette(levels(factor(meta_hm$cell_type))),
        heatmap_type = "cell_type_specific",
        top_n_markers = 15L,
        n_mark_labels = 3L,
        p_adj_threshold = 0.05,
        column_title = "Cell-type-specific phenotype markers (CyteTypeR cell_type)"
      )
    }
  }
}
```

## 6. Independent validation cohorts (GEO)

To test whether the **GSE140261** bulk phenotype signature (`ref_fshd`) separates **FSHD** from **normal / control** muscle outside the training set, we follow the validation design described in the Scissor FSHD analysis: three independent GEO series were downloaded (**[GSE115650](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE115650)**, **[GSE56787](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE56787)**, **[GSE26852](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE26852)**).

- **GSE115650** includes **34** FSHD subjects and **nine** control muscle biopsies. Supplementary **`GSE115650_biopsy_FPKM.csv.gz`** provides **FPKM**; we convert **FPKM → TPM** per sample (divide each sample’s FPKM by the sum of FPKMs and scale to \(10^6\)), then map **Ensembl** IDs to **gene symbols** (mean TPM when several Ensembl IDs map to the same symbol), matching §1.
- **GSE56787** contains **24** muscle biopsies (**15** FSHD, **nine** control) plus cultured **myotube / myoblast** samples. We use **`GSE56787_biopsy.gene.counts.csv.gz`** (biopsy counts only) and assign **FSHD** vs **Control** from **`source_name_ch1`**. Counts are **summed** per symbol, then **CPM** (counts per million per sample) is used as library-size–normalized expression for **`PhenoMap()`** (analogous to using proportional abundance for signature weighting).
- **GSE26852** includes FSHD, inflammatory myopathies, dysferlinopathy, and controls. We keep **normal muscle** and **FSHD** samples only (`characteristics_ch1.3`: “Normal Muscle” or “FSHD T2-STIR”), discarding other disease groups. The series matrix reports **Illumina** probe IDs; we collapse **`fData` `Symbol`** to **one row per symbol** (mean intensity), then **scale each sample** to sum \(10^6\) so scores are comparable in scale to TPM-like proportions.

In each cohort, samples are scored with **`PhenoMap(expression = genes × samples, reference = ref_fshd, z_score_cutoff = 2)`**. **Higher** scores indicate stronger alignment with genes that were **higher in FSHD** in **GSE140261** (given `binary_positive_reference = "first"` in §1). Group separation is summarized with **Wilcoxon** tests and **violin plots**.

**Dependencies:** **`GEOquery`** (`BiocManager::install("GEOquery")`) and network access to NCBI GEO. If downloads fail, install **`curl`** or fetch supplementary files manually.

```{r validation-helpers}
#' FPKM columns to TPM: TPM_ij = FPKM_ij / sum_i(FPKM_ij) * 1e6
fpkm_to_tpm <- function(fpkm) {
  fpkm <- as.matrix(fpkm)
  cs <- colSums(fpkm, na.rm = TRUE)
  cs[cs <= 0] <- NA_real_
  sweep(fpkm, 2L, cs, "/") * 1e6
}

#' Collapse Ensembl gene-level counts to symbols by summing (then e.g. CPM).
collapse_ensembl_counts_to_symbol_sum <- function(cnt) {
  ens <- rownames(cnt)
  ens_base <- sub("\\.[0-9]+$", "", ens)
  sym <- AnnotationDbi::mapIds(
    org.Hs.eg.db,
    keys = ens_base,
    column = "SYMBOL",
    keytype = "ENSEMBL",
    multiVals = "first"
  )
  df <- as.data.frame(cnt)
  df$gene_symbol <- sym
  df <- df[!is.na(df$gene_symbol) & nzchar(df$gene_symbol), , drop = FALSE]
  sample_cols <- setdiff(names(df), "gene_symbol")
  agg <- stats::aggregate(
    df[, sample_cols, drop = FALSE],
    by = list(gene_symbol = df$gene_symbol),
    FUN = sum,
    na.rm = TRUE
  )
  mat <- as.matrix(agg[, -1, drop = FALSE])
  rownames(mat) <- agg$gene_symbol
  mat
}
```

```{r validation-geo-cohorts, fig.width = 9, fig.height = 3.5}
validation_plot_df <- NULL
validation_wilcox <- NULL

if (vignette_run && exists("ref_fshd") && requireNamespace("GEOquery", quietly = TRUE)) {
  val_base <- tempdir()
  val_score_col <- paste0("weighted_sum_score_", colnames(ref_fshd)[1])

  tryCatch(
    {
      ## --- GSE115650: FPKM -> TPM, Ensembl -> symbol ---------------------------
      suppressMessages(
        GEOquery::getGEOSuppFiles("GSE115650", makeDirectory = TRUE, baseDir = val_base)
      )
      fp_115650 <- file.path(val_base, "GSE115650", "GSE115650_biopsy_FPKM.csv.gz")
      if (file.exists(fp_115650)) {
        fpkm_115650 <- read.csv(
          gzfile(fp_115650),
          row.names = 1,
          check.names = FALSE,
          stringsAsFactors = FALSE
        )
        tpm_115650 <- fpkm_to_tpm(fpkm_115650)
        tpm_sym_115650 <- collapse_ensembl_tpm_to_symbol(tpm_115650)
        g_115650 <- suppressMessages(GEOquery::getGEO("GSE115650", GSEMatrix = TRUE)[[1]])
        pd_115650 <- Biobase::pData(g_115650)
        st <- tolower(pd_115650$characteristics_ch1)
        is_fshd <- grepl("fshd|facioscapulohumeral", st)
        is_ctrl <- grepl("subject status: control", st)
        pd_115650$group <- NA_character_
        pd_115650$group[is_fshd] <- "FSHD"
        pd_115650$group[is_ctrl] <- "Control"
        pd_115650$title <- as.character(pd_115650$title)
        m <- match(colnames(tpm_sym_115650), pd_115650$title)
        if (!anyNA(m)) {
          pd_115650 <- pd_115650[m, , drop = FALSE]
          pd_115650$group <- factor(pd_115650$group, levels = c("FSHD", "Control"))
          colnames(tpm_sym_115650) <- pd_115650$title
          sc_115650 <- PhenoMap(
            expression = tpm_sym_115650,
            reference = ref_fshd,
            z_score_cutoff = 2,
            verbose = FALSE
          )
          x <- sc_115650[rownames(pd_115650), val_score_col, drop = TRUE]
          validation_wilcox[["GSE115650"]] <- stats::wilcox.test(x ~ pd_115650$group)
          validation_plot_df <- rbind(
            validation_plot_df,
            data.frame(
              cohort = "GSE115650",
              group = pd_115650$group,
              score = x,
              stringsAsFactors = FALSE
            )
          )
        }
      }

      ## --- GSE56787: biopsy gene counts -> CPM -------------------------------
      suppressMessages(
        GEOquery::getGEOSuppFiles("GSE56787", makeDirectory = TRUE, baseDir = val_base)
      )
      fp_56787 <- file.path(val_base, "GSE56787", "GSE56787_biopsy.gene.counts.csv.gz")
      if (file.exists(fp_56787)) {
        cnt_56787 <- read.csv(
          gzfile(fp_56787),
          row.names = 1,
          check.names = FALSE,
          stringsAsFactors = FALSE
        )
        cnt_sym_56787 <- collapse_ensembl_counts_to_symbol_sum(cnt_56787)
        lib <- colSums(cnt_sym_56787)
        cpm_56787 <- sweep(cnt_sym_56787, 2, lib, "/") * 1e6
        g_56787 <- suppressMessages(GEOquery::getGEO("GSE56787", GSEMatrix = TRUE)[[1]])
        pd_56787 <- Biobase::pData(g_56787)
        biopsy <- grepl("Muscle biopsy", pd_56787$characteristics_ch1.1)
        pd_56787 <- pd_56787[biopsy, , drop = FALSE]
        pd_56787$group <- ifelse(grepl("FSHD", pd_56787$source_name_ch1), "FSHD", "Control")
        pd_56787$group <- factor(pd_56787$group, levels = c("FSHD", "Control"))
        m2 <- match(colnames(cpm_56787), pd_56787$title)
        if (!anyNA(m2)) {
          pd_56787 <- pd_56787[m2, , drop = FALSE]
          colnames(cpm_56787) <- pd_56787$title
          sc_56787 <- PhenoMap(
            expression = cpm_56787,
            reference = ref_fshd,
            z_score_cutoff = 2,
            verbose = FALSE
          )
          x2 <- sc_56787[rownames(pd_56787), val_score_col, drop = TRUE]
          validation_wilcox[["GSE56787"]] <- stats::wilcox.test(x2 ~ pd_56787$group)
          validation_plot_df <- rbind(
            validation_plot_df,
            data.frame(
              cohort = "GSE56787",
              group = pd_56787$group,
              score = x2,
              stringsAsFactors = FALSE
            )
          )
        }
      }

      ## --- GSE26852: Illumina -> symbol, FSHD vs normal ----------------------
      g_26852 <- suppressMessages(GEOquery::getGEO("GSE26852", GSEMatrix = TRUE)[[1]])
      pd_26852 <- Biobase::pData(g_26852)
      ds <- pd_26852$characteristics_ch1.3
      keep_26852 <- grepl("Normal Muscle", ds) | grepl("FSHD T2-STIR", ds)
      pd_26852 <- pd_26852[keep_26852, , drop = FALSE]
      pd_26852$group <- ifelse(grepl("FSHD", pd_26852$characteristics_ch1.3), "FSHD", "Control")
      pd_26852$group <- factor(pd_26852$group, levels = c("FSHD", "Control"))
      fd_26852 <- Biobase::fData(g_26852)
      e_26852 <- Biobase::exprs(g_26852)[, rownames(pd_26852), drop = FALSE]
      sym_26852 <- fd_26852$Symbol
      ok_26852 <- !is.na(sym_26852) & nzchar(sym_26852)
      e_26852 <- e_26852[ok_26852, , drop = FALSE]
      sym_26852 <- sym_26852[ok_26852]
      dfl_26852 <- data.frame(sym = sym_26852, e_26852, check.names = FALSE)
      agg_26852 <- stats::aggregate(. ~ sym, data = dfl_26852, mean)
      rownames(agg_26852) <- agg_26852$sym
      mat_26852 <- as.matrix(agg_26852[, -1, drop = FALSE])
      colnames(mat_26852) <- colnames(e_26852)
      mat_26852 <- apply(mat_26852, 2, function(x) x / sum(x) * 1e6)
      sc_26852 <- PhenoMap(
        expression = mat_26852,
        reference = ref_fshd,
        z_score_cutoff = 2,
        verbose = FALSE
      )
      x3 <- sc_26852[rownames(pd_26852), val_score_col, drop = TRUE]
      validation_wilcox[["GSE26852"]] <- stats::wilcox.test(x3 ~ pd_26852$group)
      validation_plot_df <- rbind(
        validation_plot_df,
        data.frame(
          cohort = "GSE26852",
          group = pd_26852$group,
          score = x3,
          stringsAsFactors = FALSE
        )
      )

      if (!is.null(validation_plot_df) && nrow(validation_plot_df) > 0) {
        validation_plot_df$cohort <- factor(
          validation_plot_df$cohort,
          levels = c("GSE115650", "GSE56787", "GSE26852")
        )
        ggplot(validation_plot_df, aes(x = group, y = score, fill = group)) +
          geom_violin(trim = FALSE, alpha = 0.35, color = NA) +
          geom_jitter(width = 0.08, height = 0, size = 0.9, alpha = 0.85) +
          facet_wrap(~cohort, scales = "free_y", nrow = 1) +
          labs(
            title = "FSHD bulk-derived PhenoMap score: independent GEO cohorts",
            x = NULL,
            y = "PhenoMap score (ref = GSE140261)"
          ) +
          theme_bw() +
          theme(legend.position = "none")
      }
    },
    error = function(e) {
      message("Validation cohort download or scoring failed: ", conditionMessage(e))
    }
  )
} else if (vignette_run && !requireNamespace("GEOquery", quietly = TRUE)) {
  message("Install package 'GEOquery' to run Section 6 validation (BiocManager::install(\"GEOquery\")).")
}
```

```{r validation-wilcox-text, echo=FALSE, results='asis'}
if (
  vignette_run &&
  exists("validation_wilcox") &&
  length(validation_wilcox) > 0
) {
  for (nm in names(validation_wilcox)) {
    w <- validation_wilcox[[nm]]
    cat(sprintf(
      "**%s**: Wilcoxon rank-sum test, *p* = %s.\n\n",
      nm,
      format.pval(w$p.value, digits = 3)
    ))
  }
}
```

## Summary

- **Bulk reference** — **GSE140261** TPM, **FSHD vs control**, Ensembl→symbol collapse, **`derive_reference_from_bulk()`**.
- **Single-cell** — **GSE122873**, **<10% MT-** filter, **`PhenoMap()`** + **UMAP** + **`find_phenotype_markers()`** + **`plot_phenotype_markers()`** (global heatmap).
- **Independent validation (optional)** — **`GEOquery`** + supplementary tables: **GSE115650** (FPKM→TPM), **GSE56787** (biopsy counts→CPM; myotube/myoblast excluded), **GSE26852** (FSHD vs normal only); same **`ref_fshd`** scored with **`PhenoMap()`** (Section 6).
- **Relation to Scissor** — Scissor solves a **different** optimization (bulk–single-cell regression on a similarity matrix); **PhenoMapR** uses **gene-level** bulk contrasts and a **weighted sum** score on cells, but the **data sources and QC** can follow the same GEO-based workflow as in [Sun *et al.*](https://www.nature.com/articles/s41587-021-01091-3).

## References

<a id="ref-scissor"></a> Sun, D., Guan, X., Moran, A. E. *et al.* Identifying phenotype-associated subpopulations by integrating bulk and single-cell sequencing data. *Nat. Biotechnol.* **40**, 527–538 (2022). <https://doi.org/10.1038/s41587-021-01091-3>

<a id="ref-kendall"></a> Kendall, L. E. *et al.* Single-cell RNA sequencing in facioscapulohumeral muscular dystrophy disease etiology and development. *Genome Res.* **29**, 890–901 (2019). [GSE122873](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE122873)

## Session info

```{r session-info}
sessionInfo()
```
