---
title: "Applying PhenoMapR to facioscapulohumeral muscular dystrophy (FSHD) datasets"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_collapsed: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  out.width = "100%",
  cache = TRUE,
  cache.path = "cache/fshd-vignette/"
)
# vignette_run <- Sys.getenv("PHENOMAPR_FSHD_VIGNETTE", unset = "TRUE") != "FALSE"

# Seurat's FindNeighbors / FindClusters use future::future_lapply. A multisession
# or multicore plan (e.g. from .Rprofile) can trigger FutureLaunchError when
# workers see a different 'future' version than the parent, or when unloading
# 'future' fails because Seurat imports it. Sequential plan avoids forked workers.
if (requireNamespace("future", quietly = TRUE)) {
  future::plan("sequential", gc = FALSE)
}
```

## Overview

This vignette shows how to use **`PhenoMapR`** to mirror the **FSHD (facioscapulohumeral muscular dystrophy)** analysis in **Scissor** ([Sun *et al.*, *Nat. Biotechnol.* 2022](https://doi.org/10.1038/s41587-021-01091-3); [Nature article](https://www.nature.com/articles/s41587-021-01091-3)), where bulk expression and a binary phenotype are integrated with single-cell data to surface phenotype-associated cells. **Figure 5** of that paper applies Scissor to FSHD using public GEO data.

We follow the paper’s description of the GEO inputs:

- **FSHD scRNA-seq ([s](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE122873))** — 7,047 cells from muscle biopsy–derived myocyte cultures (four FSHD patients, two controls); **148 cells** with high mitochondrial expression (≥10% of total) are removed, leaving **6,899** cells for analysis (Scissor’s reported numbers).
- **FSHD bulk ([GSE140261](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE140261))** — TPM from **27** follow-up FSHD biopsies and **eight** control biopsies in the processed **year 2** supplementary files; Ensembl IDs are collapsed to **gene symbols** by **summing** expression when multiple Ensembl IDs map to the same symbol (here we use the **mean TPM** across duplicate Ensembl IDs after mapping, which is equivalent for proportional summaries).

**PhenoMapR** workflow:

1. Derive a **custom z-score reference** from bulk TPM with **`derive_reference_from_bulk()`** (FSHD vs. control).
2. Score all single cells with **`PhenoMap()`** and visualize on **UMAP**.
3. Define **adverse / favorable** tails with **`define_phenotype_groups()`** and run **`find_phenotype_markers()`** (`marker_scope = "phenotype_groups"` for global markers).
4. Draw a **global marker heatmap** with **`plot_phenotype_markers(..., heatmap_type = "global")`**.
5. **Validation** — sketch how to apply the same signature to independent GEO cohorts (**GSE115650**, **GSE56787** biopsies only, **GSE26852** FSHD vs. controls), as in the paper.

## Packages

```{r libraries}
suppressPackageStartupMessages({
  library(PhenoMapR)
  library(ggplot2)
  library(dplyr)
    library(Seurat)
    library(AnnotationDbi)
    library(org.Hs.eg.db)
  })
```

## 1. Bulk TPM and phenotype (GSE140261)

Processed **year 2** supplementary files ([GEO FTP](https://ftp.ncbi.nlm.nih.gov/geo/series/GSE140nnn/GSE140261/suppl/)):

- `GSE140261_year2_TPM.csv.gz` — TPM with **Ensembl IDs** (with version) as rows.
- `GSE140261_year2_column_data.csv.gz` — metadata; we use **`pheno_type`** (`FSHD` vs `Control`).

```{r bulk-download}
ftp_base <- "https://ftp.ncbi.nlm.nih.gov/geo/series/GSE140nnn/GSE140261/suppl/"
tpm_url <- paste0(ftp_base, "GSE140261_year2_TPM.csv.gz")
meta_url <- paste0(ftp_base, "GSE140261_year2_column_data.csv.gz")

#' Try several strategies: NCBI HTTPS often fails with default download.file (proxy/SSL/RStudio).
#' Falls back to GEOquery supplementary download, then system curl if available.
download_geo_file <- function(url, dest, geo_acc = NULL, supp_filename = NULL) {
  if (file.exists(dest) && file.info(dest)$size > 1000L) {
    return(normalizePath(dest, winslash = "/"))
  }
  dir.create(dirname(dest), showWarnings = FALSE, recursive = TRUE)

  # 1) curl package: most reliable for HTTPS (often works when download.file / RStudio fails)
  if (requireNamespace("curl", quietly = TRUE)) {
    st <- tryCatch(
      {
        curl::curl_download(url, dest, quiet = TRUE)
        0L
      },
      error = function(e) 1L
    )
    if (st == 0L && file.exists(dest) && file.info(dest)$size > 1000L) {
      return(normalizePath(dest, winslash = "/"))
    }
  }

  try_download_file <- function() {
    methods <- c("libcurl", "wininet", "auto")
    for (m in methods) {
      if (m == "wininet" && .Platform$OS.type != "windows") next
      st <- tryCatch(
        download.file(url, dest, mode = "wb", quiet = TRUE, method = m),
        error = function(e) 1L
      )
      if (isTRUE(st == 0L) && file.exists(dest) && file.info(dest)$size > 1000L) {
        return(TRUE)
      }
    }
    FALSE
  }

  if (try_download_file()) {
    return(normalizePath(dest, winslash = "/"))
  }

  if (!is.null(geo_acc) && !is.null(supp_filename) && requireNamespace("GEOquery", quietly = TRUE)) {
    base <- tempfile("geo_supp")
    dir.create(base)
    tryCatch(
      suppressMessages(
        GEOquery::getGEOSuppFiles(geo_acc, makeDirectory = TRUE, baseDir = base)
      ),
      error = function(e) invisible(NULL)
    )
    hits <- list.files(
      base,
      pattern = supp_filename,
      fixed = TRUE,
      recursive = TRUE,
      full.names = TRUE
    )
    if (length(hits) >= 1L && file.copy(hits[1], dest, overwrite = TRUE)) {
      if (file.exists(dest) && file.info(dest)$size > 1000L) {
        return(normalizePath(dest, winslash = "/"))
      }
    }
  }

  curl_bin <- Sys.which("curl")
  if (nzchar(curl_bin)) {
    st <- system2(curl_bin, c("-fsSL", "-o", dest, shQuote(url)), stdout = FALSE, stderr = FALSE)
    if (st == 0L && file.exists(dest) && file.info(dest)$size > 1000L) {
      return(normalizePath(dest, winslash = "/"))
    }
  }

  stop(
    "Download failed for:\n  ", url,
    "\nTry: install.packages('curl'); BiocManager::install('GEOquery'); ",
    "options(download.file.method = 'libcurl'); ",
    "or download the file in a browser and set PHENOMAPR_GSE122873_TAR to the local path."
  )
}

download_if_needed <- function(url, dest, geo_acc = NULL, supp_filename = basename(url)) {
  download_geo_file(url, dest, geo_acc = geo_acc, supp_filename = supp_filename)
}

#' Collapse Ensembl rows to HGNC symbols by mean TPM (duplicate Ensembl → one symbol).
collapse_ensembl_tpm_to_symbol <- function(tpm_ens_rows_samples_cols) {
  ens <- rownames(tpm_ens_rows_samples_cols)
  ens_base <- sub("\\.[0-9]+$", "", ens)
  sym <- AnnotationDbi::mapIds(
    org.Hs.eg.db,
    keys = ens_base,
    column = "SYMBOL",
    keytype = "ENSEMBL",
    multiVals = "first"
  )
  df <- as.data.frame(tpm_ens_rows_samples_cols)
  df$gene_symbol <- sym
  df <- df[!is.na(df$gene_symbol) & nzchar(df$gene_symbol), , drop = FALSE]
  sample_cols <- setdiff(names(df), "gene_symbol")
  agg <- stats::aggregate(
    df[, sample_cols, drop = FALSE],
    by = list(gene_symbol = df$gene_symbol),
    FUN = mean,
    na.rm = TRUE
  )
  mat <- as.matrix(agg[, -1, drop = FALSE])
  rownames(mat) <- agg$gene_symbol
  mat
}


  td <- tempdir()
  tpm_path <- download_if_needed(
    tpm_url,
    file.path(td, "GSE140261_year2_TPM.csv.gz"),
    geo_acc = "GSE140261",
    supp_filename = "GSE140261_year2_TPM.csv.gz"
  )
  meta_path <- download_if_needed(
    meta_url,
    file.path(td, "GSE140261_year2_column_data.csv.gz"),
    geo_acc = "GSE140261",
    supp_filename = "GSE140261_year2_column_data.csv.gz"
  )

  tpm_wide <- read.csv(gzfile(tpm_path), row.names = 1, check.names = FALSE, stringsAsFactors = FALSE)
  tpm_sym <- collapse_ensembl_tpm_to_symbol(tpm_wide)

  coldata <- read.csv(gzfile(meta_path), stringsAsFactors = FALSE, check.names = FALSE)
  coldata <- coldata[tolower(coldata$pheno_type) %in% c("fshd", "control"), , drop = FALSE]
  coldata$group <- ifelse(tolower(coldata$pheno_type) == "fshd", "FSHD", "Control")
  coldata$sample_id <- coldata$sample_name

  use_samples <- intersect(colnames(tpm_sym), coldata$sample_id)
  tpm_sym <- tpm_sym[, use_samples, drop = FALSE]
  pheno_bulk <- coldata[match(use_samples, coldata$sample_id), c("sample_id", "group")]
  pheno_bulk$group <- factor(pheno_bulk$group, levels = c("FSHD", "Control"))

  ref_fshd <- derive_reference_from_bulk(
    bulk_expression = tpm_sym,
    phenotype = pheno_bulk,
    sample_id_column = "sample_id",
    phenotype_column = "group",
    phenotype_type = "binary",
    normalize = FALSE,
    binary_positive_reference = "first",
    verbose = TRUE
  )
```

With **`levels = c("FSHD", "Control")`** and **`binary_positive_reference = "first"`**, genes with **positive** z-scores are **higher in FSHD** bulk samples.

```{r bulk-ref-plot, fig.width=6, fig.height=3.5}
if (vignette_run) {
  plot_score_distribution(ref_fshd, main = "GSE140261: FSHD vs control (gene-level z-scores)")
}
```

## 2. Single-cell data (GSE122873) and mitochondrial QC

[GSE122873](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE122873) provides **`GSE122873_RAW.tar`** with six **`GSM*_*.txt.gz`** tables (myocyte cultures: four FSHD + two control). Each table is **genes × cells** with a header row of cell barcodes ([Kendall *et al.*, *Genome Res.* 2019](https://doi.org/10.1101/gr.243734.118); [PMC6423425](https://pmc.ncbi.nlm.nih.gov/articles/PMC6423425/)). **Gene identifiers are Ensembl IDs** (with optional version suffix); **`PhenoMapR`** matches the custom reference by **HUGO symbols**, so we **map Ensembl → symbol** and **sum counts** per symbol **before** building the Seurat object.

**If downloads fail** (common behind strict firewalls or in RStudio: `download.file` to NCBI can error), install **`GEOquery`** (`BiocManager::install("GEOquery")`) so the code can fetch supplementary files via Bioconductor, or set **`PHENOMAPR_GSE122873_TAR`** to the **full path** to a manually downloaded `GSE122873_RAW.tar`.

```{r sc-download}
sc_tar_url <- "https://ftp.ncbi.nlm.nih.gov/geo/series/GSE122nnn/GSE122873/suppl/GSE122873_RAW.tar"

read_gse122873_sample <- function(path) {
  m <- as.matrix(read.delim(gzfile(path), row.names = 1, check.names = FALSE))
  storage.mode(m) <- "double"
  m
}

merge_gse122873 <- function(dir_untar) {
  fs <- list.files(dir_untar, pattern = "^GSM.*\\.txt\\.gz$", full.names = TRUE)
  if (length(fs) < 1L) stop("No GSM*.txt.gz after extracting GSE122873_RAW.tar")
  mats <- lapply(fs, read_gse122873_sample)
  genes <- Reduce(intersect, lapply(mats, rownames))
  combined <- do.call(cbind, lapply(mats, function(m) m[genes, , drop = FALSE]))
  colnames(combined) <- make.unique(colnames(combined), sep = "_")
  sn <- basename(fs)
  lab <- sub("\\.txt\\.gz$", "", sn)
  lab <- sub("^GSM[0-9]+_", "", lab)
  rep_lab <- rep(lab, times = vapply(mats, ncol, 1L))
  list(counts = combined, donor_batch = rep_lab)
}

#' Genes x cells count matrix: Ensembl rownames -> HUGO symbols (sum counts per symbol).
#' Rows already looking like symbols (not ENSG...) are kept as-is.
collapse_counts_ensembl_to_hugo <- function(counts_mat) {
  if (!requireNamespace("org.Hs.eg.db", quietly = TRUE)) {
    stop("Package 'org.Hs.eg.db' is required to map Ensembl IDs to symbols.")
  }
  x <- as.matrix(counts_mat)
  genes <- rownames(x)
  ens_base <- sub("\\.[0-9]+$", "", genes)
  is_ens <- grepl("^ENSG[0-9]+", ens_base, ignore.case = TRUE)
  sym <- genes
  if (any(is_ens)) {
    mapped <- AnnotationDbi::mapIds(
      org.Hs.eg.db,
      keys = ens_base[is_ens],
      column = "SYMBOL",
      keytype = "ENSEMBL",
      multiVals = "first"
    )
    sym[is_ens] <- as.character(mapped)
  }
  ok <- !is.na(sym) & nzchar(sym) & sym != "NA"
  x <- x[ok, , drop = FALSE]
  sym <- sym[ok]
  if (length(unique(sym)) < length(sym)) {
    x <- rowsum(x, group = sym, reorder = FALSE)
  } else {
    rownames(x) <- sym
  }
  x
}

  td_sc <- tempfile("gse122873")
  dir.create(td_sc, showWarnings = FALSE)
  tar_path <- file.path(td_sc, "GSE122873_RAW.tar")
  env_tar <- Sys.getenv("PHENOMAPR_GSE122873_TAR", unset = "")
  if (nzchar(env_tar) && file.exists(env_tar)) {
    file.copy(env_tar, tar_path, overwrite = TRUE)
  } else {
    download_geo_file(
      sc_tar_url,
      tar_path,
      geo_acc = "GSE122873",
      supp_filename = "GSE122873_RAW.tar"
    )
  }
  untar(tar_path, exdir = td_sc)
  sc_list <- merge_gse122873(td_sc)
  counts <- collapse_counts_ensembl_to_hugo(sc_list$counts)
  meta_cells <- data.frame(
    cell_id = colnames(counts),
    donor_batch = sc_list$donor_batch,
    stringsAsFactors = FALSE
  )
  meta_cells$disease <- ifelse(grepl("^CTRL", meta_cells$donor_batch), "Control", "FSHD")
  rownames(meta_cells) <- meta_cells$cell_id

  seu <- CreateSeuratObject(counts = counts, meta.data = meta_cells)
  seu[["percent.mt"]] <- PercentageFeatureSet(seu, pattern = "^MT-")
  seu <- subset(seu, subset = percent.mt < 10)
  seu <- NormalizeData(seu, verbose = FALSE)
  seu <- FindVariableFeatures(seu, verbose = FALSE)
  all.genes <- rownames(seu)
seu <- ScaleData(seu, features = all.genes)
  seu <- RunPCA(seu, verbose = FALSE)

# Cluster the cells and run UMAP #####
seu <- FindNeighbors(seu, dims = 1:10)
seu <- FindClusters(seu, resolution = 0.5)
  seu <- RunUMAP(seu, dims = 1:30, verbose = FALSE)
  seu.markers <- FindAllMarkers(object = seu, only.pos = TRUE)
```

After filtering, you should obtain **about 6,899** cells (exact counts can differ slightly with software versions).

## 3. Score single cells with the custom reference

```{r score-sc}
if (vignette_run) {
  scores_sc <- PhenoMap(
    seu,
    reference = ref_fshd,
    assay = "RNA",
    slot = "data",
    z_score_cutoff = 2,
    verbose = FALSE
  )
  seu <- add_scores_to_seurat(seu, scores_sc)
  score_col <- grep("weighted_sum_score", colnames(scores_sc), value = TRUE)[1]
}
```

```{r umap-score, fig.width=8, fig.height=4.5}
if (vignette_run) {
  pal <- c(Control = "#2166AC", FSHD = "#B2182B")
  p1 <- DimPlot(seu, group.by = "disease", cols = pal) + ggtitle("GSE122873 (culture label)")
  DimPlot(seu, group.by = "disease", cols = pal) + ggtitle("GSE122873 (culture label)")
  p2 <- FeaturePlot(seu, features = score_col, order = TRUE) +
    scale_color_gradientn(colors = c("#2166AC", "#F7F7F7", "#B2182B"))
  if (requireNamespace("patchwork", quietly = TRUE)) {
    suppressPackageStartupMessages(library(patchwork))
    print(p1 + p2)
  } else {
    print(p1)
    print(p2)
  }
}
```

## Use CyteTypeR to assign cell type labels since none are provided
```{r}
# if (requireNamespace("future", quietly = TRUE)) {
#   future::plan("sequential", gc = FALSE)
# }

library(devtools)
BiocManager::install("rhdf5")
library(rhdf5)
install_github("NygenAnalytics/CyteTypeR")
library(CyteTypeR)


# Load libraries ####
library(dplyr)
library(patchwork)
library(Matrix)
library(Seurat)

# Find markers for all Clusters #####
# seu <- FindNeighbors(object = seu, dims = 1:10)
# seu <- FindClusters(object = seu, resolution = 0.5)
# seu.markers <- FindAllMarkers(object = seu, only.pos = TRUE)

# (optional:)
seu.markers %>%
  group_by(cluster) %>%
  dplyr::filter(avg_log2FC > 1)

## Prep data for job submission to cytetype api
prepped_data <- PrepareCyteTypeR(seu,
         seu.markers,
         n_top_genes = 10,
         group_key = 'seurat_clusters',
         aggregate_metadata = TRUE,
         coordinates_key = "umap")

## Adding metadata on 
metadata <- list(
  title = 'FSHD GSE122873',
  run_label = 'initial_analysis',
  experiment_name = 'fshd_GSE122873')


## Submit job to cytetype
# seu <- CyteTypeR(obj = seu,
#                      prepped_data = prepped_data, 
#                           study_context = "FSHD samples from humans and controls", 
#                           metadata = metadata,
#                      require_artifacts = FALSE
#                           )
GetResults(job_id = "ce748d9d-6191-42bc-a212-4bace7e97e4b")
```

```{r score-by-disease,fig.width=6, fig.height=4}
if (vignette_run) {
  plot_data <- data.frame(
    disease = seu$disease,
    score = seu@meta.data[[score_col]],
    stringsAsFactors = FALSE
  )
  ggplot(plot_data, aes(x = disease, y = score, fill = disease)) +
    geom_violin(alpha = 0.6) +
    geom_jitter(width = 0.1, alpha = 0.15, size = 0.3) +
    scale_fill_manual(values = c(Control = "#2166AC", FSHD = "#B2182B")) +
    theme_minimal(base_size = 14) +
    labs(title = "Custom FSHD bulk signature (single-cell scores)", y = score_col, x = NULL)
}
```

## 4. Phenotype tails and marker genes (single-cell)

We label **Most Adverse** / **Most Favorable** / **Other** tails from the custom score (**higher** = more FSHD-like under this reference), then run **global** `find_phenotype_markers()` (`marker_scope = "phenotype_groups"`). For **cell-type–specific** markers, set `marker_scope = "cell_type_specific"` once cell types are annotated (see `vignette("single-cell", package = "PhenoMapR")`).

```{r markers-sc}
if (vignette_run) {
  groups <- define_phenotype_groups(scores_sc, percentile = 0.05, score_columns = score_col)
  group_col <- grep("^phenotype_group_", names(groups), value = TRUE)[1]

  mk <- find_phenotype_markers(
    seu,
    group_labels = groups,
    group_column = group_col,
    cell_id_column = "cell_id",
    marker_scope = "phenotype_groups",
    assay = "RNA",
    slot = "data",
    verbose = FALSE
  )
  message("Adverse markers (top 8):")
  print(utils::head(mk$adverse_markers[order(mk$adverse_markers$p_adj), ], 8))
  message("Favorable markers (top 8):")
  print(utils::head(mk$favorable_markers[order(mk$favorable_markers$p_adj), ], 8))
}
```

## 5. Phenotype marker heatmap (`plot_phenotype_markers`)

After **`find_phenotype_markers()`**, **`plot_phenotype_markers()`** builds a **ComplexHeatmap** of favorable- then adverse-marker rows (row-wise scaled expression), with column annotations for **prognostic group**, **culture label** (here we use **`disease`**: FSHD vs control as a simple strip—not fine-grained cell type), and the continuous **PhenoMapR score**. Install **`ComplexHeatmap`** and **`circlize`** if needed. **`heatmap_type = "global"`** matches **`marker_scope = "phenotype_groups"`** above.

```{r plot-phenotype-markers-fshd, fig.align='center', fig.width=12, fig.height=8}
if (vignette_run && exists("mk") && exists("groups") && exists("group_col")) {
  if (!requireNamespace("ComplexHeatmap", quietly = TRUE) || !requireNamespace("circlize", quietly = TRUE)) {
    message("Install packages 'ComplexHeatmap' and 'circlize' to draw plot_phenotype_markers().")
  } else {
    expr_mat <- tryCatch(
      Seurat::GetAssayData(seu, layer = "data", assay = "RNA"),
      error = function(e) Seurat::GetAssayData(seu, slot = "data", assay = "RNA")
    )
    meta_hm <- as.data.frame(seu@meta.data)
    meta_hm$cell_id <- colnames(seu)
    meta_hm[[group_col]] <- groups[[group_col]][match(meta_hm$cell_id, groups$cell_id)]
    pal_dis <- PhenoMapR::get_celltype_palette(levels(factor(meta_hm$disease)))

    plot_phenotype_markers(
      markers = mk,
      expr_mat = expr_mat,
      meta = meta_hm,
      cell_id_col = "cell_id",
      group_col = group_col,
      score_col = score_col,
      celltype_col = "disease",
      celltype_palette = pal_dis,
      heatmap_type = "global",
      top_n_markers = 20L,
      n_mark_labels = 5L,
      p_adj_threshold = 0.05,
      column_title = "FSHD bulk-derived signature: single-cell marker heatmap"
    )
  }
}
```

## 6. Validation bulk cohorts (optional)

The paper evaluates the FSHD signature on **GSE115650** (34 FSHD vs nine controls; FPKM→TPM), **GSE56787** (muscle biopsies only: 15 FSHD vs nine controls), and **GSE26852** (FSHD vs normal controls only). A minimal pattern is:

1. Download the **series matrix** or supplementary expression table from GEO.
2. Map identifiers to **gene symbols** and align genes with `ref_fshd`.
3. **`PhenoMap(expression = bulk_genes_x_samples, reference = ref_fshd, normalize = …)`** (or pass TPM with `normalize = FALSE` in `derive_reference` only; for scoring single samples, use log-normalized expression consistent with your bulk pipeline).
4. Compare **score by group** (FSHD vs control) with a violin or ROC.

```{r validation-template, eval=FALSE}
# Example skeleton (not run): load GEO series matrix with GEOquery, then score.
# library(GEOquery)
# gse <- getGEO("GSE115650", GSEMatrix = TRUE)[[1]]
# expr <- Biobase::exprs(gse)  # often log-transformed; match preprocessing to your validation paper
# pheno <- Biobase::pData(gse)
# ... build binary labels from pheno columns ...
# scores_val <- PhenoMap(expression = expr, reference = ref_fshd, z_score_cutoff = 2)
```

## Summary

- **Bulk reference** — **GSE140261** TPM, **FSHD vs control**, Ensembl→symbol collapse, **`derive_reference_from_bulk()`**.
- **Single-cell** — **GSE122873**, **<10% MT-** filter, **`PhenoMap()`** + **UMAP** + **`find_phenotype_markers()`** + **`plot_phenotype_markers()`** (global heatmap).
- **Relation to Scissor** — Scissor solves a **different** optimization (bulk–single-cell regression on a similarity matrix); **PhenoMapR** uses **gene-level** bulk contrasts and a **weighted sum** score on cells, but the **data sources and QC** can follow the same GEO-based workflow as in [Sun *et al.*](https://www.nature.com/articles/s41587-021-01091-3).

## References

<a id="ref-scissor"></a> Sun, D., Guan, X., Moran, A. E. *et al.* Identifying phenotype-associated subpopulations by integrating bulk and single-cell sequencing data. *Nat. Biotechnol.* **40**, 527–538 (2022). <https://doi.org/10.1038/s41587-021-01091-3>

<a id="ref-kendall"></a> Kendall, L. E. *et al.* Single-cell RNA sequencing in facioscapulohumeral muscular dystrophy disease etiology and development. *Genome Res.* **29**, 890–901 (2019). [GSE122873](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE122873)

## Session info

```{r session-info}
sessionInfo()
```
