---
title: "GSE111672: Single-cell pancreatic adenocarcinoma (PAAD)"
author: "PhenoMapR"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GSE111672 single-cell}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Overview

[GSE111672](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE111672) integrates microarray-based spatial transcriptomics and single-cell RNA-seq to reveal tissue architecture in **pancreatic ductal adenocarcinoma (PDAC)**. This vignette uses the PAAD single-cell Seurat object (`PAAD_GSE111672_seurat.rds`) in the `Vignettes` directory to demonstrate PhenoMapR: scoring cells with the built-in PRECOG **Pancreatic** reference, comparing score distributions across cell types, and optionally running prognostic group and marker analysis.

## 1. Load the PAAD GSE111672 Seurat object

Load the preprocessed Seurat object from the vignettes directory. The object includes metadata such as `seurat_clusters`, `Cell_Type_Broad`, and (where available) `Tumor_Normal` or `Sample_Site`.

```{r load-setup}
library(PhenoMapR)
library(Seurat)

# Load the PAAD GSE111672 Seurat object (in Vignettes directory)
rds_path <- "Vignettes/PAAD_GSE111672_seurat.rds"
if (!file.exists(rds_path)) rds_path <- "PAAD_GSE111672_seurat.rds"
seurat <- readRDS(rds_path)

dim(seurat)
head(colnames(seurat@meta.data))
```

## 2. Score cells with PRECOG Pancreatic

Score each cell with PhenoMapR using the built-in PRECOG **Pancreatic** reference. (The object may already contain scores; we run `score_expression` here to demonstrate the API.)

```{r score-cells}
# Score with PRECOG primary pancreatic reference
scores_primary <- score_expression(
  expression = seurat,
  reference = "precog",
  cancer_type = "Pancreatic",
  assay = "RNA",
  slot = "data",
  verbose = TRUE
)
seurat <- add_scores_to_seurat(seurat, scores_primary)
```

## 3. Cell type score distributions

Compare prognostic score distributions across **major lineage** (`Celltype..major.lineage.`) using a distinct color palette.

```{r score-by-celltype}
library(ggplot2)

df <- seurat@meta.data
score_col <- "weighted_sum_score_Pancreatic"
celltype_col <- "Celltype..major.lineage."

# Custom color palette (Tol bright scheme–like; distinct from default ggplot)
n_lineages <- length(unique(df[[celltype_col]]))
pal_lineage <- if (n_lineages <= 12) {
  c("#4477AA", "#EE6677", "#228833", "#CCBB44", "#66CCEE", "#AA3377",
    "#BBBBBB", "#EE8866", "#FFAABB", "#99DDFF", "#DDDDDD", "#77AADD")[seq_len(n_lineages)]
} else {
  colorRampPalette(c("#4477AA", "#EE6677", "#228833", "#CCBB44", "#66CCEE"))(n_lineages)
}
names(pal_lineage) <- sort(unique(as.character(df[[celltype_col]])))

# Boxplot of Pancreatic score by major lineage
if (celltype_col %in% names(df)) {
  ggplot(df, aes(
    x = reorder(.data[[celltype_col]], .data[[score_col]], FUN = median),
    y = .data[[score_col]],
    fill = .data[[celltype_col]]
  )) +
    geom_boxplot(outlier.alpha = 0.3) +
    scale_fill_manual(values = pal_lineage, name = "Major lineage") +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "right",
      legend.title = element_text(size = 9)
    ) +
    labs(y = "PRECOG Pancreatic score", x = "Cell type (major lineage)")
} else {
  ggplot(df, aes(x = seurat_clusters, y = .data[[score_col]], fill = factor(seurat_clusters))) +
    geom_boxplot(outlier.alpha = 0.3) +
    scale_fill_brewer(palette = "Set3", name = "Cluster") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") +
    labs(y = "PRECOG Pancreatic score", x = "Cluster")
}
```

## 4. Prognostic groups and marker gene analysis

Define adverse (high score) and favorable (low score) prognostic groups and run marker gene analysis with Seurat. For faster Wilcoxon rank-sum tests, install [presto](https://github.com/immunogenomics/presto)—Seurat will use it automatically when available.

```{r presto-setup, eval = TRUE}
# Optional: install presto for much faster FindMarkers (Seurat uses it automatically)
if (!requireNamespace("presto", quietly = TRUE) && requireNamespace("devtools", quietly = TRUE)) {
  devtools::install_github("immunogenomics/presto")
}
```

```{r markers}
scores_df <- seurat@meta.data[, grep("weighted_sum_score", names(seurat@meta.data)), drop = FALSE]
groups <- define_prognostic_groups(scores_df, percentile = 0.05)
seurat <- AddMetaData(seurat, groups)

# Find marker genes between adverse vs rest and favorable vs rest
group_col <- grep("prognostic_group_weighted_sum_score", names(seurat@meta.data), value = TRUE)[1]
markers <- NULL
if (!is.na(group_col)) {
  markers <- find_prognostic_markers(
    seurat,
    group_labels = seurat@meta.data[[group_col]],
    group_column = NULL,
    assay = "RNA",
    slot = "data"
  )
  if (!is.null(markers)) {
    head(markers$adverse_markers)
    head(markers$favorable_markers)
  }
}
```

## 5. Heatmap: top marker genes across all cells

Visualize the top adverse and favorable marker genes with **all cells as columns** and **genes as rows**. Column annotations show prognostic score, cell type (major lineage), and prognostic group.

```{r heatmap-markers}
n_top <- 15
# GetAssayData: use 'layer' in Seurat v5+, 'slot' in v4
expr <- tryCatch(
  Seurat::GetAssayData(seurat, layer = "data", assay = "RNA"),
  error = function(e) Seurat::GetAssayData(seurat, slot = "data", assay = "RNA")
)

# Top 15 positively expressed marker genes for adverse and favorable (avg_log2FC > 0)
adverse_pos <- markers$adverse_markers[markers$adverse_markers$avg_log2FC > 0, ]
favorable_pos <- markers$favorable_markers[markers$favorable_markers$avg_log2FC > 0, ]
top_genes <- c(
  head(adverse_pos$gene[order(adverse_pos$p_val_adj)], n_top),
  head(favorable_pos$gene[order(favorable_pos$p_val_adj)], n_top)
)
top_genes <- unique(top_genes[top_genes %in% rownames(expr)])
if (length(top_genes) == 0) top_genes <- head(rownames(expr), 20)

mat <- as.matrix(expr[top_genes, , drop = FALSE])
# Row-scale for visualization; clip to [-3, 3] for consistent color scale
mat_scaled <- t(scale(t(mat)))
mat_scaled[mat_scaled < -3] <- -3
mat_scaled[mat_scaled > 3] <- 3

# Order columns: by prognostic group (adverse, middle, favorable) then by score
meta <- seurat@meta.data
meta$cell_id <- colnames(seurat)
ord <- order(
  match(meta[[group_col]], c("adverse", "middle", "favorable")),
  -meta[[score_col]]
)
mat_scaled <- mat_scaled[, ord, drop = FALSE]
meta_ord <- meta[ord, ]

# Column annotations: prognostic score (binned), cell type, prognostic group
score_bin <- cut(meta_ord[[score_col]], breaks = 5, labels = paste0("Q", 1:5), include.lowest = TRUE)
ann_col <- data.frame(
  `Prognostic score` = score_bin,
  `Cell type` = factor(meta_ord[["Celltype..major.lineage."]]),
  `Prognostic group` = factor(meta_ord[[group_col]], levels = c("adverse", "middle", "favorable")),
  check.names = FALSE
)
rownames(ann_col) <- colnames(mat_scaled)

# Palettes for column annotations (non-default)
pal_score <- setNames(
  colorRampPalette(c("#2166AC", "#F7F7F7", "#B2182B"))(5),
  levels(score_bin)
)
pal_group <- c(adverse = "#B2182B", middle = "#F7F7F7", favorable = "#2166AC")
n_ct <- length(unique(ann_col$`Cell type`))
pal_celltype <- setNames(
  colorRampPalette(c("#7F3C8D", "#11A579", "#3969AC", "#F2B701", "#E73F74"))(max(n_ct, 5))[seq_len(n_ct)],
  sort(unique(as.character(ann_col$`Cell type`)))
)
ann_colors <- list(
  `Prognostic score` = pal_score,
  `Cell type` = pal_celltype,
  `Prognostic group` = pal_group
)

# Plot heatmap (genes = rows, cells = columns): viridis fill, scale limits ±3
heatmap_breaks <- seq(-3, 3, length.out = 101)
heatmap_colors <- if (requireNamespace("viridis", quietly = TRUE)) {
  viridis::viridis(100)
} else {
  colorRampPalette(c("#2166AC", "#F7F7F7", "#B2182B"))(100)
}
if (requireNamespace("pheatmap", quietly = TRUE)) {
  pheatmap::pheatmap(
    mat_scaled,
    scale = "none",
    cluster_cols = FALSE,
    cluster_rows = TRUE,
    show_colnames = FALSE,
    annotation_col = ann_col,
    annotation_colors = ann_colors,
    color = heatmap_colors,
    breaks = heatmap_breaks,
    main = "Top adverse & favorable marker genes (all cells)",
    fontsize_row = 8
  )
} else {
  heatmap(mat_scaled, scale = "none", Colv = NA, col = heatmap_colors,
          labCol = FALSE, main = "Top marker genes (all cells)")
}
```

## 6. References

- **GSE111672**: [Integrating microarray-based spatial transcriptomics and single-cell RNA-seq reveals tissue architecture in pancreatic ductal adenocarcinomas](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE111672) (GEO).
- **PhenoMapR PRECOG**: Built-in reference `Pancreatic` from PRECOG 2.0 (Benard et al., *Nucleic Acids Research* 2026).
