---
title: "Scoring single-cell data with PhenoMapR"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_collapsed: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, out.width = "100%")

# When knit from inside the PhenoMapR source tree, load the in-tree
# dev package instead of a stale system install (knitr spawns a fresh R
# process, so an interactive devtools::load_all() does not carry over).
.try_load_phenomapr_dev <- function(max_walk = 6L) {
  start_dir <- tryCatch(
    normalizePath(knitr::opts_knit$get("output.dir") %||% getwd(),
                  mustWork = FALSE),
    error = function(e) getwd()
  )
  if (is.null(start_dir) || !nzchar(start_dir)) start_dir <- getwd()
  cur <- start_dir
  for (i in seq_len(max_walk)) {
    desc <- file.path(cur, "DESCRIPTION")
    if (file.exists(desc)) {
      pkg <- tryCatch(unname(read.dcf(desc, fields = "Package")[[1, 1]]),
                      error = function(e) NA_character_)
      if (identical(pkg, "PhenoMapR") &&
          requireNamespace("pkgload", quietly = TRUE)) {
        pkgload::load_all(cur, quiet = TRUE, helpers = FALSE,
                          export_all = FALSE)
        message("Loaded PhenoMapR (dev) from ", cur)
        return(TRUE)
      }
    }
    parent <- dirname(cur)
    if (identical(parent, cur)) break
    cur <- parent
  }
  FALSE
}
`%||%` <- function(x, y) if (is.null(x)) y else x
```

## Overview

This vignette demonstrates how to use **`PhenoMapR`** on a single-cell RNA-seq dataset. We use the **CRA001160** pancreatic cancer dataset from [Peng et al. 2019](https://doi.org/10.1038/s41422-019-0195-y) [[1]](#ref1). This dataset comprises single-cell RNA-seq of 57,530 pancreatic cells from 25 primary PAAD tumors and 9 control pancreases. Pre-processed expression and metadata were obtained from [TISCH2](https://tisch.compbio.cn/home/)  [[2]](#ref2). We score all samples in the dataset with **TCGA** and **PRECOG** meta-z signatures [[3]](#ref3) then use the metadata and score outputs to visualize cell types of interest, define prognostic groups of cells, and identify marker genes for the most phenotype associated cell populations. Results are compared with those obtained using the **SIDISH** method from [Jolasun et al. 2025](https://www.nature.com/articles/s41467-025-66162-4) [[4]](#ref4).

## Load data

We will use pre-processed expression and annotation files from TISCH2 for the CRA001160 dataset. Download the expression matrix and cell metadata from Google Drive and visualize cell type distributions across all samples in the dataset.

```{r load-data, fig.align='center', fig.width=10, fig.height=7}
suppressPackageStartupMessages({
  if (!isTRUE(.try_load_phenomapr_dev())) {
    library(PhenoMapR)
  }
  library(googledrive)
  library(ggplot2)
  library(ggpubr)
  library(dplyr)
  library(tidyr)
  library(Seurat)
  library(patchwork)
  library(ComplexHeatmap)
  library(circlize)
  library(ggchicklet2)
})

options(googledrive_quiet = TRUE)
googledrive::drive_deauth()
googledrive::drive_download(googledrive::as_id("1PolTXggREz8XmhutCLTQJGCfKxFAzqMl"), "PAAD_CRA001160_expression.h5", overwrite = TRUE)
googledrive::drive_download(googledrive::as_id("17mqxnKOZJn0jW2iD9RV0wZeWsilAIwdu"), "PAAD_CRA001160_CellMetainfo_table.tsv", overwrite = TRUE)

# Expression matrix
expr_mat <- Seurat::Read10X_h5("PAAD_CRA001160_expression.h5")

# Cell and sample metadata
meta <- read.delim("PAAD_CRA001160_CellMetainfo_table.tsv", stringsAsFactors = FALSE, check.names = FALSE) %>%
   mutate(`Celltype (original)` = gsub(" cell", "", `Celltype (original)`)) %>%
   dplyr::rename(celltype_malignancy = `Celltype (malignancy)`,
         celltype_original = `Celltype (original)`)

# visuals for cohort
paad_vs_ctrl_clors <- c(
  "Normal" = "#b35806",
  "Tumor" = "#8073ac"
)

pal_cells <- PhenoMapR::get_celltype_palette(meta$celltype_original)

malignant_colors <- c(
  "Others" = "#007D82FF",
  "Immune cells" = "#388F30FF",
  "Stromal cells" = "#C38961FF",
  "Malignant cells" = "#950404FF"
)

# UMAPS for the cell type clustering
# PDAC vs control
umap_1 <- ggscatter(meta, 
          x = "UMAP_1", 
          y = "UMAP_2",
          size = .5,
          color = "Source",
          palette = paad_vs_ctrl_clors,
          legend = "right"
) + guides(color = guide_legend(override.aes = list(size = 2))) +
    labs(color = "Source") +
  theme_void()


# Broad cell group labels
umap_2 <- ggscatter(meta, 
          x = "UMAP_1", 
          y = "UMAP_2",
          size = .5,
          color = "celltype_malignancy",
          palette = malignant_colors,
          legend = "right"
) + guides(color = guide_legend(override.aes = list(size = 2))) +
    labs(color = "Cell Type (Broad)") +
  theme_void()

# Original cell type labels from the authors
umap_3 <- ggscatter(meta,
          x = "UMAP_1", 
          y = "UMAP_2",
          size = .5,
          color = "celltype_original",
          palette = pal_cells,
          legend = "right"
) + guides(color = guide_legend(override.aes = list(size = 2), legend.spacing.x = unit(0.5, 'cm')))+
  labs(color = "Cell Type", ) +
  theme_void() +
   theme(
    legend.spacing.y = unit(0.1, "cm"),  # reduce vertical spacing
    legend.key.height = unit(0.3, "cm")  # reduce key height
  )


# Order samples by malignant-cell proportion (cohort overview).
# Use plain geom_col() for dense per-sample bars: ggchicklet rounded
# geoms can collapse to invisible segments when many x categories share
# a narrow patchwork panel (seen as empty bars with only one sample drawn).
target_celltype <- "Malignant cells"

prop_data_mal <- meta %>%
  count(Patient, celltype_malignancy, name = "n") %>%
  group_by(Patient) %>%
  mutate(proportion = n / sum(n)) %>%
  ungroup()

sample_order <- prop_data_mal %>%
  filter(celltype_malignancy == target_celltype) %>%
  tidyr::complete(Patient = unique(meta$Patient), fill = list(proportion = 0, n = 0L)) %>%
  arrange(desc(proportion)) %>%
  pull(Patient) %>%
  unique()

prop_data_mal <- prop_data_mal %>%
  mutate(Patient = factor(Patient, levels = sample_order))

prop_data_src <- meta %>%
  count(Patient, Source, name = "n") %>%
  group_by(Patient) %>%
  mutate(proportion = n / sum(n)) %>%
  ungroup() %>%
  mutate(Patient = factor(Patient, levels = sample_order))

prop_data <- meta %>%
  count(Patient, celltype_original, name = "n") %>%
  group_by(Patient) %>%
  mutate(proportion = n / sum(n)) %>%
  ungroup() %>%
  mutate(Patient = factor(Patient, levels = sample_order))

patient_counts <- meta %>%
  mutate(Patient = factor(Patient, levels = sample_order)) %>%
  count(Patient, name = "n", .drop = FALSE)

.bar_theme_top <- theme_pubclean(base_size = 8) +
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = "none"
  )

p_num <- ggplot(patient_counts, aes(x = Patient, y = n)) +
  geom_col(width = 0.9, fill = "grey55") +
  .bar_theme_top +
  labs(x = NULL, y = "# of cells", title = NULL)

p_src <- ggplot(prop_data_src, aes(x = Patient, y = proportion, fill = Source)) +
  geom_col(width = 0.9, position = "stack") +
  scale_y_continuous(labels = scales::percent, expand = c(0, 0)) +
  scale_fill_manual(values = paad_vs_ctrl_clors) +
  .bar_theme_top +
  labs(x = NULL, y = NULL, fill = "Source")

p_mal <- ggplot(prop_data_mal, aes(x = Patient, y = proportion, fill = celltype_malignancy)) +
  geom_col(width = 0.9, position = "stack") +
  scale_y_continuous(labels = scales::percent, expand = c(0, 0)) +
  scale_fill_manual(values = malignant_colors, drop = FALSE) +
  .bar_theme_top +
  labs(x = NULL, y = NULL, fill = "Cell Type (Broad)")

p_cell <- ggplot(prop_data, aes(x = Patient, y = proportion, fill = celltype_original)) +
  geom_col(width = 0.9, position = "stack") +
  scale_y_continuous(labels = scales::percent, expand = c(0, 0)) +
  scale_fill_manual(values = pal_cells, drop = FALSE) +
  theme_pubclean(base_size = 8) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    legend.position = "none"
  ) +
  labs(x = NULL, y = "Proportion of cells", fill = "Cell Type", title = NULL)

left_plots <- (p_num / p_src / p_mal / p_cell) +
  plot_layout(heights = c(1.2, 0.55, 0.55, 4.2), guides = "collect")

right_plots <- (umap_1 / umap_2 / umap_3) +
  plot_layout(heights = c(1, 1, 1), guides = "collect")

(left_plots | right_plots) +
  plot_layout(widths = c(1.15, 0.55))

```

## Score cells with PhenoMapR

Because we have survival meta-z signatures from both TCGA and PRECOG PAAD resources, we will score the expression matrix with both references and compare. We then merge score columns into metadata for downstream analysis.

```{r score-cells, fig.align='center'}
# Score all cells against the TCGA PAAD meta-z signature
scores_tcga <- PhenoMap(expression = expr_mat, 
                        reference = "tcga", 
                        cancer_type = "PAAD", 
                        verbose = TRUE)

# Score all cells against the PRECOG Primary Pancreatic meta-z signature
scores_precog <- PhenoMap(expression = expr_mat, 
                          reference = "precog", 
                          cancer_type = "Pancreatic", 
                          verbose = TRUE)

for (col in names(scores_tcga)) meta[[col]] <- scores_tcga[meta$Cell, col]
for (col in names(scores_precog)) meta[[col]] <- scores_precog[meta$Cell, col]

score_tcga_col <- grep("PhenoMapR_.*PAAD", names(meta), value = TRUE, ignore.case = TRUE)[1]
score_precog_col <- grep("PhenoMapR_.*Pancreatic", names(meta), value = TRUE, ignore.case = TRUE)[1]
if (is.na(score_tcga_col)) score_tcga_col <- names(scores_tcga)[1]
if (is.na(score_precog_col)) score_precog_col <- names(scores_precog)[1]
```

## Compare TCGA vs PRECOG PhenoMapR scores

Here, we take the resulting **`PhenoMapR`** scores from using TCGA and PRECOG as references and check to make sure they are correlated within each of the four main cell type categories.

```{r scatter-tcga-precog, fig.align='center'}
df_scatter <- data.frame(
  x = meta[[score_precog_col]],
  y = meta[[score_tcga_col]],
  Cell_Type = factor(meta$celltype_malignancy)
)

ggpubr::ggscatter(df_scatter, x = "x", y = "y", color = "Cell_Type", palette = malignant_colors,
    add = "reg.line", conf.int = TRUE, size = 0.5, alpha = 0.3) +
    ggpubr::stat_cor(aes(color = Cell_Type), label.x.npc = "left", size = 3.5, show.legend = F) +
  scale_color_manual(values = malignant_colors) +
  scale_fill_manual(values = malignant_colors) +
    scale_x_continuous("PRECOG score") +
    scale_y_continuous("TCGA score") +
    labs(title = "TCGA vs. PRECOG PhenoMapR scores by broad cell type") +
    theme_pubr(base_size = 10) +
    theme(legend.position = "right", legend.title = element_text(size = 12), legend.text = element_text(size = 10))

```

```{r scatter-tcga-precog-detailed, fig.align='center'}
df_scatter <- data.frame(
  x = meta[[score_precog_col]],
  y = meta[[score_tcga_col]],
  Cell_Type = factor(meta$celltype_original)
)

ggpubr::ggscatter(df_scatter, x = "x", y = "y", color = "Cell_Type", palette = pal_cells,
    add = "reg.line", conf.int = TRUE, size = 0.5, alpha = 0.3) +
    ggpubr::stat_cor(aes(color = Cell_Type), label.x.npc = "left", size = 3.5, show.legend = F) +
  scale_color_manual(values = pal_cells) +
  scale_fill_manual(values = pal_cells) +
    scale_x_continuous("PRECOG score") +
    scale_y_continuous("TCGA score") +
    labs(title = "TCGA vs. PRECOG PhenoMapR scores by cell type") +
    theme_pubr(base_size = 10) +
    theme(legend.position = "right", legend.title = element_text(size = 12), legend.text = element_text(size = 10))

```

As expected, **`PhenoMapR`** scores are strongly correlated between TCGA and PRECOG references across all major cell types.


## PhenoMapR score UMAP

```{r PhenoMapR-UMAP, fig.height=2.5, fig.align='center'}

# PRECOG
meta_ordered <- meta %>%
  arrange(abs(PhenoMapR_Pancreatic))

# scale the score for easier visualization
meta_ordered$score_scaled <- scale(meta_ordered$PhenoMapR_Pancreatic)

precog_scaled_umap <- ggscatter(meta_ordered,
          x = "UMAP_1", 
          y = "UMAP_2",
          size = 1,
          color = "score_scaled",
          legend = "right", 
          title = NULL
) + 
  scale_color_gradient2(
    low = "#2166AC",
    mid = "#F7F7F7",
    high = "#B2182B",
    midpoint = 0,  
    name = "PhenoMapR\nScore"
  ) +
  theme_void() +
    theme(plot.title = element_text(hjust = 0.5)) 

# Let's plot next to the cell type UMAP for comparison
(umap_2 | precog_scaled_umap) + 
  plot_layout(widths = c(1, 1))
```
It seems like the most adversely scoring cells are enriched in several of the Ductal type 2 cell clusters while the most favorable are enriching in the acinar and endocrine clusters.


## Prognostic score distribution by Cell Type

The qualitative assessment of PhenoMapR score enrichment across UMAP clusters nominated some cell types as potentially more associated with the phenotypic signature than others. Here, we plot the absolute PhenoMapR score (PRECOG and TCGA scores) across all cell types in the dataset.

```{r score-distribution-boxplots, fig.align='center', fig.height=3.5, fig.width=5.5}
med_precog <- setNames(
  tapply(meta[[score_precog_col]], meta$celltype_malignancy, median, na.rm = TRUE),
  levels(factor(meta$celltype_malignancy))
)
# med_precog <- med_precog[!is.na(med_precog)]
ct_order <- names(sort(med_precog))
meta$celltype_malignancy <- factor(meta$celltype_malignancy, levels = ct_order)

dl <- rbind(
  data.frame(Reference = "TCGA PAAD", Score = meta[[score_tcga_col]], Cell_type = meta$celltype_malignancy, stringsAsFactors = FALSE),
  data.frame(Reference = "PRECOG Pancreatic", Score = meta[[score_precog_col]], Cell_type = meta$celltype_malignancy, stringsAsFactors = FALSE)
)
dl$Reference <- factor(dl$Reference, levels = c("TCGA PAAD", "PRECOG Pancreatic"))
ggplot(dl, aes(x = Cell_type, y = Score, fill = Cell_type)) +
  # geom_boxplot(outlier.alpha = 0.2) +
  ggchicklet2::geom_chicklet_boxplot(radius = grid::unit(1, "pt"),outlier.alpha = 0.5, median.linewidth = 0.5,
                 outlier.fill = NULL,
                 outlier.color = NULL, outlier.shape = 21) +
  facet_wrap(~ Reference, ncol = 2) +
  scale_fill_manual(values = malignant_colors, name = "Cell Type (Broad)") +
  theme_pubr(base_size = 10) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "right", 
        strip.text = element_text(size = 12)) +
  labs(x = NULL, y = "PhenoMapR score", title = "Score distribution by reference and cell type")
```

**`PhenoMapR`** successfully identifies the **Malignant** cell compartment as most associated with the adverse prognostic signature in both TCGA and PRECOG.

## Score distribution by refined cell type

To see if the **`PhenoMapR`** score assignment is enriched in more granular cell types, we use the original cell type labels provided by the authors (*Peng et al.* 2019).

```{r score-distribution-celltype-original, fig.align='center', fig.height=3.5, fig.width=7}
med_precog <- setNames(
  tapply(meta[[score_precog_col]], meta$celltype_original, median, na.rm = TRUE),
  levels(factor(meta$celltype_original))
)
# med_precog <- med_precog[!is.na(med_precog)]
ct_order <- names(sort(med_precog))
meta$celltype_original <- factor(meta$celltype_original, levels = ct_order)

dl <- rbind(
  data.frame(Reference = "TCGA PAAD", Score = meta[[score_tcga_col]], Cell_type = meta$celltype_original, stringsAsFactors = FALSE),
  data.frame(Reference = "PRECOG Pancreatic", Score = meta[[score_precog_col]], Cell_type = meta$celltype_original, stringsAsFactors = FALSE)
)
dl$Reference <- factor(dl$Reference, levels = c("TCGA PAAD", "PRECOG Pancreatic"))
ggplot(dl, aes(x = Cell_type, y = Score, fill = Cell_type)) +
  # geom_boxplot(outlier.alpha = 0.2) +
  ggchicklet2::geom_chicklet_boxplot(radius = grid::unit(1, "pt"),
                                     outlier.alpha = 0.5, median.linewidth = 0.5,
                 outlier.fill = NULL,
                 outlier.color = NULL, outlier.shape = 21) +
  facet_wrap(~ Reference, ncol = 2) +
  scale_fill_manual(values = pal_cells, name = "Cell Type") +
  theme_pubr(base_size = 10) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "right", strip.text = element_text(size = 12)) +
  labs(x = NULL, y = "PhenoMapR score", title = "Score distribution by reference and cell type")

```

In both TCGA and PRECOG results, we see that the Ductal cell type 2 is the most significantly associated cell type with an adverse prognostic signature in PAAD. This agrees with the results from **SIDISH** [Jolasun et al.](https://www.nature.com/articles/s41467-025-66162-4/figures/2), where they find **over 55% of their high-risk cells were Ductal cell type 2**. 


## Score distribution by sample source

Theoretically, the PhenoMapR scores should be differentially distributed between tumor or normal samples. We plot the scores per cell type, split by tumor/normal source:

```{r, fig.height=5}
library(ggsignif)
library(purrr)
library(dplyr)

dl <- 
  data.frame(Reference = "PRECOG Pancreatic", Source = meta$Source, Score = meta[[score_precog_col]], Cell_type = meta$celltype_original, stringsAsFactors = FALSE)

cell_levels <- levels(factor(dl$Cell_type))  # or your factor order

# For each cell type, check which sources are present
source_offsets <- c("Normal" = -0.2, "Tumor" = 0.2)  # matches dodge width = 0.8 → each side ±0.2

# --- 2. Compute p-values per cell type ---
pval_df <- dl %>%
  group_by(Cell_type) %>%
  summarise(
    n_sources = n_distinct(Source),
    p_val = ifelse(
      n_distinct(Source) > 1,
      wilcox.test(scale(Score)[Source == "Normal"],
                  scale(Score)[Source == "Tumor"])$p.value,
      NA_real_
    ),
    .groups = "drop"
  ) %>%
  filter(!is.na(p_val)) %>%
  mutate(
    # Use the SAME factor used in ggplot
    x_pos  = as.integer(factor(Cell_type, levels = levels(factor(dl$Cell_type)))),
    xmin   = x_pos - 0.2,
    xmax   = x_pos + 0.2,
    # Per-group y_pos based on actual data max for that cell type
    y_pos  = max(scale(dl$Score), na.rm = TRUE) * 1.05,
    label  = case_when(
      p_val < 0.001 ~ "***",
      p_val < 0.01  ~ "**",
      p_val < 0.05  ~ "*",
      TRUE          ~ "ns"
    )
  )

# --- 3. Plot ---
ggplot(dl, aes(x = Cell_type, y = scale(Score), fill = Cell_type, color = Source)) +
  # geom_boxplot(outlier.alpha = 0.2, position = position_dodge(width = 0.8)) +
    #   geom_boxplot(outlier.alpha = 0.5, median.linewidth = 0.5,
    #              outlier.fill = NULL,
    #              outlier.color = NULL, outlier.shape = 21) +
    # # geom_boxplot(outlier.alpha = 0.2) +
  ggchicklet2::geom_chicklet_boxplot(radius = grid::unit(1, "pt"),
                                     outlier.alpha = 0.5, median.linewidth = 0.5,
                 outlier.fill = NULL,
                 outlier.color = NULL, outlier.shape = 21) +
  geom_signif(
    data        = pval_df,
    aes(xmin = xmin, xmax = xmax, annotations = label, y_position = y_pos),
    manual      = TRUE,
    tip_length  = 0.02,
    textsize    = 4,
    color       = "black",   # override the color aes from the main plot
    inherit.aes = T
  ) +
  scale_fill_manual(values = pal_cells, name = "\nCell Type") +
  scale_color_manual(values = c("Normal" = "lightgrey", "Tumor" = "black"), name = "Source") +
  guides(
    fill  = guide_legend(order = 1),
    color = guide_legend(order = 2)
  ) +
  theme_pubr(base_size = 14) +
  theme(
    axis.text.x     = element_text(angle = 45, hjust = 1),
    legend.position = "right",
    strip.text      = element_text(size = 12),
    plot.title      = element_text(hjust = 0.5)
  ) +
  labs(x = NULL, y = "PhenoMapR score (scaled)", title = "PhenoMapR Score distribution by Source and Cell Type")
```
Indeed, tumor samples are significantly enriched for more adverse scores across all cell types except B cells.

## Cell types in the most prognostic populations

Cell type counts in the top 5% (Most Adverse) and bottom 5% (Most Favorable) of cells by PRECOG score.

```{r celltype-in-extreme-percentiles, fig.align='center', fig.height=2, fig.width=4}
scores_tail <- meta[, score_precog_col, drop = FALSE]
rownames(scores_tail) <- meta$Cell
groups_tail <- define_phenotype_groups(
  scores_tail,
  percentile = 0.05,
  score_columns = score_precog_col
)
tail_col <- grep("^phenotype_group_", names(groups_tail), value = TRUE)[1]
pg_tmp <- groups_tail[[tail_col]][match(meta$Cell, groups_tail$cell_id)]
idx_extreme <- pg_tmp %in% c("Most Adverse", "Most Favorable")

ct_in_extreme <- meta[idx_extreme, , drop = FALSE]
ct_in_extreme$pg <- pg_tmp[idx_extreme]
ct_counts <- as.data.frame(
  table(Cell_type = ct_in_extreme$celltype_malignancy, pg = ct_in_extreme$pg),
  stringsAsFactors = FALSE
)
ct_wide <- tidyr::pivot_wider(
  ct_counts,
  names_from = pg,
  values_from = Freq,
  values_fill = 0
)
for (c in c("Most Adverse", "Most Favorable")) {
  if (!c %in% names(ct_wide)) ct_wide[[c]] <- 0L
}
ct_wide <- ct_wide[order(ct_wide[["Most Adverse"]], decreasing = TRUE), , drop = FALSE]
ct_long <- rbind(
  data.frame(Cell_type = ct_wide$Cell_type, pg = "Most Favorable", n = -ct_wide[["Most Favorable"]]),
  data.frame(Cell_type = ct_wide$Cell_type, pg = "Most Adverse", n = ct_wide[["Most Adverse"]])
)
ct_long$pg <- factor(ct_long$pg, levels = c("Most Favorable", "Most Adverse"))
max_n <- max(abs(ct_long$n), 1)
ggplot(ct_long, aes(x = Cell_type, y = n, fill = pg)) +
  geom_col(width = 0.75) +
  coord_flip() +
  scale_fill_manual(
    values = c(`Most Adverse` = "#B2182B", `Most Favorable` = "#2166AC"),
    name = "Prognostic group"
  ) +
  scale_y_continuous(
    breaks = seq(-max_n, max_n, length.out = 5),
    labels = abs(seq(-max_n, max_n, length.out = 5))
  ) +
  theme_pubr(base_size = 10) +
  theme(
    axis.text.y = element_text(size = 9),
    legend.position = c(.8, .7),
    plot.title = element_text(hjust = 0.5, size = 10)
  ) +
  labs(
    x = NULL,
    y = "Cell count (in 5th percentile)",
    title = "Broad cell type abundance in \nadverse vs. favorable extremes"
  )
```
At a broad cell type level, PhenoMapR assigns the vast majority of the top 5% of adverse cells to the malignant compartment, however the immune, stromal, and other cell types do contain some of the most adversely scoring cells.


We then use the original cell type annotations from the authors to provide a more refined understanding of the prognostic score assignment across cell types.

```{r celltype-original-in-extreme-percentiles, fig.align='center', fig.height=3, fig.width=4}
ct_orig_in_extreme <- meta[idx_extreme, , drop = FALSE]
ct_orig_in_extreme$pg <- pg_tmp[idx_extreme]
ct_orig_counts <- as.data.frame(
  table(Cell_type_original = ct_orig_in_extreme$celltype_original, pg = ct_orig_in_extreme$pg),
  stringsAsFactors = FALSE
)
ct_orig_wide <- tidyr::pivot_wider(
  ct_orig_counts,
  names_from = pg,
  values_from = Freq,
  values_fill = 0
)
for (c in c("Most Adverse", "Most Favorable")) {
  if (!c %in% names(ct_orig_wide)) ct_orig_wide[[c]] <- 0L
}
ct_orig_wide <- ct_orig_wide[order(ct_orig_wide[["Most Adverse"]], decreasing = TRUE), , drop = FALSE]
ct_orig_long <- rbind(
  data.frame(Cell_type_original = ct_orig_wide$Cell_type_original, pg = "Most Favorable", n = -ct_orig_wide[["Most Favorable"]]),
  data.frame(Cell_type_original = ct_orig_wide$Cell_type_original, pg = "Most Adverse", n = ct_orig_wide[["Most Adverse"]])
)
ct_orig_long$pg <- factor(ct_orig_long$pg, levels = c("Most Favorable", "Most Adverse"))
max_n_orig <- max(abs(ct_orig_long$n), 1)
ggplot(ct_orig_long, aes(x = Cell_type_original, y = n, fill = pg)) +
  geom_col(width = 0.75) +
  coord_flip() +
  scale_fill_manual(
    values = c(`Most Adverse` = "#B2182B", `Most Favorable` = "#2166AC"),
    name = "Prognostic group"
  ) +
  scale_y_continuous(
    breaks = round(seq(-max_n_orig, max_n_orig, length.out = 5)),
    labels = abs(round(seq(-max_n_orig, max_n_orig, length.out = 5)))
  ) +
  theme_pubr(base_size = 10) +
  theme(
    axis.text.y = element_text(size = 9),
    legend.position = c(.8, .75),
    plot.title = element_text(hjust = 0.5, size = 10)
  ) +
  labs(
    x = NULL,
    y = "Cell count (in 5th percentile)",
    title = "Cell Type abundance in adverse\nvs. favorable extremes"
  )
```
Here, we see that the vase majority of the top 5th percentile of adverse PhenoMapR cells are from the Ductal type 2 cells, in line with the results of [Jolasun et al.](https://www.nature.com/articles/s41467-025-66162-4/figures/2). The majority of favorably scoring cells were Acinar and B cell populations. 



## Phenotype groups and marker genes

**`PhenoMapR`** provides some high-level analysis functions to help characterize the most phenotypically relevant cells. We define the "Most Adverse" and "Most Favorable" cells from the **PRECOG PhenoMapR** score (5th/95th percentiles), then run marker identification (adverse vs rest, favorable vs rest) using the original cell type annotation for context.

```{r prognostic-groups-markers, cache = FALSE}
scores_df <- meta[, c(score_tcga_col, score_precog_col), drop = FALSE]
rownames(scores_df) <- meta$Cell
groups <- define_phenotype_groups(
  scores_df,
  percentile = 0.05,
  score_columns = score_precog_col
)
group_col <- grep("^phenotype_group_", names(groups), value = TRUE)[1]
meta[[group_col]] <- groups[[group_col]][match(meta$Cell, groups$cell_id)]

group_df <- data.frame(
  cell_id = meta$Cell,
  phenotype_group = meta[[group_col]],
  cell_type = meta$celltype_original,
  stringsAsFactors = FALSE
)

markers <- find_phenotype_markers(
  expr_mat,
  group_labels = group_df,
  group_column = "phenotype_group",
  cell_id_column = "cell_id",
  marker_scope = "phenotype_groups",
  pval_threshold = 0.05,
  max_cells_per_ident = 5000L,
  verbose = FALSE
)
if (!is.null(markers)) {
  message("Adverse markers (top 5):"); print(head(markers$adverse_markers, 5))
  message("Favorable markers (top 5):"); print(head(markers$favorable_markers, 5))
}
```

## Marker genes heatmap

After identifying marker genes for the adverse and favorable **`PhenoMapR`** populations, we take up to **10** positive markers per tail with **FDR < 0.05** and the **largest `avg_log2FC`** (same thresholds as the cell-type–specific heatmap). The helper **`plot_phenotype_markers()`** subsets the expression matrix to those genes and the ordered cells, then applies **row-wise `scale`** only to that submatrix. **ComplexHeatmap** shows favorable-marker rows, then adverse-marker rows, with **`anno_mark`** for the top five genes in each block.

```{r heatmap-markers, fig.align='center', fig.width=12, fig.height=4}
plot_phenotype_markers(
    markers = markers,
    expr_mat = expr_mat,
    meta = meta,
    cell_id_col = "Cell",
    group_col = group_col,
    score_col = score_precog_col,
    celltype_col = "celltype_original",
    celltype_palette = pal_cells,
    heatmap_type = "global",
    top_n_markers = 10L,
    n_mark_labels = 10L,
    p_adj_threshold = 0.05
  )
```
It's clear from the heatmap that the cell type driving the favorable prognostic signal is the Acinar cell type. Cell-type specific marker identification within the most prgnostic group would further resolve other cell type markers associated with favorable outcomes in this dataset. 

## Cell-type-specific phenotype marker genes
The `find_phenotype_markers()` function can also identify marker genes **for each cell type** by contrasting cells in the phenotype tails (Most Adverse or Most Favorable) against a reference group. By default (`celltype_contrast = "within_cell_type"`), the reference is **only other cells of the same type** (`Other` and the opposite tail)—this reduces ubiquitous genes appearing in many blocks. The **original** cohort-wide contrast `(cell type ∩ tail)` vs **all other cells** is available as `celltype_contrast = "vs_cohort_rest"`. By default (`celltype_unique_genes = TRUE`), each gene is kept in a **single** cell-type–phenotype block (the one with the strongest `avg_log2FC`), so ubiquitous transcripts are not repeated across many blocks. We visualize the top marker genes per cell type and phenotype bin in a heatmap that includes *all* cells, ordered by:
1) phenotype bin (Most Favorable, Other, Most Adverse), then
2) cell type, then
3) phenotype score.

For the heatmap, we take up to **20** positive markers per cell type and phenotype tail with the **largest `avg_log2FC`**, restricting to genes that are **FDR-significant** (`p_adj` from Benjamini–Hochberg `< 0.05`). Rows follow the same column order: **Most Favorable** (each cell type), **Other** cells, **Most Adverse** (each cell type). **`plot_phenotype_markers(..., heatmap_type = "cell_type_specific")`** draws row splits between blocks; **`anno_mark`** labels the **top five** genes (by log2FC) within each (phenotype tail × cell type) group, with **cell type** and **phenotype** strips on the left. Setting **`color_mark_labels_by_celltype = TRUE`** colors each labeled gene with its cell-type palette so markers are easier to read against the strips.

```{r celltype-specific-phenotype-markers, fig.width=15, fig.height=7}
markers_ct <- find_phenotype_markers(
  expr_mat,
  group_labels = group_df,
  group_column = "phenotype_group",
  cell_id_column = "cell_id",
  cell_type_column = "cell_type",
  marker_scope = "cell_type_specific",
  celltype_contrast = "within_cell_type",
  pval_threshold = 0.05,
  max_cells_per_ident = 5000L,
  verbose = FALSE
)

plot_phenotype_markers(
  markers = markers_ct,
  expr_mat = expr_mat,
  meta = meta,
  cell_id_col = "Cell",
  group_col = group_col,
  score_col = score_precog_col,
  celltype_col = "celltype_original",
  celltype_palette = pal_cells,
  heatmap_type = "cell_type_specific",
  celltype_contrast = "within_cell_type",
  top_n_markers = 10L,
  n_mark_labels = 5L,
  color_mark_labels_by_celltype = TRUE,
  p_adj_threshold = 0.05,
  outline_marker_blocks = TRUE
)
```

```{r celltype-specific-against-tail-phenotype-markers, fig.width=12, fig.height=6}
markers_ct_vs_tail <- find_phenotype_markers(
  expr_mat,
  group_labels = group_df,
  group_column = "phenotype_group",
  cell_id_column = "cell_id",
  cell_type_column = "cell_type",
  marker_scope = "cell_type_specific",
  celltype_contrast = "vs_opposite_tail",
  pval_threshold = 0.05,
  max_cells_per_ident = 5000L,
  verbose = FALSE
)

plot_phenotype_markers(
  markers = markers_ct_vs_tail,
  expr_mat = expr_mat,
  meta = meta,
  cell_id_col = "Cell",
  group_col = group_col,
  score_col = score_precog_col,
  celltype_col = "celltype_original",
  celltype_palette = pal_cells,
  heatmap_type = "cell_type_specific",
  celltype_contrast = "vs_opposite_tail",
  outline_marker_blocks = TRUE,
  block_outline_color = "black",
  top_n_markers = 5L,
  n_mark_labels = 5L,
  color_mark_labels_by_celltype = TRUE,
  p_adj_threshold = 0.05
)
```

## Dataset summary of PhenoMapR results

Here, we summarize the enrichment of prognostic PhenoMapR scores across samples and cell types to highlight the difference in signal between healthy and tumor samples as well as between different cell populations across the sample groups. The **top panel** shows **one PRECOG score per pseudo-bulk sample profile** without building a **`Seurat`** object: we map each cell’s barcode to a sample ID (same column used elsewhere in this figure), **sum counts per gene across all cells in that sample**, and assemble a **genes × samples** table (`expr_agg_df`, one column per sample). That object is the usual **bulk-style** input to **`PhenoMap()`** (rows = genes, columns = samples), so each column is scored once with the same weighted-sum PRECOG logic used for bulk data. (The **`pseudobulk = TRUE`** argument applies when **`PhenoMap`** aggregates from a **`Seurat`** object; here aggregation is explicit in the table, which avoids large **`Seurat`** allocations and matches **sum-per-sample** pseudo-bulk expression.) Values are **z-scaled across samples** so bars are comparable for relative high/low bulk-like signal per sample. The **stacked bar** uses the same **5th / 95th percentile** PRECOG tails as elsewhere in the vignette (**`define_phenotype_groups(..., percentile = 0.05)`**), applied again in this chunk so every cell receives a **Most Adverse**, **Most Favorable**, or **Other** label and **per-sample proportions of tail cells always sum sensibly** over the full cohort. **All panels share one sample axis order** (`sample_lev`): factor **levels** are used when the sample column is already a factor (e.g. **`Patient`** after ordering earlier in the vignette); otherwise **first appearance in `meta`**. **`merge()`** output for bar heights is **re-sorted** to that order so horizontal positions match the pseudo-bulk and dot plots.

```{r proportion-by-sample-celltype, fig.height=5.6, fig.align='center'}
sample_col <- NULL
for (cand in c("Sample", "sample", "Patient", "patient", "Sample_ID", "orig.ident")) {
  if (cand %in% names(meta)) {
    sample_col <- cand
    break
  }
}
if (is.null(sample_col)) sample_col <- names(meta)[1]

meta_plot <- meta
meta_plot$sample <- meta_plot[[sample_col]]
meta_plot$cell_type <- meta_plot$celltype_malignancy
# meta_plot$cell_type_original <- if (!is.null(celltype_original) && celltype_original %in% names(meta_plot)) meta_plot$celltype_original else NA_character_
meta_plot$cell_type_original <- meta$celltype_original
## Tail labels for p_bar / dots: same rule as marker chunk (5% / 95% on PRECOG score), keyed by Cell
scores_tail <- meta[, score_precog_col, drop = FALSE]
rownames(scores_tail) <- meta$Cell
groups_tail <- define_phenotype_groups(scores_tail, percentile = 0.05, score_columns = score_precog_col)
tail_col_bar <- grep("^phenotype_group_", names(groups_tail), value = TRUE)[1]
meta_plot$prognostic_grp <- unname(groups_tail[as.character(meta_plot$Cell), tail_col_bar])
## Canonical sample order for every panel: preserve factor levels if set upstream (e.g. Patient),
## else first appearance in `meta` (stable, not alphabetical).
sc_ids <- meta[[sample_col]]
sample_lev <- if (is.factor(sc_ids)) {
  levels(droplevels(sc_ids))
} else {
  unique(as.character(sc_ids))
}
meta_plot$sample <- factor(as.character(meta_plot[[sample_col]]), levels = sample_lev)
n_per_sample <- setNames(as.numeric(table(meta[[sample_col]])[sample_lev]), sample_lev)
n_per_sample[is.na(n_per_sample)] <- 0

## Pseudo-bulk table: cell → sample mapping, sum counts per gene per sample, score with PhenoMap (no Seurat)
cells_pb <- intersect(colnames(expr_mat), meta$Cell)
expr_pb <- expr_mat[, cells_pb, drop = FALSE]
meta_pb <- meta[match(cells_pb, meta$Cell), , drop = FALSE]
rownames(meta_pb) <- cells_pb
sample_ids <- as.character(meta_pb[[sample_col]])
ns <- length(sample_lev)
pb_mat <- matrix(
  0,
  nrow = nrow(expr_pb),
  ncol = ns,
  dimnames = list(rownames(expr_pb), as.character(sample_lev))
)
col_filled <- rep(FALSE, ns)
for (j in seq_len(ns)) {
  idx <- sample_ids == as.character(sample_lev[j])
  if (!any(idx)) next
  col_filled[j] <- TRUE
  sm <- expr_pb[, idx, drop = FALSE]
  pb_mat[, j] <- if (inherits(sm, "Matrix") || inherits(sm, "sparseMatrix")) {
    as.numeric(Matrix::rowSums(sm))
  } else {
    rowSums(sm)
  }
}
expr_agg_df <- as.data.frame(pb_mat, stringsAsFactors = FALSE, check.names = FALSE)
scores_precog_pb <- PhenoMap(
  expression = expr_agg_df,
  reference = "precog",
  cancer_type = "Pancreatic",
  verbose = FALSE
)
pb_score_col <- if (score_precog_col %in% names(scores_precog_pb)) {
  score_precog_col
} else {
  grep("^PhenoMapR_", names(scores_precog_pb), value = TRUE)[1]
}
pb_mean <- as.numeric(scores_precog_pb[[pb_score_col]][match(as.character(sample_lev), rownames(scores_precog_pb))])
pb_mean[!col_filled] <- NA_real_
ok_pb <- is.finite(pb_mean)
pb_scaled <- rep(NA_real_, length(pb_mean))
if (sum(ok_pb) >= 2L && stats::sd(pb_mean[ok_pb], na.rm = TRUE) > 0) {
  pb_scaled[ok_pb] <- as.numeric(scale(pb_mean[ok_pb]))
} else if (any(ok_pb)) {
  pb_scaled[ok_pb] <- 0
}
pb_df <- data.frame(
  sample = factor(sample_lev, levels = sample_lev),
  pb_score_scaled = pb_scaled,
  stringsAsFactors = FALSE
)
p_pbulk <- ggplot(pb_df, aes(x = .data$sample, y = .data$pb_score_scaled, fill = .data$pb_score_scaled)) +
  geom_col(width = 0.85, na.rm = FALSE) +
  scale_fill_gradient2(
    low = "#2166AC",
    mid = "#FFFFFF",
    high = "#B2182B",
    midpoint = 0,
    na.value = "#d9d9d9",
    guide = "none"
  ) +
  geom_hline(yintercept = 0, linewidth = 0.35, linetype = "dashed", color = "grey40") +
  scale_x_discrete(drop = FALSE) +
  scale_y_continuous(expand = expansion(mult = c(0.08, 0.08))) +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.x = element_blank(),
    legend.position = "none"
  ) +
  labs(y = "Sample\nPseudobulk\nscore", title = NULL)

# Per-sample proportion of adverse and favorable (5th / 95th % tails on PRECOG) for stacked bar
meta_adv_fav <- meta_plot[meta_plot$prognostic_grp %in% c("Most Adverse", "Most Favorable"), ]
bar_df <- as.data.frame(table(sample = meta_adv_fav$sample, pg = meta_adv_fav$prognostic_grp, useNA = "no"))
bar_df <- bar_df[bar_df$pg %in% c("Most Adverse", "Most Favorable"), ]
bar_df$total <- n_per_sample[as.character(bar_df$sample)]
bar_df$proportion <- ifelse(bar_df$total > 0, bar_df$Freq / bar_df$total, 0)
fav <- bar_df[bar_df$pg == "Most Favorable", c("sample", "proportion")]
adv <- bar_df[bar_df$pg == "Most Adverse", c("sample", "proportion")]
names(fav)[2] <- "p_fav"
names(adv)[2] <- "p_adv"
bar_stack <- merge(data.frame(sample = sample_lev, stringsAsFactors = FALSE), fav, by = "sample", all.x = TRUE)
bar_stack <- merge(bar_stack, adv, by = "sample", all.x = TRUE)
bar_stack$p_fav[is.na(bar_stack$p_fav)] <- 0
bar_stack$p_adv[is.na(bar_stack$p_adv)] <- 0
## merge() sorts by `by`; restore cohort order so x positions match p_pbulk and dot plots
bar_stack <- bar_stack[match(sample_lev, as.character(bar_stack$sample)), , drop = FALSE]
bar_stack$sample <- factor(bar_stack$sample, levels = sample_lev)

# Stacked bar: per-sample fraction of cohort tail cells (from define_phenotype_groups above)
bar_long <- rbind(
  data.frame(sample = bar_stack$sample, pg = "Most Favorable", proportion = bar_stack$p_fav),
  data.frame(sample = bar_stack$sample, pg = "Most Adverse", proportion = bar_stack$p_adv)
)
bar_long$pg <- factor(bar_long$pg, levels = c("Most Favorable", "Most Adverse"))
bar_long$sample <- factor(as.character(bar_long$sample), levels = sample_lev)

p_bar <- ggplot(bar_long, aes(x = sample, y = proportion, fill = pg)) +
  geom_col(width = 0.85, position = "stack") +
  scale_fill_manual(values = c(`Most Adverse` = "#B2182B", `Most Favorable` = "#2166AC"),
                    name = "Prognostic group") +
  scale_x_discrete(drop = FALSE) +
  scale_y_continuous(expand = c(0, 0)) +
  theme_minimal() +
  theme(
    axis.text.x  = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.x = element_blank(),
    legend.position = "none"
  ) +
  labs(y = "Proportion\nof cells\n(in 5th %)", title = NULL)

meta_plot_dots <- meta_plot[meta_plot$prognostic_grp %in% c("Most Adverse", "Most Favorable"), ]
p_dots <- NULL
p_dots_orig <- NULL
if (nrow(meta_plot_dots) > 0) {
  ## Full `sample_lev` level order so x positions match p_pbulk / p_bar (subset can drop unused levels)
  meta_plot_dots$sample <- factor(as.character(meta_plot_dots$sample), levels = sample_lev)
  meta_plot_dots$cell_type <- factor(meta_plot_dots$cell_type)
  counts <- as.data.frame(table(meta_plot_dots$sample, meta_plot_dots$cell_type, meta_plot_dots$prognostic_grp, dnn = c("sample", "cell_type", "pg")))
  totals <- aggregate(counts$Freq, by = list(sample = counts$sample, pg = counts$pg), FUN = sum)
  names(totals)[3] <- "total"
  counts <- merge(counts, totals, by = c("sample", "pg"))
  counts$proportion <- counts$Freq / counts$total
  counts$proportion[counts$total == 0] <- 0
  counts$sample <- factor(as.character(counts$sample), levels = sample_lev)
  counts$x_num <- as.numeric(counts$sample) + ifelse(counts$pg == "Most Adverse", -0.18, 0.18)

  p_dots <- ggplot(counts, aes(x = x_num, y = cell_type, size = proportion, color = pg)) +
    geom_point(alpha = 0.85) +
    scale_color_manual(values = c(`Most Adverse` = "#B2182B", `Most Favorable` = "#2166AC"), name = "Prognostic group") +
    scale_size_continuous(range = c(0, 5), name = "Proportion of cell type\nper sample in\nprognostic groups") +
    scale_x_continuous(
      breaks = seq_along(sample_lev),
      labels = sample_lev,
      limits = c(0.5, length(sample_lev) + 0.5),
      expand = c(0, 0)
    ) +
    theme_minimal() +
    theme(panel.grid.major.y = element_line(color = "grey90"), 
          axis.title = element_text(size = 10), 
          axis.text.x = element_blank(), 
          axis.ticks.x = element_blank(), 
          legend.title = element_text(hjust = 0.5),
          legend.position = "right") +
    guides(
  color = guide_legend(override.aes = list(size = 3)),
  # size  = guide_legend(override.aes = list(size = c(2, 3, 4, 5, 6)))
) +
    labs(x = NULL, y = "Cell type\n(broad)", title = NULL)

  if (!all(is.na(meta_plot_dots$cell_type_original))) {
    meta_plot_dots$cell_type_original <- factor(meta_plot_dots$cell_type_original)
    counts_orig <- as.data.frame(table(meta_plot_dots$sample, meta_plot_dots$cell_type_original, meta_plot_dots$prognostic_grp, dnn = c("sample", "cell_type_original", "pg")))
    totals_orig <- aggregate(counts_orig$Freq, by = list(sample = counts_orig$sample, pg = counts_orig$pg), FUN = sum)
    names(totals_orig)[3] <- "total"
    counts_orig <- merge(counts_orig, totals_orig, by = c("sample", "pg"))
    counts_orig$proportion <- counts_orig$Freq / counts_orig$total
    counts_orig$proportion[counts_orig$total == 0] <- 0
    counts_orig$sample <- factor(as.character(counts_orig$sample), levels = sample_lev)
    counts_orig$x_num <- as.numeric(counts_orig$sample) + ifelse(counts_orig$pg == "Most Adverse", -0.18, 0.18)

    p_dots_orig <- ggplot(counts_orig, aes(x = x_num, y = cell_type_original, size = proportion, color = pg)) +
      geom_point(alpha = 0.85) +
      scale_color_manual(values = c(`Most Adverse` = "#B2182B", `Most Favorable` = "#2166AC"), name = "Prognostic group") +
      scale_size_continuous(range = c(0, 5), name = "Proportion") +
      scale_x_continuous(
        breaks = seq_along(sample_lev),
        labels = sample_lev,
        limits = c(0.5, length(sample_lev) + 0.5),
        expand = c(0, 0)
      ) +
      theme_minimal() +
      theme(panel.grid.major.y = element_line(color = "grey90"), axis.title = element_text(size = 10),
            axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
            legend.position = "none") +
      labs(x = "Sample", y = "Cell type\n(detailed)", title = NULL) +
      guides(color = guide_legend(override.aes = list(size = 3)))
  }
} else {
  warning("No cells labeled Most Adverse or Most Favorable after define_phenotype_groups; check PRECOG scores (e.g. all NA). Cell-type dot panels omitted.")
}

if (requireNamespace("patchwork", quietly = TRUE)) {
  suppressPackageStartupMessages(library(patchwork))
  if (!is.null(p_dots_orig)) {
    print(p_pbulk / p_bar / p_dots / p_dots_orig + plot_layout(heights = c(0.75, 1, 1, 2)))
  } else if (!is.null(p_dots)) {
    print(p_pbulk / p_bar / p_dots + plot_layout(heights = c(0.75, 1, 2)))
  } else {
    print(p_pbulk / p_bar + plot_layout(heights = c(0.75, 1)))
  }
} else {
  print(p_pbulk)
  print(p_bar)
  if (!is.null(p_dots)) print(p_dots)
  if (!is.null(p_dots_orig)) print(p_dots_orig)
}
```

## Conclusions

Here, we demonstrate that using **`PhenoMapR`** on single-cell RNAseq data in a pancreatic cancer dataset successfully identified cells known to be associated with disease (primarily Malignant and Ductal type 2). These results agree with those of [Jolasun et al.](https://www.nature.com/articles/s41467-025-66162-4/figures/2), however our **`PhenoMapR`** results provide an increased level of granularity compared to other methods, since we retain absolute and rank-ordered information regarding all cell's phenotype association. We also nominate cells most associated with favorable outcomes in PAAD, highlighting potential areas for additional therapeutic focus.

## References

<a id="ref1"></a>
**[1]** Peng, J.  et al. Single-cell RNA-seq highlights intra-tumoral heterogeneity and malignant progression in pancreatic ductal adenocarcinoma. *Cell Research* 29, 725–738 (2019). [https://doi.org/10.1038/s41422-019-0195-y](https://doi.org/10.1038/s41422-019-0195-y)

<a id="ref2"></a>
**[2]** Han, Y.  et al. TISCH2: expanded datasets and new tools for single-cell transcriptome analyses of the tumor microenvironment. *Nucleic Acids Res* 51, D1425–D1431 (2023). [https://doi.org/10.1093/nar/gkac959](https://doi.org/10.1093/nar/gkac959)

<a id="ref4"></a>
**[3]** Benard, B. et al. PRECOG update: an augmented resource of clinical outcome associations with gene expression for adult, pediatric, and immunotherapy cohorts. *Nucleic Acids Res.* 54, D1579–D1589 (2026). [https://doi.org/10.1093/nar/gkaf1215](https://doi.org/10.1093/nar/gkaf1215)

<a id="ref4"></a>
**[4]** Jolasun, Y. et al. SIDISH integrates single-cell and bulk transcriptomics to identify high-risk cells and guide precision therapeutics through in silico perturbation. *Nat Commun* 16, 11271 (2025). [https://doi.org/10.1038/s41467-025-66162-4](https://doi.org/10.1038/s41467-025-66162-4)

## Session Info

```{r session-info}
sessionInfo()
```
