---
title: "Scoring spatial transcriptomics data with PhenoMapR"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_collapsed: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, out.width = "100%")
```

## Overview

This vignette demonstrates **{PhenoMapR}** on spatial transcriptomics in two parts. **Part 1** uses **spot-level** 10X Visium data (`HT270P1-S1H2Fc2U1Z1Bs1-H2Bs2-Test.hgnc.rds`): we show the **H&E** with `SpatialPlot`, then score with PRECOG **Pancreatic** and map **z-scaled** scores with **`stat_summary_hex`** (same coordinates and aspect as H&E; no jitter). **Part 2** uses the same sample after [**CytoSPACE**](https://www.nature.com/articles/s41587-023-01697-9) mapped single cells onto spots (`HT270P1-S1H2Fc2U1Z1Bs1-H2Bs2-Test_processed.rds`); we then mirror the single-cell workflow (score by cell type, markers, heatmaps) at **cell** resolution and assess **co-localization** of prognostic groups with **[`spatialCooccur`](https://github.com/juninamo/spatialCooccur)** (**`nhood_enrichment()`**, **`cooccur_local()`**).

The sample is from a pancreatic cancer dataset available from [HTAN](https://humantumoratlas.org/).

## Download data and vizualize H&E histology
```{r load-setup}
suppressPackageStartupMessages({
  library(PhenoMapR)
  library(Seurat)
  library(SeuratObject)
  library(ggplot2)
  library(dplyr)
})

knitr::opts_chunk$set(fig.width = 12, out.width = "100%", warning = FALSE)
theme_set(theme_minimal(base_size = 14))

options(googledrive_quiet = TRUE)
googledrive::drive_deauth()

gd_id_spot <- "1OkIr7ksAWxKVjtdlGqYHMidvHZZsySEE"
rds_spot <- "HT270P1-S1H2Fc2U1Z1Bs1-H2Bs2-Test.hgnc.rds"

googledrive::drive_download(googledrive::as_id(gd_id_spot), rds_spot, overwrite = TRUE)

seurat_spot <- readRDS("HT270P1-S1H2Fc2U1Z1Bs1-H2Bs2-Test.hgnc.rds")

SpatialPlot(object = seurat_spot,
  features = NULL,
  image.alpha = 1,
  pt.size.factor = 0) +
  guides(fill = "none") 
```

## Score spots with PhenoMapR
```{r score-spots-part1}
# Score spots
scores_spot <- PhenoMapR::PhenoMap(
    expression = seurat_spot,
    reference = "precog",
    cancer_type = "Pancreatic",
    assay = "Spatial",
    slot = "data",
    verbose = FALSE
  )

# Add scores to Seurat object
seurat_spot <- PhenoMapR::add_scores_to_seurat(seurat_spot, scores_spot)

score_col_spot <- grep("weighted_sum_score", names(scores_spot), value = TRUE)[1]
## Z-scaled PhenoMapR score on spots (for hex / metadata plots)
sc_sp <- suppressWarnings(as.numeric(seurat_spot@meta.data[[score_col_spot]]))
seurat_spot$phenomapr_scaled_spot <- as.numeric(scale(sc_sp))

plot_score_distribution(
  seurat_spot$phenomapr_scaled_spot,
  main = "PRECOG Pancreatic score distribution (Visium spots)",
  base_size = 14
)
```

## PhenoMapR score distribution across spots 
```{r}
p <- SpatialPlot(
  object = seurat_spot,
  features = "weighted_sum_score_Pancreatic", image.alpha = 0
)
cols_pheno <- grDevices::colorRampPalette(c("#2166AC", "#F7F7F7", "#B2182B"))(100)

df <- p$data  # extract data from SpatialPlot

spot_hex_bins <- max(10, min(50L, as.integer(round(sqrt(nrow(df))))))

ggplot(df, aes(
  x = x,
  y = -y,
  z = scale(weighted_sum_score_Pancreatic)
)) +
  stat_summary_hex(fun = mean, bins = spot_hex_bins) +
  scale_fill_gradient2(
    low = "#2166AC",
    mid = "#F7F7F7",
    high = "#B2182B",
    midpoint = 0,
    trans = scales::pseudo_log_trans(sigma = 0.2),
    # limits = c(-4, 4),
    name = "PhenoMapR\nz-score"
  )+
  coord_fixed() +
  theme_void()
```

```{r}
df <- SpatialPlot(
  object = seurat_spot,
  features = "weighted_sum_score_Pancreatic", image.alpha = 0
)$data %>%
  as.data.frame()

groups_spot <- PhenoMapR::define_phenotype_groups(df, score_columns = "weighted_sum_score_Pancreatic", percentile = 0.05)

df <- left_join(df, groups_spot, by = c("cell" = "cell_id"))

# Helper function for mode
mode_val <- function(z) {
  z <- z[is.finite(z)]
  if (!length(z)) return(NA_real_)
  as.numeric(names(sort(table(z), decreasing = TRUE)[1L]))
}

ggplot(df, aes(
  x = x,
  y = -y,
  z = as.numeric(factor(
    phenotype_group_weighted_sum_score_Pancreatic,
    levels = c("Most Favorable", "Other", "Most Adverse")
  ))
)) +
  stat_summary_hex(
    aes(
      fill = after_stat(factor(
        value,
        levels = 1:3,
        labels = c("Most Favorable", "Other", "Most Adverse")
      ))
    ),
    bins = 50,
    fun = mode_val,
    colour = NA
  ) +
  scale_fill_manual(
    values = c(
      "Most Favorable" = "#2166AC",
      "Other"            = "#F7F7F7",
      "Most Adverse"     = "#B2182B"
    ),
    drop = FALSE,
    name = "PhenoMapR\nGroup"
  ) +
  coord_fixed() +
  theme_void()
```
Here, we can see that the most favorable spots seem to cluster together, while a subset of the most adverse also tend to co-localize.


# Part 2: CytoSPACE-mapped cells

Load the **CytoSPACE** object (single cells placed on Visium coordinates).

```{r load-cytospace-object}
rds_cyto <- "HT270P1-S1H2Fc2U1Z1Bs1-H2Bs2-Test_processed.rds"
gd_id_cyto <- "1gcOyLriW9bIFNbDuQN6Vi1UsrMGKDxll"

googledrive::drive_download(googledrive::as_id(gd_id_cyto), rds_cyto, overwrite = TRUE)

seurat <- readRDS(rds_cyto)
```

## Score sample with PhenoMapR (CytoSPACE)

```{r score-spots}
  scores_spatial <- PhenoMapR::PhenoMap(
    expression = seurat,
    reference = "precog",
    cancer_type = "Pancreatic",
    assay = "Spatial",
    slot = "data",
    verbose = FALSE
  )
  seurat <- PhenoMapR::add_scores_to_seurat(seurat, scores_spatial)
score_col <- grep("weighted_sum_score", names(scores_spatial), value = TRUE)[1]

plot_score_distribution(
  seurat@meta.data[[score_col]],
  main = "PRECOG Pancreatic score distribution (CytoSPACE cells)",
  base_size = 14
)
```

## Score by cell type

Before looking at any spatial information, we plot the **PhenoMapR** score by cell type to see cell types most enriched in the adverse and favorable prognostic groups.

```{r score-by-annotation, fig.width=7.5, fig.height=5}
spatial_celltype_pal <- NULL
spatial_celltype_col <- NULL
meta_names <- names(seurat@meta.data)
celltype_col <- "CellType"

df <- seurat@meta.data
n_meta <- nrow(df)
ann_vec <- NULL
if (!is.null(celltype_col)) {
  raw <- df[[celltype_col]]
  if (is.list(raw)) {
    ann_vec <- vapply(raw, function(x) if (is.null(x) || length(x) == 0) NA_character_ else as.character(x)[1], character(1))
  } else {
    ann_vec <- as.vector(raw)
  }
  if (length(ann_vec) != n_meta) ann_vec <- NULL
}

  df$annotation <- factor(ann_vec, exclude = NULL)

  pal <- PhenoMapR::get_celltype_palette(levels(df$annotation))
print(ggplot(df, aes(
    x = reorder(.data$annotation, .data[[score_col]], FUN = median),
    y = .data[[score_col]],
    fill = .data$annotation
  )) +
    geom_boxplot(outlier.alpha = 0.3) +
    scale_fill_manual(values = pal, name = celltype_col) +
    coord_cartesian(ylim = c(-10000, 15000)) +
    guides(fill = guide_legend(ncol = 2)) +
    theme_minimal(base_size = 14) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
      legend.position = "right",
      plot.title = element_text(hjust = 0.5)
    ) +
    labs(y = "PRECOG Pancreatic score", x = celltype_col, title = "PhenoMapR Score by Cell-Type"))
```
As seen in the other pancreatic cancer vignettes, the ductal cell type is the most associated with the adverse prognostic signal, while the beta and delta cells are most associated with favorable signal. Interestingly, plasma cells show quite a wide range of score and contain some of the most favorably prognostic cells across the sample. 



## Where are the different cell types?

Here, we use the coordinates information from the spatial seurat object and pair them with the cell level metadata in order to plot the results in spatial context.

```{r}
cell_locations <- seurat@meta.data %>%
  as.data.frame() %>%
  dplyr::select("Cell", "row", "col", 
                "CellType", "weighted_sum_score_Pancreatic") %>%
    group_by(.data$row, .data$col) %>%
    mutate(points_per_location = n()) %>%
    ungroup()

cell_locations$CellType <- as.factor(cell_locations$CellType)

  # Shared jitter and point size so all three spatial plots match (multiple cells per spot)
  rng_row <- diff(range(cell_locations$row, na.rm = TRUE))
  rng_col <- diff(range(-cell_locations$col, na.rm = TRUE))
  spatial_jitter_w <- max(0.25, if (rng_row > 0) rng_row * 0.025 else 0.15)
  spatial_jitter_h <- max(0.25, if (rng_col > 0) rng_col * 0.025 else 0.15)
  spatial_point_range <- c(0.5, 1.6)

  spatial_celltype_pal <- PhenoMapR::get_celltype_palette(levels(cell_locations$CellType))
  
    ct_freq <- sort(table(cell_locations$CellType, useNA = "no"), decreasing = TRUE)
  ct_order <- names(ct_freq)
  cell_locations$celltype_zorder <- as.numeric(factor(as.character(cell_locations$CellType), levels = ct_order))
  ct_pal <- if (!is.null(spatial_celltype_pal)) spatial_celltype_pal else PhenoMapR::get_celltype_palette(levels(cell_locations$CellType))
 
   ggplot(cell_locations, aes(x = .data$row, y = -.data$col, color = .data$CellType,
    size = points_per_location, zorder = .data$celltype_zorder)) +
    geom_jitter(alpha = 0.8, width = spatial_jitter_w, height = spatial_jitter_h, shape = 16) +
    scale_color_manual(values = ct_pal, name = "Cell Type", na.value = "grey90") +
    scale_size_continuous(range = spatial_point_range, trans = "reverse", name = "Cells per spot") +
    guides(
      color = guide_legend(override.aes = list(size = 4), ncol = 2)
      # size  = guide_legend(override.aes = list(size = 4))
    ) +
    theme_minimal(base_size = 14) +
    theme(
      plot.title  = element_blank(),
      axis.text   = element_blank(),
      axis.ticks  = element_blank(),
      axis.title  = element_blank()
    ) +
       coord_fixed(ratio = 0.6) +
  theme_void()
  
```



### Where raw PhenoMapR scores are

Spatial map of PhenoMapR score (z-scaled for color gradient). Blue = more favorable, red = more adverse.

```{r}
cell_locations <- cell_locations[order(abs(cell_locations$weighted_sum_score_Pancreatic)), ]

ggplot(cell_locations, aes(
  x = .data$row,
  y = -.data$col,
  color = scale(weighted_sum_score_Pancreatic),
  size = .data$points_per_location
)) +
  geom_jitter(alpha = 0.8, width = spatial_jitter_w, height = spatial_jitter_h, shape = 16) +
  scale_size_continuous(range = spatial_point_range, trans = "reverse", name = "Cells per spot") +
  scale_color_gradient2(
    low = "#2166AC", mid = "#F7F7F7", high = "#B2182B",
    midpoint = 0,
    trans = scales::pseudo_log_trans(sigma = 0.2),
    name = "PhenoMapR\nz-score"
  ) +
  coord_fixed(ratio = 0.6) +
  theme_void()
```

### Where 5th percentile cells are

In order to make the visualization more apparent, we restrict the spatial map of prognostic groups to: top 5% (Most Adverse), bottom 5% (Most Favorable), and the rest (Other).

```{r}
cell_locations <- cell_locations %>% 
  mutate(percentile = percent_rank(weighted_sum_score_Pancreatic)) %>%
  mutate(prognostic_group = case_when(
    percentile < 0.05 ~ "Most Favorable",
    percentile >= 0.95 ~ "Most Adverse",
    TRUE ~ "Other"
  ))

  df_other <- cell_locations %>%
    dplyr::filter(prognostic_group=="Other")
  
  df_extreme <- cell_locations %>%
    dplyr::filter(prognostic_group!="Other")
  
  ggplot() +
    geom_jitter(data = df_other, aes(x = .data$row, y = -.data$col, color = .data$prognostic_group,
      size = points_per_location), alpha = 0.8, width = spatial_jitter_w, height = spatial_jitter_h, shape = 16) +
    geom_jitter(data = df_extreme, aes(x = .data$row, y = -.data$col, color = .data$prognostic_group,
      size = points_per_location), alpha = 0.8, width = spatial_jitter_w, height = spatial_jitter_h, shape = 16) +
    # ggtitle("5th percentile: Most Adverse vs Most Favorable") +
    scale_color_manual(
      values = c(`Most Adverse` = "#B2182B", Other = "#f7f7f7", `Most Favorable` = "#2166AC"),
      name = "Prognostic group",
      na.value = "grey90",
      drop = FALSE
    ) +
     guides(
      color = guide_legend(override.aes = list(size = 4))
      # size  = guide_legend(override.aes = list(size = 4))
    ) +
    scale_size_continuous(range = spatial_point_range, trans = "reverse", name = "Cells per spot") +
    theme_minimal(base_size = 14) +
    theme(
      plot.title = element_text(hjust = 0.5, size = 14),
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      axis.title = element_blank()
    ) +
  coord_fixed(ratio = 0.6) +
  theme_void()

```

<!-- Visually, adverse and favorable cells can appear to segregate on the tissue. The next section quantifies **neighborhood co-occurrence** of prognostic groups with **[`spatialCooccur`](https://github.com/juninamo/spatialCooccur)** (Inamo *et al.*, [medRxiv 2025](https://doi.org/10.1101/2025.08.05.25332835)). -->

<!-- ## Co-localization of prognostic cells (CytoSPACE) -->

<!-- We use the same **Visium row / column** coordinates as the spatial plots, assembled as a **`data.frame`** with **`x`**, **`y`**, and **`prognostic_group`** (and **`cell_type`** as a copy of the latter for **`cooccur_local()`**, which expects that column name). -->

<!-- Install **`spatialCooccur`** from GitHub if needed: `remotes::install_github("juninamo/spatialCooccur")`. -->

<!-- **`spatialCooccur::nhood_enrichment()`** builds a **kNN graph** (**`Seurat::FindNeighbors()`** inside the package), optionally normalizes adjacency, aggregates **co-occurrence** between cluster pairs, and compares the observed matrix to **permutation** nulls to produce a **z-score** matrix. We use **`n_jobs = 1`** so the vignette runs on a single core (set higher locally if you install the package). **`spatialCooccur::cooccur_local()`** scores each cell by whether its **radius** neighborhood contains both **Most Adverse** and **Most Favorable** cells, then applies a short diffusion-style step when **`maxnsteps > 0`**. Parameters **`neighbors.k`**, **`n_perms`**, and **`radius`** are reduced when **`PHENOMAPR_SPATIAL_LITE`** or **CI** is set; duplicate spot coordinates (many CytoSPACE cells per spot) get a small jitter. -->

<!-- ```{r cytospace-spatialcooccur, fig.width=9, fig.height=4.5, out.width=NULL} -->
<!-- scoc_nhood <- NULL -->
<!-- scoc_local <- NULL -->

<!-- if (exists("has_spatial_coords") && isTRUE(has_spatial_coords) && exists("spatial_df") && -->
<!--     all(c("row", "col", "prognostic_group", "cell_id") %in% names(spatial_df)) && -->
<!--     requireNamespace("spatialCooccur", quietly = TRUE)) { -->
<!--   dfb <- spatial_df[ -->
<!--     stats::complete.cases(spatial_df[, c("row", "col", "prognostic_group")]), -->
<!--     , -->
<!--     drop = FALSE -->
<!--   ] -->
<!--   scoc_df <- data.frame( -->
<!--     x = as.numeric(dfb$row), -->
<!--     y = as.numeric(dfb$col), -->
<!--     prognostic_group = as.character(dfb$prognostic_group), -->
<!--     stringsAsFactors = FALSE -->
<!--   ) -->
<!--   rownames(scoc_df) <- as.character(dfb$cell_id) -->
<!--   scoc_df$cell_type <- scoc_df$prognostic_group -->

<!--   max_cells_sc <- if (exists("spatial_ci_lite") && isTRUE(spatial_ci_lite)) 1500L else 4000L -->
<!--   if (nrow(scoc_df) > max_cells_sc) { -->
<!--     set.seed(3L) -->
<!--     scoc_df <- scoc_df[sample.int(nrow(scoc_df), max_cells_sc), , drop = FALSE] -->
<!--   } -->

<!--   xy_mat <- as.matrix(scoc_df[, c("x", "y"), drop = FALSE]) -->
<!--   if (any(duplicated(xy_mat) | duplicated(xy_mat, fromLast = TRUE))) { -->
<!--     rng <- max(diff(range(xy_mat[, 1], na.rm = TRUE)), diff(range(xy_mat[, 2], na.rm = TRUE)), na.rm = TRUE) -->
<!--     eps <- if (is.finite(rng) && rng > 0) rng * 1e-5 else 1e-6 -->
<!--     scoc_df$x <- scoc_df$x + stats::runif(nrow(scoc_df), 0, eps) -->
<!--     scoc_df$y <- scoc_df$y + stats::runif(nrow(scoc_df), 0, eps) -->
<!--   } -->

<!--   k_sc <- if (exists("spatial_ci_lite") && isTRUE(spatial_ci_lite)) 12L else 20L -->
<!--   n_perm_sc <- if (exists("spatial_ci_lite") && isTRUE(spatial_ci_lite)) 40L else 100L -->
<!--   k_sc <- min(k_sc, max(5L, nrow(scoc_df) - 1L)) -->

<!--   n_grp <- length(unique(scoc_df$prognostic_group)) -->
<!--   if (nrow(scoc_df) >= (k_sc + 3L) && n_grp >= 2L) { -->
<!--     scoc_nhood <- tryCatch( -->
<!--       spatialCooccur::nhood_enrichment( -->
<!--         scoc_df, -->
<!--         cluster_key = "prognostic_group", -->
<!--         neighbors.k = k_sc, -->
<!--         connectivity_key = "nn", -->
<!--         transformation = TRUE, -->
<!--         n_perms = n_perm_sc, -->
<!--         seed = 42L, -->
<!--         n_jobs = 1L -->
<!--       ), -->
<!--       error = function(e) { -->
<!--         message("spatialCooccur::nhood_enrichment failed: ", conditionMessage(e)) -->
<!--         NULL -->
<!--       } -->
<!--     ) -->
<!--   } -->

<!--   rad_sc <- max(3, 0.04 * max( -->
<!--     diff(range(scoc_df$x, na.rm = TRUE)), -->
<!--     diff(range(scoc_df$y, na.rm = TRUE)), -->
<!--     na.rm = TRUE -->
<!--   )) -->
<!--   if (nrow(scoc_df) >= 20L && -->
<!--       sum(scoc_df$prognostic_group == "Most Adverse") >= 3L && -->
<!--       sum(scoc_df$prognostic_group == "Most Favorable") >= 3L) { -->
<!--     scoc_local <- tryCatch( -->
<!--       spatialCooccur::cooccur_local( -->
<!--         scoc_df, -->
<!--         cluster_x = "Most Adverse", -->
<!--         cluster_y = "Most Favorable", -->
<!--         connectivity_key = "nn", -->
<!--         neighbors.k = k_sc, -->
<!--         radius = rad_sc, -->
<!--         maxnsteps = if (exists("spatial_ci_lite") && isTRUE(spatial_ci_lite)) 1L else 3L -->
<!--       ), -->
<!--       error = function(e) { -->
<!--         message("spatialCooccur::cooccur_local failed: ", conditionMessage(e)) -->
<!--         NULL -->
<!--       } -->
<!--     ) -->
<!--   } -->
<!-- } else if (!requireNamespace("spatialCooccur", quietly = TRUE)) { -->
<!--   message( -->
<!--     "Install spatialCooccur for co-localization: remotes::install_github(\"juninamo/spatialCooccur\")" -->
<!--   ) -->
<!-- } -->

<!-- if (!is.null(scoc_nhood) && !is.null(scoc_nhood$zscore)) { -->
<!--   zm <- as.matrix(scoc_nhood$zscore) -->
<!--   lab_clean <- function(nm) gsub("^Cluster", "", nm) -->
<!--   ij <- expand.grid(i = seq_len(nrow(zm)), j = seq_len(ncol(zm)), stringsAsFactors = FALSE) -->
<!--   zlong <- data.frame( -->
<!--     from = lab_clean(rownames(zm)[ij$i]), -->
<!--     to = lab_clean(colnames(zm)[ij$j]), -->
<!--     z = zm[cbind(ij$i, ij$j)], -->
<!--     stringsAsFactors = FALSE -->
<!--   ) -->
<!--   print(scoc_nhood$count) -->
<!--   p_nh <- ggplot(zlong, aes(x = .data$to, y = .data$from, fill = .data$z)) + -->
<!--     geom_tile(color = "white", linewidth = 0.2) + -->
<!--     scale_fill_gradient2( -->
<!--       low = "#2166AC", -->
<!--       mid = "white", -->
<!--       high = "#B2182B", -->
<!--       midpoint = 0, -->
<!--       name = "Z" -->
<!--     ) + -->
<!--     labs( -->
<!--       title = "Neighborhood enrichment (spatialCooccur::nhood_enrichment)", -->
<!--       subtitle = "Z-scores for co-occurrence on kNN graph vs permuted labels", -->
<!--       x = "Neighbor label", -->
<!--       y = "Cell label" -->
<!--     ) + -->
<!--     theme_minimal(base_size = 11) + -->
<!--     theme(axis.text.x = element_text(angle = 35, hjust = 1)) -->
<!--   print(p_nh) -->
<!-- } -->

<!-- if (!is.null(scoc_local) && ncol(scoc_local) >= 1L) { -->
<!--   cn <- colnames(scoc_local)[1] -->
<!--   loc_vec <- scoc_local[[1]] -->
<!--   # rownames() hold barcodes; names(loc_vec) is often NULL after diffusion (as.numeric(matrix)) -->
<!--   nr <- length(loc_vec) -->
<!--   cell_ids <- rownames(scoc_local) -->
<!--   if (length(cell_ids) != nr) cell_ids <- names(loc_vec) -->
<!--   if (is.null(cell_ids) || length(cell_ids) != nr) { -->
<!--     cell_ids <- if (exists("scoc_df") && nrow(scoc_df) == nr) rownames(scoc_df) else as.character(seq_len(nr)) -->
<!--   } -->
<!--   loc_df <- data.frame( -->
<!--     score = as.numeric(loc_vec), -->
<!--     cell = as.character(cell_ids), -->
<!--     stringsAsFactors = FALSE -->
<!--   ) -->
<!--   if (exists("scoc_df")) { -->
<!--     loc_df$prognostic_group <- scoc_df$prognostic_group[match(loc_df$cell, rownames(scoc_df))] -->
<!--   } -->
<!--   p_loc <- ggplot(loc_df, aes(x = stats::reorder(.data$prognostic_group, .data$score, median), y = .data$score, fill = .data$prognostic_group)) + -->
<!--     geom_boxplot(outlier.alpha = 0.3, show.legend = FALSE) + -->
<!--     labs( -->
<!--       title = "Local co-occurrence score (spatialCooccur::cooccur_local)", -->
<!--       subtitle = paste0("Neighborhoods (radius ≈ ", signif(rad_sc, 3), ") containing both Most Adverse and Most Favorable cells"), -->
<!--       x = "Prognostic group", -->
<!--       y = cn -->
<!--     ) + -->
<!--     theme_minimal(base_size = 11) + -->
<!--     theme(axis.text.x = element_text(angle = 30, hjust = 1)) -->
<!--   print(p_loc) -->
<!-- } -->

<!-- if (is.null(scoc_nhood) && requireNamespace("spatialCooccur", quietly = TRUE)) { -->
<!--   message("spatialCooccur neighborhood enrichment skipped (need enough cells and distinct prognostic groups).") -->
<!-- } -->
<!-- ``` -->


<!-- ## Prognostic markers -->

<!-- `find_phenotype_markers()` supports the same **marker scopes** on spatial data (cells or spots mapped to spots) as in the single-cell vignette: -->

<!-- - **Cell type agnostic** — `marker_scope = "phenotype_groups"` (**default**): **Most Adverse** vs all other cells with a phenotype label, and **Most Favorable** vs all other cells, **ignoring** cell type. This asks which genes distinguish the prognostic extremes in the whole mixture on the slide. -->
<!-- - **Cell type specific** — `marker_scope = "cell_type_specific"`: for each cell type in metadata (here, CytoSPACE-mapped types), contrasts that type in the adverse or favorable tail vs a reference set controlled by `celltype_contrast`. **Default** `celltype_contrast = "within_cell_type"` compares to other phenotype bins **within the same type** (reduces housekeeping genes repeating across type × tail blocks). **`celltype_contrast = "vs_cohort_rest"`** restores the **original** cohort-wide contrast: (type ∩ tail) vs **every other cell**. With **`celltype_unique_genes = TRUE`** (default), each gene is kept in at most one adverse/favorable row (strongest `avg_log2FC`). -->

<!-- Below, **Step 1–2** use **cell-type-agnostic** markers; **Step 2** draws **`plot_phenotype_markers(..., heatmap_type = "global")`** when **ComplexHeatmap** and **circlize** are installed (same helper as the **single-cell** vignette), with a **pheatmap** fallback otherwise. **Step 3** runs **cell-type-specific** markers and **`plot_phenotype_markers(..., heatmap_type = "cell_type_specific")`** when those packages are available. -->

<!-- ### Step 1: Markers for adverse vs. favorable cells (cell type agnostic) -->

<!-- ```{r markers-spatial, cache = FALSE} -->
<!-- markers <- NULL -->
<!-- assay_markers <- NULL -->
<!-- if (!is.null(group_col)) { -->
<!--   cells <- colnames(seurat) -->
<!--   group_vec <- seurat@meta.data[cells, group_col] -->
<!--   group_df <- data.frame( -->
<!--     cell_id = cells, -->
<!--     phenotype_group = as.character(group_vec), -->
<!--     stringsAsFactors = FALSE -->
<!--   ) -->
<!--   for (a in unique(c(assay_use, "RNA", "SCT"))) { -->
<!--     if (!a %in% names(seurat@assays)) next -->
<!--     markers <- tryCatch( -->
<!--       PhenoMapR::find_phenotype_markers( -->
<!--         seurat, -->
<!--         group_labels = group_df, -->
<!--         group_column = "phenotype_group", -->
<!--         cell_id_column = "cell_id", -->
<!--         marker_scope = "phenotype_groups", -->
<!--         assay = a, -->
<!--         slot = "data", -->
<!--         max_cells_per_ident = spatial_max_cells, -->
<!--         verbose = FALSE -->
<!--       ), -->
<!--       error = function(e) { -->
<!--         message("find_phenotype_markers (assay ", a, ") error: ", conditionMessage(e)) -->
<!--         NULL -->
<!--       } -->
<!--     ) -->
<!--     if (!is.null(markers) && (nrow(markers$adverse_markers) > 0 || nrow(markers$favorable_markers) > 0)) { -->
<!--       assay_markers <- a -->
<!--       break -->
<!--     } -->
<!--   } -->
<!--   if (!is.null(markers)) { -->
<!--     message("Adverse markers (top 5):") -->
<!--     print(head(markers$adverse_markers, 5)) -->
<!--     message("Favorable markers (top 5):") -->
<!--     print(head(markers$favorable_markers, 5)) -->
<!--   } else { -->
<!--     message("Marker analysis returned no results. Check that Most Adverse and Most Favorable groups have enough cells.") -->
<!--   } -->
<!-- } -->
<!-- ``` -->

<!-- ### Step 2: Heatmap of adverse vs. favorable markers (global phenotype groups) -->

<!-- **`plot_phenotype_markers(..., heatmap_type = "global")`** matches the single-cell vignette: favorable- then adverse-marker rows, column order by PhenoMapR score, and **pheatmap** only if ComplexHeatmap is unavailable or marker tables are incomplete. -->

<!-- ```{r heatmap-markers-spatial, fig.align='center', fig.width=10, fig.height=4} -->
<!-- if (exists("markers") && !is.null(markers)) { -->
<!--   n_top <- if (exists("spatial_ci_lite") && isTRUE(spatial_ci_lite)) 10L else 15L -->
<!--   expr <- NULL -->
<!--   assay_order <- unique(c( -->
<!--     if (exists("assay_markers") && !is.null(assay_markers)) assay_markers else character(0), -->
<!--     assay_use, "RNA", "SCT" -->
<!--   )) -->
<!--   for (a in assay_order) { -->
<!--     if (!a %in% names(seurat@assays)) next -->
<!--     expr <- tryCatch( -->
<!--       Seurat::GetAssayData(seurat, layer = "data", assay = a), -->
<!--       error = function(e) tryCatch(Seurat::GetAssayData(seurat, slot = "data", assay = a), error = function(e2) NULL) -->
<!--     ) -->
<!--     if (!is.null(expr) && nrow(expr) > 0 && ncol(expr) > 0) break -->
<!--     expr <- tryCatch( -->
<!--       SeuratObject::LayerData(seurat, layer = "data", assay = a), -->
<!--       error = function(e) NULL -->
<!--     ) -->
<!--     if (is.null(expr) || ncol(expr) == 0) { -->
<!--       expr <- tryCatch( -->
<!--         SeuratObject::LayerData(seurat, layer = "counts", assay = a), -->
<!--         error = function(e) tryCatch(SeuratObject::LayerData(seurat, layer = NULL, assay = a), error = function(e2) NULL) -->
<!--       ) -->
<!--     } -->
<!--     if (!is.null(expr) && nrow(expr) > 0 && ncol(expr) > 0) break -->
<!--   } -->
<!--   if (!is.null(expr) && ncol(expr) > 0) { -->
<!--     cells_expr <- colnames(expr) -->
<!--     if (is.null(cells_expr)) cells_expr <- character(0) -->
<!--     cells_obj <- colnames(seurat) -->
<!--     cells_meta <- rownames(seurat@meta.data) -->
<!--     cells_use <- intersect(cells_expr, cells_obj) -->
<!--     if (length(cells_use) == 0 && length(cells_meta) > 0) { -->
<!--       cells_use <- intersect(cells_expr, cells_meta) -->
<!--     } -->
<!--     if (length(cells_use) == 0 && length(cells_expr) > 0 && any(grepl("_", cells_expr, fixed = TRUE))) { -->
<!--       stripped <- sub("^[^_]+_", "", cells_expr) -->
<!--       cells_use <- intersect(stripped, cells_obj) -->
<!--       if (length(cells_use) == 0) cells_use <- intersect(stripped, cells_meta) -->
<!--       if (length(cells_use) > 0) { -->
<!--         expr_cols <- vapply(cells_use, function(c) cells_expr[which(stripped == c)[1]], character(1)) -->
<!--         expr <- expr[, expr_cols, drop = FALSE] -->
<!--         colnames(expr) <- cells_use -->
<!--       } -->
<!--     } -->
<!--     if (length(cells_use) == 0) { -->
<!--       message("No overlapping cells between expression and metadata; skipping heatmap. ", -->
<!--               "expr ncol=", length(cells_expr), ", obj ncol=", length(cells_obj), -->
<!--               if (length(cells_expr) > 0) paste0("; expr sample: ", head(cells_expr, 2)) else "") -->
<!--     } else { -->
<!--       expr_pm <- as.matrix(expr[, cells_use, drop = FALSE]) -->
<!--       meta_pm <- seurat@meta.data[cells_use, , drop = FALSE] -->
<!--       meta_pm$cell_id_plot <- cells_use -->
<!--       group_col_hm <- paste0("phenotype_group_", score_col) -->
<!--       if (!group_col_hm %in% names(meta_pm)) group_col_hm <- group_col -->

<!--       ct_col_hm <- NULL -->
<!--       if (exists("spatial_df_celltype_col") && !is.null(spatial_df_celltype_col) && -->
<!--           spatial_df_celltype_col %in% names(meta_pm)) { -->
<!--         ct_col_hm <- spatial_df_celltype_col -->
<!--       } else if (exists("celltype_col") && !is.null(celltype_col) && celltype_col %in% names(meta_pm)) { -->
<!--         ct_col_hm <- celltype_col -->
<!--       } -->
<!--       if (is.null(ct_col_hm)) { -->
<!--         meta_pm$celltype_plot <- factor(rep("Cell", nrow(meta_pm))) -->
<!--         ct_col_hm <- "celltype_plot" -->
<!--       } else { -->
<!--         meta_pm[[ct_col_hm]] <- factor(as.character(meta_pm[[ct_col_hm]])) -->
<!--       } -->
<!--       pal_ct <- PhenoMapR::get_celltype_palette(levels(meta_pm[[ct_col_hm]])) -->

<!--       drew_ch <- FALSE -->
<!--       if (requireNamespace("ComplexHeatmap", quietly = TRUE) && requireNamespace("circlize", quietly = TRUE)) { -->
<!--         adv_ok <- !is.null(markers$adverse_markers) && nrow(markers$adverse_markers) > 0L -->
<!--         fav_ok <- !is.null(markers$favorable_markers) && nrow(markers$favorable_markers) > 0L -->
<!--         if (adv_ok && fav_ok) { -->
<!--           ch_ht <- PhenoMapR::plot_phenotype_markers( -->
<!--             markers = markers, -->
<!--             expr_mat = expr_pm, -->
<!--             meta = meta_pm, -->
<!--             cell_id_col = "cell_id_plot", -->
<!--             group_col = group_col_hm, -->
<!--             score_col = score_col, -->
<!--             celltype_col = ct_col_hm, -->
<!--             celltype_palette = pal_ct, -->
<!--             heatmap_type = "global", -->
<!--             top_n_markers = if (exists("spatial_ci_lite") && isTRUE(spatial_ci_lite)) 12L else 20L, -->
<!--             n_mark_labels = 15L, -->
<!--             p_adj_threshold = 0.05, -->
<!--             column_title = "Global phenotype markers (spatial, CytoSPACE cells)" -->
<!--           ) -->
<!--           drew_ch <- !is.null(ch_ht) -->
<!--         } -->
<!--       } -->
<!--       if (!drew_ch && requireNamespace("pheatmap", quietly = TRUE)) { -->
<!--         adverse_pos <- markers$adverse_markers -->
<!--         if (is.null(adverse_pos)) adverse_pos <- data.frame() -->
<!--         if (nrow(adverse_pos) > 0 && "avg_log2FC" %in% names(adverse_pos)) adverse_pos <- adverse_pos[adverse_pos$avg_log2FC > 0, ] -->
<!--         favorable_pos <- markers$favorable_markers -->
<!--         if (is.null(favorable_pos)) favorable_pos <- data.frame() -->
<!--         if (nrow(favorable_pos) > 0 && "avg_log2FC" %in% names(favorable_pos)) favorable_pos <- favorable_pos[favorable_pos$avg_log2FC > 0, ] -->
<!--         adv_ref <- markers$adverse_markers -->
<!--         pcol <- if (!is.null(adv_ref) && "p_adj" %in% names(adv_ref)) "p_adj" else if (!is.null(adv_ref) && "p_val_adj" %in% names(adv_ref)) "p_val_adj" else "p_val" -->
<!--         top_genes <- unique(c( -->
<!--           if (nrow(adverse_pos) > 0 && "gene" %in% names(adverse_pos)) head(adverse_pos$gene[order(adverse_pos[[pcol]])], n_top) else character(0), -->
<!--           if (nrow(favorable_pos) > 0 && "gene" %in% names(favorable_pos)) head(favorable_pos$gene[order(favorable_pos[[pcol]])], n_top) else character(0) -->
<!--         )) -->
<!--         top_genes <- top_genes[top_genes %in% rownames(expr_pm)] -->
<!--         if (length(top_genes) == 0) top_genes <- head(rownames(expr_pm), 20) -->
<!--         mat <- as.matrix(expr_pm[top_genes, , drop = FALSE]) -->
<!--         mat_scaled <- t(scale(t(mat))) -->
<!--         mat_scaled[mat_scaled < -3] <- -3 -->
<!--         mat_scaled[mat_scaled > 3] <- 3 -->
<!--         ord <- order(meta_pm[[score_col]]) -->
<!--         mat_scaled <- mat_scaled[, ord, drop = FALSE] -->
<!--         meta_ord <- meta_pm[ord, , drop = FALSE] -->
<!--         ann_col <- data.frame( -->
<!--           `PhenoMapR Score` = meta_ord[[score_col]], -->
<!--           `Prognostic group` = factor(meta_ord[[group_col_hm]], levels = c("Most Adverse", "Other", "Most Favorable")), -->
<!--           check.names = FALSE -->
<!--         ) -->
<!--         if (ct_col_hm %in% names(meta_ord)) { -->
<!--           ann_col$`Cell type` <- factor(meta_ord[[ct_col_hm]]) -->
<!--         } -->
<!--         rownames(ann_col) <- colnames(mat_scaled) -->
<!--         pal_score <- colorRampPalette(c("#2166AC", "#F7F7F7", "#B2182B"))(100) -->
<!--         pal_group <- c(`Most Adverse` = "#B2182B", Other = "#f7f7f7", `Most Favorable` = "#2166AC") -->
<!--         pal_celltype <- if ("Cell type" %in% names(ann_col)) { -->
<!--           PhenoMapR::get_celltype_palette(levels(ann_col$`Cell type`)) -->
<!--         } else list() -->
<!--         ann_colors <- list(`PhenoMapR Score` = pal_score, `Prognostic group` = pal_group) -->
<!--         if (length(pal_celltype) > 0) ann_colors$`Cell type` <- pal_celltype -->
<!--         heatmap_colors <- if (requireNamespace("paletteer", quietly = TRUE)) { -->
<!--           colorRampPalette(paletteer::paletteer_d("MexBrewer::Vendedora"))(100) -->
<!--         } else { -->
<!--           colorRampPalette(c("#2166AC", "#F7F7F7", "#B2182B"))(100) -->
<!--         } -->
<!--         message("Using pheatmap fallback (ComplexHeatmap unavailable or one marker table empty).") -->
<!--         pheatmap::pheatmap( -->
<!--           mat_scaled, -->
<!--           scale = "none", -->
<!--           cluster_cols = FALSE, -->
<!--           cluster_rows = TRUE, -->
<!--           show_colnames = FALSE, -->
<!--           annotation_col = ann_col, -->
<!--           annotation_colors = ann_colors, -->
<!--           color = heatmap_colors, -->
<!--           breaks = seq(-3, 3, length.out = 101), -->
<!--           main = "Top adverse & favorable marker genes (pheatmap fallback)", -->
<!--           fontsize = 12, -->
<!--           fontsize_row = 10, -->
<!--           treeheight_row = 10 -->
<!--         ) -->
<!--       } else if (!drew_ch) { -->
<!--         message("Install ComplexHeatmap + circlize for plot_phenotype_markers; pheatmap for fallback.") -->
<!--       } -->
<!--     } -->
<!--   } else { -->
<!--     msg <- "Could not retrieve expression data for heatmap" -->
<!--     if (!is.null(expr)) { -->
<!--       msg <- paste0(msg, " (expr nrow=", nrow(expr), ", ncol=", ncol(expr), "; try LayerData for Seurat v5)") -->
<!--     } -->
<!--     message(msg, ".") -->
<!--   } -->
<!-- } -->
<!-- ``` -->

<!-- ### Stacked barplot: adverse vs. favorable by cell type -->

<!-- Each column is a cell type; the height is filled by the number of adverse (5th percentile) or favorable (5th percentile) cells. -->

<!-- ```{r stacked-bar-adverse-favorable-by-celltype, fig.width=5, fig.height=5} -->
<!-- if (!is.null(group_col)) { -->
<!--   ct_col_bar <- NULL -->
<!--   if (exists("spatial_df_celltype_col") && !is.null(spatial_df_celltype_col) && spatial_df_celltype_col %in% names(seurat@meta.data)) { -->
<!--     ct_col_bar <- spatial_df_celltype_col -->
<!--   } else if (exists("celltype_col") && !is.null(celltype_col) && celltype_col %in% names(seurat@meta.data)) { -->
<!--     ct_col_bar <- celltype_col -->
<!--   } else { -->
<!--     for (c in c("CellType", "Celltype..major.lineage.", "cell_type", "celltype", "annotation", "seurat_clusters")) { -->
<!--       if (c %in% names(seurat@meta.data) && length(unique(na.omit(seurat@meta.data[[c]]))) >= 2) { -->
<!--         ct_col_bar <- c -->
<!--         break -->
<!--       } -->
<!--     } -->
<!--   } -->
<!--   if (!is.null(ct_col_bar)) { -->
<!--     meta <- seurat@meta.data -->
<!--     meta$pg <- meta[[group_col]] -->
<!--     meta$ct <- as.character(meta[[ct_col_bar]]) -->
<!--     meta$ct_ok <- !is.na(meta$ct) & nzchar(trimws(meta$ct)) -->
<!--     idx_extreme <- meta$pg %in% c("Most Adverse", "Most Favorable") & meta$ct_ok -->
<!--     df_bar <- as.data.frame(table( -->
<!--       CellType = meta$ct[idx_extreme], -->
<!--       Prognostic_group = meta$pg[idx_extreme], -->
<!--       useNA = "no" -->
<!--     )) -->
<!--     if (nrow(df_bar) > 0) { -->
<!--       df_bar$Prognostic_group <- factor(df_bar$Prognostic_group, levels = c("Most Favorable", "Most Adverse")) -->
<!--       pal_bar <- c(`Most Adverse` = "#B2182B", `Most Favorable` = "#2166AC") -->
<!--       adverse <- df_bar[df_bar$Prognostic_group == "Most Adverse", c("CellType", "Freq")] -->
<!--       fav <- df_bar[df_bar$Prognostic_group == "Most Favorable", c("CellType", "Freq")] -->
<!--       names(adverse)[2] <- "n_adverse" -->
<!--       names(fav)[2] <- "n_fav" -->
<!--       df_labels <- merge(adverse, fav, by = "CellType", all = TRUE) -->
<!--       df_labels$n_adverse[is.na(df_labels$n_adverse)] <- 0 -->
<!--       df_labels$n_fav[is.na(df_labels$n_fav)] <- 0 -->
<!--       df_labels$label <- paste0(df_labels$n_adverse, "/", df_labels$n_fav) -->
<!--       df_labels$total <- df_labels$n_adverse + df_labels$n_fav -->
<!--       ct_ord <- levels(reorder(df_bar$CellType, df_bar$Freq, function(x) -sum(x))) -->
<!--       df_labels$CellType <- factor(df_labels$CellType, levels = ct_ord) -->
<!--       p_bar <- ggplot(df_bar, aes(x = reorder(.data$CellType, .data$Freq, function(x) -sum(x)), y = .data$Freq, fill = .data$Prognostic_group)) + -->
<!--         geom_col(position = "stack") + -->
<!--         geom_text(data = df_labels, aes(x = .data$CellType, y = .data$total, label = .data$n_adverse), -->
<!--                   inherit.aes = FALSE, vjust = -0.3, size = 3.5, color = "#B2182B", hjust = 1.05) + -->
<!--         geom_text(data = df_labels, aes(x = .data$CellType, y = .data$total, label = paste0("/", .data$n_fav)), -->
<!--                   inherit.aes = FALSE, vjust = -0.3, size = 3.5, color = "#2166AC", hjust = -0.05) + -->
<!--         scale_fill_manual(values = pal_bar) + -->
<!--         labs(x = "Cell type", y = "Number of cells", fill = "Prognostic group", -->
<!--              title = "Adverse vs. favorable cells (5th percentile) by cell type") + -->
<!--         theme_minimal(base_size = 14) + -->
<!--         theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12), legend.position = "top") -->
<!--       print(p_bar) -->
<!--     } -->
<!--   } -->
<!-- } -->
<!-- ``` -->
<!-- In line with the other Pancreatic cancer vignettes, PhenoMapR identifies ductal cells (and some fibroblasts) as the most associated cell type with an adverse PhenoMapR prognostic score. Alpha, plasma, and beta cells are the most associated with the more favorable prognostic signal. Interestingly, fibroblasts seems to comprise both adverse and favorable phenotypes. -->


<!-- ### Step 3: Cell-type-specific phenotype markers -->

<!-- Here we call `find_phenotype_markers()` with **`marker_scope = "cell_type_specific"`**, using the same cell type column as elsewhere in this vignette. The default **`celltype_contrast = "within_cell_type"`** contrasts each prognostic tail **within** a mapped cell type to the other bins in that type. For the legacy cohort-wide reference (tail vs all other cells), set **`celltype_contrast = "vs_cohort_rest"`** (see `?find_phenotype_markers`). -->

<!-- When both adverse and favorable marker tables are non-empty and **ComplexHeatmap** is installed, **`plot_phenotype_markers(..., heatmap_type = "cell_type_specific")`** matches the single-cell vignette (column order: phenotype bin, cell type, score; row blocks with optional gene labels). -->

<!-- ```{r markers-celltype-specific-spatial, cache = FALSE, fig.align='center', fig.width=14, fig.height=10, eval=!spatial_ci_lite} -->
<!-- markers_ct_spatial <- NULL -->
<!-- ct_col_spatial <- NULL -->
<!-- if (!is.null(group_col)) { -->
<!--   if (exists("spatial_df_celltype_col") && !is.null(spatial_df_celltype_col) && -->
<!--       spatial_df_celltype_col %in% names(seurat@meta.data)) { -->
<!--     ct_col_spatial <- spatial_df_celltype_col -->
<!--   } else if (!is.null(celltype_col) && celltype_col %in% names(seurat@meta.data)) { -->
<!--     ct_col_spatial <- celltype_col -->
<!--   } else { -->
<!--     for (c in c("CellType", "Celltype..major.lineage.", "cell_type", "celltype", "annotation", "seurat_clusters")) { -->
<!--       if (c %in% names(seurat@meta.data) && length(unique(na.omit(seurat@meta.data[[c]]))) >= 2) { -->
<!--         ct_col_spatial <- c -->
<!--         break -->
<!--       } -->
<!--     } -->
<!--   } -->
<!--   if (!is.null(ct_col_spatial)) { -->
<!--     cells_ct <- colnames(seurat) -->
<!--     group_df_ct <- data.frame( -->
<!--       cell_id = cells_ct, -->
<!--       phenotype_group = as.character(seurat@meta.data[cells_ct, group_col]), -->
<!--       cell_type = as.character(seurat@meta.data[cells_ct, ct_col_spatial]), -->
<!--       stringsAsFactors = FALSE -->
<!--     ) -->
<!--     assay_ct <- if (exists("assay_markers") && !is.null(assay_markers)) assay_markers else assay_use -->
<!--     markers_ct_spatial <- tryCatch( -->
<!--       PhenoMapR::find_phenotype_markers( -->
<!--         seurat, -->
<!--         group_labels = group_df_ct, -->
<!--         group_column = "phenotype_group", -->
<!--         cell_id_column = "cell_id", -->
<!--         cell_type_column = "cell_type", -->
<!--         marker_scope = "cell_type_specific", -->
<!--         celltype_contrast = "within_cell_type", -->
<!--         assay = assay_ct, -->
<!--         slot = "data", -->
<!--         max_cells_per_ident = spatial_max_cells, -->
<!--         verbose = FALSE -->
<!--       ), -->
<!--       error = function(e) { -->
<!--         message("cell_type_specific find_phenotype_markers: ", conditionMessage(e)) -->
<!--         NULL -->
<!--       } -->
<!--     ) -->
<!--     if (!is.null(markers_ct_spatial)) { -->
<!--       message("Cell-type-specific adverse markers (top 5):") -->
<!--       print(utils::head(markers_ct_spatial$adverse_markers, 5)) -->
<!--       message("Cell-type-specific favorable markers (top 5):") -->
<!--       print(utils::head(markers_ct_spatial$favorable_markers, 5)) -->
<!--     } -->
<!--     expr_pm <- NULL -->
<!--     for (a in unique(c(assay_ct, assay_use, "RNA", "SCT"))) { -->
<!--       if (!a %in% names(seurat@assays)) next -->
<!--       expr_pm <- tryCatch( -->
<!--         as.matrix(Seurat::GetAssayData(seurat, layer = "data", assay = a)), -->
<!--         error = function(e) tryCatch( -->
<!--           as.matrix(Seurat::GetAssayData(seurat, slot = "data", assay = a)), -->
<!--           error = function(e2) NULL -->
<!--         ) -->
<!--       ) -->
<!--       if (!is.null(expr_pm) && nrow(expr_pm) > 0 && ncol(expr_pm) > 0) break -->
<!--       expr_pm <- tryCatch( -->
<!--         as.matrix(SeuratObject::LayerData(seurat, layer = "data", assay = a)), -->
<!--         error = function(e) NULL -->
<!--       ) -->
<!--       if (!is.null(expr_pm) && nrow(expr_pm) > 0 && ncol(expr_pm) > 0) break -->
<!--     } -->
<!--     if (!is.null(markers_ct_spatial) && !is.null(expr_pm) && ncol(expr_pm) > 0) { -->
<!--       meta_pm <- as.data.frame(seurat@meta.data[cells_ct, , drop = FALSE]) -->
<!--       meta_pm$cell_id_plot <- cells_ct -->
<!--       if (all(cells_ct %in% colnames(expr_pm))) { -->
<!--         expr_pm <- expr_pm[, cells_ct, drop = FALSE] -->
<!--       } -->
<!--       pal_ct <- PhenoMapR::get_celltype_palette(levels(factor(meta_pm[[ct_col_spatial]]))) -->
<!--       if (requireNamespace("ComplexHeatmap", quietly = TRUE) && requireNamespace("circlize", quietly = TRUE)) { -->
<!--         PhenoMapR::plot_phenotype_markers( -->
<!--           markers = markers_ct_spatial, -->
<!--           expr_mat = expr_pm, -->
<!--           meta = meta_pm, -->
<!--           cell_id_col = "cell_id_plot", -->
<!--           group_col = group_col, -->
<!--           score_col = score_col, -->
<!--           celltype_col = ct_col_spatial, -->
<!--           celltype_palette = pal_ct, -->
<!--           heatmap_type = "cell_type_specific", -->
<!--           top_n_markers = 20L, -->
<!--           n_mark_labels = 5L, -->
<!--           p_adj_threshold = 0.05, -->
<!--           column_title = paste0( -->
<!--             "Cell-type-specific phenotype markers (spatial; within_cell_type contrast; ", -->
<!--             "use celltype_contrast = \"vs_cohort_rest\" for cohort-wide DE)" -->
<!--           ) -->
<!--         ) -->
<!--       } else { -->
<!--         message("Install packages 'ComplexHeatmap' and 'circlize' to draw the cell-type-specific marker heatmap.") -->
<!--       } -->
<!--     } -->
<!--   } else { -->
<!--     message("No suitable cell type column in metadata; skipping cell_type_specific markers.") -->
<!--   } -->
<!-- } -->
<!-- ``` -->

<!-- ## Summary -->

<!-- - **Part 1 (spots)**: Spot-level object `HT270P1-S1H2Fc2U1Z1Bs1-H2Bs2-Test.hgnc.rds`; **H&E** via `SpatialPlot`; PhenoMapR scoring; **hex** summary of z-scaled scores plus **point** map of 5th-percentile groups (**same `coord_fixed` / limits as H&E**; no jitter). -->
<!-- - **Part 2 (CytoSPACE)**: `HT270P1-S1H2Fc2U1Z1Bs1-H2Bs2-Test_processed.rds`; same PRECOG **Pancreatic** reference at **cell** resolution; score distribution, score by cell type, prognostic groups, and spatial maps (jittered cells per spot). -->
<!-- - **Co-localization**: **[`spatialCooccur`](https://github.com/juninamo/spatialCooccur)** **`nhood_enrichment()`** and **`cooccur_local()`** on Visium coordinates and prognostic groups (Part 2 only; install from GitHub). -->
<!-- - **Markers**: (1) **Cell-type-agnostic** `find_phenotype_markers(..., marker_scope = "phenotype_groups")` with **`plot_phenotype_markers(..., heatmap_type = "global")`** (or **pheatmap** fallback); (2) **Cell-type-specific** `marker_scope = "cell_type_specific"` with **`plot_phenotype_markers(..., heatmap_type = "cell_type_specific")`** when **ComplexHeatmap** is available. -->

<!-- Optional download URLs: `PHENOMAPR_SPATIAL_SPOT_RDS_URL`, `PHENOMAPR_SPATIAL_CYTO_RDS_URL`; `PHENOMAPR_SPATIAL_RDS_URL` still applies to the **CytoSPACE** `.rds` if the cyto-specific variable is unset. -->

<!-- ## References -->

<!-- <a id="ref1"></a> -->
<!-- **[1]** Benard, B. A. et al. PRECOG update: an augmented resource of clinical outcome associations with gene expression for adult, pediatric, and immunotherapy cohorts. Nucleic Acids Res. 54, D1579–D1589 (2026). -->

<!-- ## Session Info -->

<!-- ```{r session-info} -->
<!-- sessionInfo() -->
<!-- ``` -->
