---
title: "Scoring spatial transcriptomics data with PhenoMapR"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_collapsed: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, out.width = "100%")
```

## Overview

This vignette demonstrates applying `{PhenoMapR}` to spatial transcriptomics data in two parts. **Part 1** uses **spot-level** 10X Visium data from an HTAN PDAC sample to show broad tissue regions of phenotypic interest. **Part 2** uses the same sample after we mapped paired single-cell data to the spots using [**CytoSPACE**](https://www.nature.com/articles/s41587-023-01697-9); we then mirror the single-cell workflow (score by cell type, identify marker genes, and plot heatmaps) at **cell** resolution and assess **co-localization** of prognostic cell types.

The sample (`HT270P1-S1H2Fc2U1Z1Bs1-H2Bs2-Test`) is from a pancreatic cancer dataset available from [HTAN](https://humantumoratlas.org/).

## Download data and vizualize H&E
```{r load-setup}
suppressPackageStartupMessages({
  library(PhenoMapR)
  library(Seurat)
  library(SeuratObject)
  library(ggplot2)
  library(dplyr)
  library(patchwork)
  library(ggchicklet2)
})

knitr::opts_chunk$set(fig.width = 12, out.width = "100%", warning = FALSE)
theme_set(theme_minimal(base_size = 14))

options(googledrive_quiet = TRUE)
googledrive::drive_deauth()

gd_id_spot <- "1OkIr7ksAWxKVjtdlGqYHMidvHZZsySEE"
rds_spot <- "HT270P1-S1H2Fc2U1Z1Bs1-H2Bs2-Test.hgnc.rds"

googledrive::drive_download(googledrive::as_id(gd_id_spot), rds_spot, overwrite = TRUE)

seurat_spot <- readRDS("HT270P1-S1H2Fc2U1Z1Bs1-H2Bs2-Test.hgnc.rds")

h_and_e <- SpatialPlot(object = seurat_spot,
  features = NULL,
  image.alpha = 1,
  pt.size.factor = 0) +
  guides(fill = "none") 

print(h_and_e)
```

## Score spots with PhenoMapR
```{r score-spots-part1}
# Score spots
scores_spot <- PhenoMapR::PhenoMap(
    expression = seurat_spot,
    reference = "precog",
    cancer_type = "Pancreatic",
    assay = "Spatial",
    slot = "data",
    verbose = FALSE
  )

# Add scores to Seurat object
seurat_spot <- PhenoMapR::add_scores_to_seurat(seurat_spot, scores_spot)

score_col_spot <- grep("weighted_sum_score", names(scores_spot), value = TRUE)[1]
## Z-scaled PhenoMapR score on spots (for hex / metadata plots)
sc_sp <- suppressWarnings(as.numeric(seurat_spot@meta.data[[score_col_spot]]))
seurat_spot$phenomapr_scaled_spot <- as.numeric(scale(sc_sp))

plot_score_distribution(
  seurat_spot$phenomapr_scaled_spot,
  main = "PRECOG Pancreatic score distribution (Visium spots)",
  base_size = 14
)
```

## PhenoMapR score distribution across spots 
```{r plot-spot-scores}
p <- SpatialPlot(
  object = seurat_spot,
  features = "weighted_sum_score_Pancreatic", image.alpha = 0
)
cols_pheno <- grDevices::colorRampPalette(c("#2166AC", "#F7F7F7", "#B2182B"))(100)

df <- p$data  # extract data from SpatialPlot

spot_hex_bins <- max(10, min(50L, as.integer(round(sqrt(nrow(df))))))

hex_phenomapr <- ggplot(df, aes(
  x = x,
  y = -y,
  z = scale(weighted_sum_score_Pancreatic)
)) +
  stat_summary_hex(fun = mean, bins = spot_hex_bins) +
  scale_fill_gradient2(
    low = "#2166AC",
    mid = "#F7F7F7",
    high = "#B2182B",
    midpoint = 0,
    trans = scales::pseudo_log_trans(sigma = 0.2),
    # limits = c(-4, 4),
    name = "PhenoMapR\nz-score"
  )+
  coord_fixed() +
  theme_void()


(h_and_e | hex_phenomapr) + 
  plot_layout(guides = "collect") & 
  theme(aspect.ratio = 1)
```
We can see that there are clear morphological associations with where adverse and favorable signals are being mapped to.


Next, let's define the most adverse and favorable groups by selecting the top and bottom 5th percentile of spots based on PhenoMapR scores.
```{r}
df <- SpatialPlot(
  object = seurat_spot,
  features = "weighted_sum_score_Pancreatic", image.alpha = 0
)$data %>%
  as.data.frame()

groups_spot <- PhenoMapR::define_phenotype_groups(df, score_columns = "weighted_sum_score_Pancreatic", percentile = 0.05)

df <- left_join(df, groups_spot, by = c("cell" = "cell_id"))

# Helper function for mode
mode_val <- function(z) {
  z <- z[is.finite(z)]
  if (!length(z)) return(NA_real_)
  as.numeric(names(sort(table(z), decreasing = TRUE)[1L]))
}

hex_phenomapr_groups <- ggplot(df, aes(
  x = x,
  y = -y,
  z = as.numeric(factor(
    phenotype_group_weighted_sum_score_Pancreatic,
    levels = c("Most Favorable", "Other", "Most Adverse")
  ))
)) +
  stat_summary_hex(
    aes(
      fill = after_stat(factor(
        value,
        levels = 1:3,
        labels = c("Most Favorable", "Other", "Most Adverse")
      ))
    ),
    bins = 50,
    fun = mode_val,
    colour = NA
  ) +
  scale_fill_manual(
    values = c(
      "Most Favorable" = "#2166AC",
      "Other"            = "#F7F7F7",
      "Most Adverse"     = "#B2182B"
    ),
    drop = FALSE,
    name = "PhenoMapR\nGroup"
  ) +
  coord_fixed() +
  theme_void()


(hex_phenomapr | hex_phenomapr_groups) + 
  plot_layout(guides = "collect") & 
  theme(aspect.ratio = 1)

```
Here, we can still see that the most favorable spots seem to cluster together, while a subset of the most adverse also tend to co-localize.


# Part 2: CytoSPACE-mapped cells

This sample from HTAN contains a paired single-cell sample from the same tissue block. In order to leverage the increased resolution of the single-cell data in a spatial context, we use CytoSPACE to map the single-cells to their spots.

## Load and score sample with PhenoMapR (CytoSPACE)
Here, we load the pre-processed Seurat object with the cells already maped.
```{r score-spots}
# Load the **CytoSPACE** object (single cells placed on Visium coordinates).
rds_cyto <- "HT270P1-S1H2Fc2U1Z1Bs1-H2Bs2-Test_processed.rds"
gd_id_cyto <- "1gcOyLriW9bIFNbDuQN6Vi1UsrMGKDxll"

googledrive::drive_download(googledrive::as_id(gd_id_cyto), rds_cyto, overwrite = TRUE)

seurat <- readRDS(rds_cyto)

  scores_spatial <- PhenoMapR::PhenoMap(
    expression = seurat,
    reference = "precog",
    cancer_type = "Pancreatic",
    assay = "Spatial",
    slot = "data",
    verbose = FALSE
  )
  seurat <- PhenoMapR::add_scores_to_seurat(seurat, scores_spatial)
score_col <- grep("weighted_sum_score", names(scores_spatial), value = TRUE)[1]

plot_score_distribution(
  seurat@meta.data[[score_col]],
  main = "PRECOG Pancreatic score distribution (CytoSPACE cells)",
  base_size = 14
)
```

## Score by cell type

Before looking at any spatial information, we plot the **PhenoMapR** score by cell type to see cell types most enriched in the adverse and favorable prognostic groups.

```{r score-by-annotation, fig.width=7.5, fig.height=5}
spatial_celltype_pal <- NULL
spatial_celltype_col <- NULL
meta_names <- names(seurat@meta.data)
celltype_col <- "CellType"

df <- seurat@meta.data
n_meta <- nrow(df)
ann_vec <- NULL
if (!is.null(celltype_col)) {
  raw <- df[[celltype_col]]
  if (is.list(raw)) {
    ann_vec <- vapply(raw, function(x) if (is.null(x) || length(x) == 0) NA_character_ else as.character(x)[1], character(1))
  } else {
    ann_vec <- as.vector(raw)
  }
  if (length(ann_vec) != n_meta) ann_vec <- NULL
}

  df$annotation <- factor(ann_vec, exclude = NULL)

  pal <- PhenoMapR::get_celltype_palette(levels(df$annotation))
print(ggplot(df, aes(
    x = reorder(.data$annotation, .data[[score_col]], FUN = median),
    y = .data[[score_col]],
    fill = .data$annotation
  )) +
        # geom_boxplot(outlier.alpha = 0.5, median.linewidth = 0.5,
        #          outlier.fill = NULL,
        #          outlier.color = NULL, outlier.shape = 21) +
      ggchicklet2::geom_chicklet_boxplot(radius = grid::unit(1, "pt"),
                                     outlier.alpha = 0.5, median.linewidth = 0.5,
                 outlier.fill = NULL,
                 outlier.color = NULL, outlier.shape = 21) +
    scale_fill_manual(values = pal, name = celltype_col) +
    coord_cartesian(ylim = c(-10000, 15000)) +
    guides(fill = guide_legend(ncol = 2)) +
    theme_minimal(base_size = 14) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
      legend.position = "right",
      plot.title = element_text(hjust = 0.5)
    ) +
    labs(y = "PRECOG Pancreatic score", x = celltype_col, title = "PhenoMapR Score by Cell Type"))
```
As seen in the other pancreatic cancer vignettes, the ductal cell type is the most associated with the adverse prognostic signal, while the beta and delta cells are most associated with favorable signal. Interestingly, plasma cells show quite a wide range of score and contain some of the most favorably prognostic cells across the sample. 



## Where are the different cell types?

Here, we use the single-cell coordinates from the spatial seurat object and pair them with the cell level metadata in order to plot the results in spatial context.

```{r}
cell_locations <- seurat@meta.data %>%
  as.data.frame() %>%
  dplyr::select("Cell", "row", "col", 
                "CellType", "weighted_sum_score_Pancreatic") %>%
    group_by(.data$row, .data$col) %>%
    mutate(points_per_location = n()) %>%
    ungroup()

cell_locations$CellType <- as.factor(cell_locations$CellType)

  # Shared jitter and point size so all three spatial plots match (multiple cells per spot)
  rng_row <- diff(range(cell_locations$row, na.rm = TRUE))
  rng_col <- diff(range(-cell_locations$col, na.rm = TRUE))
  spatial_jitter_w <- max(0.25, if (rng_row > 0) rng_row * 0.025 else 0.15)
  spatial_jitter_h <- max(0.25, if (rng_col > 0) rng_col * 0.025 else 0.15)
  spatial_point_range <- c(0.5, 1.6)

  spatial_celltype_pal <- PhenoMapR::get_celltype_palette(levels(cell_locations$CellType))
  
    ct_freq <- sort(table(cell_locations$CellType, useNA = "no"), decreasing = TRUE)
  ct_order <- names(ct_freq)
  cell_locations$celltype_zorder <- as.numeric(factor(as.character(cell_locations$CellType), levels = ct_order))
  ct_pal <- if (!is.null(spatial_celltype_pal)) spatial_celltype_pal else PhenoMapR::get_celltype_palette(levels(cell_locations$CellType))
 
 cytospace_loc <-  ggplot(cell_locations, aes(x = .data$row, y = -.data$col, color = .data$CellType,
    size = points_per_location, zorder = .data$celltype_zorder)) +
    geom_jitter(alpha = 0.8, width = spatial_jitter_w, height = spatial_jitter_h, shape = 16) +
    scale_color_manual(values = ct_pal, name = "Cell Type", na.value = "grey90") +
    scale_size_continuous(range = spatial_point_range, trans = "reverse", name = "Cells per spot") +
    guides(
      color = guide_legend(override.aes = list(size = 4), ncol = 2)
      # size  = guide_legend(override.aes = list(size = 4))
    ) +
    theme_minimal(base_size = 14) +
    theme(
      plot.title  = element_blank(),
      axis.text   = element_blank(),
      axis.ticks  = element_blank(),
      axis.title  = element_blank()
    ) +
       coord_fixed(ratio = 0.6) +
  theme_void()
  
 print(cytospace_loc)
```
Interestingly, it appears that several cell types tend to co-localize with themselves, some of which seem to be localized to the areas where we saw the greatest adverse and favorable scores mapping to at the spot level.


### Where raw PhenoMapR scores are

Spatial map of PhenoMapR score (z-scaled for color gradient). Blue = more favorable, red = more adverse.

```{r, fig.width=12}
cell_locations <- cell_locations[order(abs(cell_locations$weighted_sum_score_Pancreatic)), ]

sc_phenomapr <- ggplot(cell_locations, aes(
  x = .data$row,
  y = -.data$col,
  color = scale(weighted_sum_score_Pancreatic),
  size = .data$points_per_location
)) +
  geom_jitter(alpha = 0.8, width = spatial_jitter_w, height = spatial_jitter_h, shape = 16) +
  scale_size_continuous(range = spatial_point_range, trans = "reverse", name = "Cells per spot") +
  scale_color_gradient2(
    low = "#2166AC", mid = "#F7F7F7", high = "#B2182B",
    midpoint = 0,
    trans = scales::pseudo_log_trans(sigma = 0.2),
    name = "PhenoMapR\nz-score"
  ) +
  coord_fixed(ratio = 0.6) +
  theme_void()

(cytospace_loc | sc_phenomapr) + 
  # plot_layout(guides = "collect") & 
  theme(aspect.ratio = 1)
```



### Where 5th percentile cells are

In order to make the visualization more apparent, we restrict the spatial map of prognostic groups to: top 5% (Most Adverse), bottom 5% (Most Favorable), and the rest (Other).

```{r}
cell_locations <- cell_locations %>% 
  mutate(percentile = percent_rank(weighted_sum_score_Pancreatic)) %>%
  mutate(prognostic_group = case_when(
    percentile < 0.05 ~ "Favorable",
    percentile >= 0.95 ~ "Adverse",
    TRUE ~ "Other"
  ))

  df_other <- cell_locations %>%
    dplyr::filter(prognostic_group=="Other")
  
  df_extreme <- cell_locations %>%
    dplyr::filter(prognostic_group!="Other")
  
 sc_phenomapr_5 <- ggplot() +
    geom_jitter(data = df_other, aes(x = .data$row, y = -.data$col, color = .data$prognostic_group,
      size = points_per_location), alpha = 0.8, width = spatial_jitter_w, height = spatial_jitter_h, shape = 16) +
    geom_jitter(data = df_extreme, aes(x = .data$row, y = -.data$col, color = .data$prognostic_group,
      size = points_per_location), alpha = 0.8, width = spatial_jitter_w, height = spatial_jitter_h, shape = 16) +
    # ggtitle("5th percentile: Most Adverse vs Most Favorable") +
    scale_color_manual(
      values = c(`Adverse` = "#B2182B", Other = "#f7f7f7", `Favorable` = "#2166AC"),
      name = "Prognostic group",
      na.value = "grey90",
      drop = FALSE
    ) +
     guides(
      color = guide_legend(override.aes = list(size = 4))
      # size  = guide_legend(override.aes = list(size = 4))
    ) +
    scale_size_continuous(range = spatial_point_range, trans = "reverse", name = "Cells per spot") +
    theme_minimal(base_size = 14) +
    theme(
      plot.title = element_text(hjust = 0.5, size = 14),
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      axis.title = element_blank()
    ) +
  coord_fixed(ratio = 0.6) +
  theme_void()

  (sc_phenomapr | sc_phenomapr_5) + 
  # plot_layout(guides = "collect") & 
  theme(aspect.ratio = 1)
  
```

Visually, adverse and favorable cells can appear to segregate on the tissue. The next section quantifies **neighborhood co-occurrence** of prognostic groups with **[`spatialCooccur`](https://github.com/juninamo/spatialCooccur)** (Inamo *et al.*, [medRxiv 2025](https://doi.org/10.1101/2025.08.05.25332835)).

## Co-localization of prognostic cells (CytoSPACE)

We use the same **Visium row / column** coordinates as the spatial plots, assembled as a **`data.frame`** with **`x`**, **`y`**, and a **combined label** that includes **cell type and prognostic group**. Each cell is labeled as `CellType_prognostic_group` (e.g. `Ductal_Adverse`). This lets us ask whether, for example, **adverse ductal cells** co-localize differently than **favorable ductal cells**.

Install **`spatialCooccur`** from GitHub if needed: `remotes::install_github("juninamo/spatialCooccur")`.

**`spatialCooccur::nhood_enrichment()`** builds a **kNN graph** (**`Seurat::FindNeighbors()`** inside the package), optionally normalizes adjacency, aggregates **co-occurrence** between cluster pairs, and compares the observed matrix to **permutation** nulls to produce a **z-score** matrix. We use **`n_jobs = 1`** so the vignette runs on a single core (set higher locally if you install the package). **`spatialCooccur::cooccur_local()`** scores each cell by whether its **radius** neighborhood contains both **Adverse** and **Favorable** labels, then applies a short diffusion-style step when **`maxnsteps > 0`**. Duplicate spot coordinates (many CytoSPACE cells per spot) get a small jitter.

The neighborhood **Z-score** matrix is shown with **ComplexHeatmap** when available: **row and column annotations** split **`CellType`** and **`Adverse` / `Favorable` / `Other`** (combined labels remain `CellType_prognostic_group`). If **ComplexHeatmap** is not installed, a **ggplot** tile heatmap is used instead.

```{r cytospace-spatialcooccur, fig.width=7, fig.height=6, out.width=NULL}
scoc_nhood <- NULL
scoc_local_by_ct <- NULL

if (!requireNamespace("spatialCooccur", quietly = TRUE)) {
  message("Install spatialCooccur for co-localization: remotes::install_github(\"juninamo/spatialCooccur\")")
} else if (!exists("cell_locations") ||
           !all(c("Cell", "row", "col", "CellType", "prognostic_group") %in% names(cell_locations))) {
  message("Co-localization skipped: expected `cell_locations` with columns Cell, row, col, CellType, prognostic_group.")
} else {
  dfb <- cell_locations[
    stats::complete.cases(cell_locations[, c("Cell", "row", "col", "CellType", "prognostic_group")]),
    ,
    drop = FALSE
  ]
  dfb$Cell <- as.character(dfb$Cell)
  dfb$row <- as.numeric(dfb$row)
  dfb$col <- as.numeric(dfb$col)
  dfb$CellType <- as.character(dfb$CellType)
  dfb$prognostic_group <- as.character(dfb$prognostic_group)
  dfb$prognostic_group <- dplyr::recode(
    dfb$prognostic_group,
    `Most Adverse` = "Adverse",
    `Most Favorable` = "Favorable"
  )
  dfb <- dfb[dfb$prognostic_group %in% c("Adverse", "Favorable", "Other"), , drop = FALSE]
  dfb <- dfb[!is.na(dfb$CellType) & nzchar(trimws(dfb$CellType)), , drop = FALSE]
  dfb$ct_pg <- paste0(dfb$CellType, "_", dfb$prognostic_group)

  scoc_df <- data.frame(
    x = dfb$row,
    y = dfb$col,
    ct_pg = dfb$ct_pg,
    stringsAsFactors = FALSE
  )
  rownames(scoc_df) <- dfb$Cell
  # spatialCooccur::cooccur_local expects the label column to be named `cell_type`
  scoc_df$cell_type <- scoc_df$ct_pg

  # Keep runtime bounded while preserving enough cells for stable permutation nulls.
  # max_cells_sc <- 4000L
  # if (nrow(scoc_df) > max_cells_sc) {
  #   set.seed(3L)
  #   scoc_df <- scoc_df[sample.int(nrow(scoc_df), max_cells_sc), , drop = FALSE]
  # }

  # Many CytoSPACE cells share identical spot coordinates; jitter to avoid duplicates in kNN/radius logic.
  xy_mat <- as.matrix(scoc_df[, c("x", "y"), drop = FALSE])
  if (any(duplicated(xy_mat) | duplicated(xy_mat, fromLast = TRUE))) {
    rng <- max(
      diff(range(xy_mat[, 1], na.rm = TRUE)),
      diff(range(xy_mat[, 2], na.rm = TRUE)),
      na.rm = TRUE
    )
    eps <- if (is.finite(rng) && rng > 0) rng * 1e-5 else 1e-6
    set.seed(4L)
    scoc_df$x <- scoc_df$x + stats::runif(nrow(scoc_df), 0, eps)
    scoc_df$y <- scoc_df$y + stats::runif(nrow(scoc_df), 0, eps)
  }

  k_sc <- min(20L, max(5L, nrow(scoc_df) - 1L))
  n_perm_sc <- 100L

  n_grp <- length(unique(scoc_df$ct_pg))
  if (nrow(scoc_df) >= (k_sc + 3L) && n_grp >= 2L) {
    scoc_nhood <- tryCatch(
      spatialCooccur::nhood_enrichment(
        scoc_df,
        cluster_key = "ct_pg",
        neighbors.k = k_sc,
        connectivity_key = "nn",
        transformation = TRUE,
        n_perms = n_perm_sc,
        seed = 42L,
        n_jobs = 1L
      ),
      error = function(e) {
        message("spatialCooccur::nhood_enrichment failed: ", conditionMessage(e))
        NULL
      }
    )
  }

  rad_sc <- max(3, 0.04 * max(
    diff(range(scoc_df$x, na.rm = TRUE)),
    diff(range(scoc_df$y, na.rm = TRUE)),
    na.rm = TRUE
  ))

  # Local co-occurrence within each cell type:
  # compare CellType_Adverse neighborhoods for presence of CellType_Favorable (and vice versa).
  df_local_list <- list()
  cell_types <- sort(unique(sub("_(Adverse|Favorable|Other)$", "", unique(scoc_df$ct_pg))))
  for (ct in cell_types) {
    xlab <- paste0(ct, "_Adverse")
    ylab <- paste0(ct, "_Favorable")
    if (!(xlab %in% scoc_df$ct_pg && ylab %in% scoc_df$ct_pg)) next
    if (sum(scoc_df$ct_pg == xlab) < 3L || sum(scoc_df$ct_pg == ylab) < 3L) next

    loc_one <- tryCatch(
      spatialCooccur::cooccur_local(
        scoc_df,
        cluster_x = xlab,
        cluster_y = ylab,
        connectivity_key = "nn",
        neighbors.k = k_sc,
        radius = rad_sc,
        maxnsteps = 3L
      ),
      error = function(e) {
        message("spatialCooccur::cooccur_local failed for ", ct, ": ", conditionMessage(e))
        NULL
      }
    )
    if (is.null(loc_one) || ncol(loc_one) < 1L) next

    cn <- colnames(loc_one)[1]
    v <- loc_one[[1]]
    ids <- rownames(loc_one)
    if (is.null(ids) || length(ids) != length(v)) ids <- names(v)
    if (is.null(ids) || length(ids) != length(v)) ids <- as.character(seq_len(length(v)))

    df_local_list[[ct]] <- data.frame(
      cell = as.character(ids),
      cell_type = ct,
      score = as.numeric(v),
      metric = cn,
      stringsAsFactors = FALSE
    )
  }
  if (length(df_local_list) > 0) {
    scoc_local_by_ct <- do.call(rbind, df_local_list)
  }
}

if (!is.null(scoc_nhood) && !is.null(scoc_nhood$zscore)) {
  # print(scoc_nhood$count)

  zm <- as.matrix(scoc_nhood$zscore)
  lab_clean <- function(nm) gsub("^Cluster", "", nm)
  rn <- lab_clean(rownames(zm))
  cn <- lab_clean(colnames(zm))

  suffix_pg <- c("Adverse", "Favorable", "Other")
  .parse_ct_pg <- function(lab) {
    for (s in suffix_pg) {
      suff <- paste0("_", s)
      if (nzchar(lab) && endsWith(lab, suff)) {
        ct <- substr(lab, 1L, nchar(lab) - nchar(suff))
        return(c(ct = ct, pg = s))
      }
    }
    c(ct = lab, pg = NA_character_)
  }

  all_labs <- sort(unique(c(rn, cn)))
  meta_labs <- as.data.frame(
    do.call(rbind, lapply(all_labs, function(l) {
      p <- .parse_ct_pg(l)
      data.frame(label = l, CellType = p[["ct"]], Prognostic = p[["pg"]], stringsAsFactors = FALSE)
    })),
    stringsAsFactors = FALSE
  )
  meta_labs$Prognostic <- factor(meta_labs$Prognostic, levels = suffix_pg)
  meta_labs <- meta_labs[!is.na(meta_labs$Prognostic), , drop = FALSE]
  ord <- order(meta_labs$CellType, meta_labs$Prognostic)
  labs_ord <- meta_labs$label[ord]
  labs_ord <- labs_ord[labs_ord %in% rn & labs_ord %in% cn]
  if (length(labs_ord) < 2L) {
    message("Not enough overlapping labels in zscore matrix for heatmap.")
  } else {
    mat <- zm[match(labs_ord, rn), match(labs_ord, cn), drop = FALSE]
    dimnames(mat) <- list(labs_ord, labs_ord)

    row_ct <- as.character(meta_labs$CellType[match(rownames(mat), meta_labs$label)])
    row_pg <- as.character(meta_labs$Prognostic[match(rownames(mat), meta_labs$label)])
    col_ct <- as.character(meta_labs$CellType[match(colnames(mat), meta_labs$label)])
    col_pg <- as.character(meta_labs$Prognostic[match(colnames(mat), meta_labs$label)])

    pal_pg <- c(Adverse = "#B2182B", Favorable = "#2166AC", Other = "#f7f7f7")
    uct <- sort(unique(c(row_ct, col_ct)))
    pal_ct <- PhenoMapR::get_celltype_palette(uct)

    ## Cell counts per combined label (cells entering nhood_enrichment, same order as heatmap columns).
    col_ncells <- rep(0L, ncol(mat))
    names(col_ncells) <- colnames(mat)
    if (exists("scoc_df") && is.data.frame(scoc_df) && "ct_pg" %in% names(scoc_df)) {
      tab_ct <- table(scoc_df$ct_pg)
      hit <- names(col_ncells) %in% names(tab_ct)
      col_ncells[hit] <- as.integer(tab_ct[names(col_ncells)[hit]])
    }

    ## Two-sided p-values from |Z| (normal approximation; spatialCooccur returns zscore + count only).
    mat_p <- 2 * stats::pnorm(-abs(mat))
    mat_p[!is.finite(mat)] <- NA_real_
    ## Benjamini–Hochberg FDR across all matrix cells (multiple pairwise label comparisons).
    mat_p_adj <- mat_p
    pv_flat <- as.vector(mat_p)
    ok_flat <- is.finite(pv_flat)
    mat_p_adj[] <- NA_real_
    mat_p_adj[ok_flat] <- stats::p.adjust(pv_flat[ok_flat], method = "BH")

    drew_ch <- FALSE
    if (requireNamespace("ComplexHeatmap", quietly = TRUE) && requireNamespace("circlize", quietly = TRUE)) {
      mabs <- suppressWarnings(max(abs(as.numeric(mat)), na.rm = TRUE))
      if (!is.finite(mabs) || mabs <= 0) mabs <- 1
      
      # "#7F312FFF", "#BE4A47FF", "#FD817EFF", "#FEA19EFF", "#FEC0BFFF", "#F5F5F5FF", "#99E3DDFF", "#66D4CCFF", "#33C6BBFF", "#008A80FF", "#005C55FF"
      
      col_fun <- circlize::colorRamp2(c(-mabs, 0, mabs), c("#7F312FFF", "#f7f7f7", "#005C55FF"))

      row_ha <- ComplexHeatmap::rowAnnotation(
        CellType = row_ct,
        `Prognostic Group` = row_pg,
        col = list(CellType = pal_ct, `Prognostic Group` = pal_pg),
        # annotation_name_side = NULL,
        show_annotation_name = FALSE,
        simple_anno_size = grid::unit(4, "mm"),
        show_legend = c(FALSE)
      )
      col_ha <- ComplexHeatmap::HeatmapAnnotation(
                `# cells` = ComplexHeatmap::anno_barplot(
          col_ncells,
          gp = grid::gpar(fill = "#666666"),
          border = FALSE,
          height = grid::unit(14, "mm"),
          ylim = c(0, max(col_ncells, 1L, na.rm = TRUE))
        ),
        CellType = col_ct,
        `Prognostic Group` = col_pg,
        col = list(CellType = pal_ct, `Prognostic Group` = pal_pg),
        annotation_name_side = "right",
        simple_anno_size = grid::unit(4, "mm")
      )

ht <- ComplexHeatmap::Heatmap(
  mat,
  name = "Z",
  col = col_fun,
  cluster_rows    = FALSE,
  cluster_columns = FALSE,
  show_row_names    = FALSE,
  show_column_names = FALSE,
  show_row_dend    = FALSE,
  show_column_dend = FALSE,
  row_names_gp    = grid::gpar(fontsize = 7),
  column_names_gp = grid::gpar(fontsize = 7),
  left_annotation = row_ha,
  top_annotation  = col_ha,
  row_title    = "Reference Cell Type",    
  column_title = "Neighborhood Cell Type",  
  row_title_gp    = grid::gpar(fontsize = 12),
  column_title_gp = grid::gpar(fontsize = 12),
    row_title_side    = "left",       
  column_title_side = "bottom", 
  heatmap_legend_param = list(title = "Z-score"),
  border = TRUE,
  cell_fun = function(j, i, x, y, w, h, fill) {
    qv <- mat_p_adj[i, j]
    if (!is.finite(qv)) return(invisible(NULL))
    sym <- if (qv < 0.001) "***" else if (qv < 0.01) "**" else if (qv < 0.05) "*" else return(invisible(NULL))
    grid::grid.text(sym, x, y, gp = grid::gpar(fontsize = 9))
  }
)

# Draw with main title separately
ComplexHeatmap::draw(
  ht,
  column_title     = "Neighborhood Enrichment of Prognostic Cell Types",
  column_title_gp  = grid::gpar(fontsize = 14, fontface = "bold"),
  heatmap_legend_side = "right", annotation_legend_side = "right"
)
      drew_ch <- TRUE
    }
  }
}

if (!is.null(scoc_local_by_ct) && nrow(scoc_local_by_ct) > 0L) {
  p_loc <- ggplot(scoc_local_by_ct, aes(x = stats::reorder(.data$cell_type, .data$score, median), y = .data$score)) +
    geom_boxplot(outlier.alpha = 0.3) +
    coord_flip() +
    labs(
      title = "Local co-occurrence score (spatialCooccur::cooccur_local)",
      subtitle = paste0("Within-cell-type neighborhoods (radius ≈ ", signif(rad_sc, 3), "): adverse vs favorable"),
      x = "Cell type",
      y = unique(scoc_local_by_ct$metric)[1]
    ) +
    theme_minimal(base_size = 11)
  # print(p_loc)
}
```
Stars use Benjamini–Hochberg FDR on two-sided p(|Z|): *** q<0.001, ** q<0.01, * q<0.05

## Now, we cluster to find broader patterns
```{r, fig.width=7, fig.height=6, out.width=NULL}
ht <- ComplexHeatmap::Heatmap(
  mat,
  name = "Z",
  col = col_fun,
  cluster_rows    = T,
  cluster_columns = T,
  show_row_names    = FALSE,
  show_column_names = FALSE,
  show_row_dend    = FALSE,
  show_column_dend = FALSE,
  row_names_gp    = grid::gpar(fontsize = 7),
  column_names_gp = grid::gpar(fontsize = 7),
  left_annotation = row_ha,
  top_annotation  = col_ha,
  row_title    = "Reference Cell Type",    
  column_title = "Neighborhood Cell Type",  
  row_title_gp    = grid::gpar(fontsize = 12),
  column_title_gp = grid::gpar(fontsize = 12),
    row_title_side    = "left",       
  column_title_side = "bottom", 
  heatmap_legend_param = list(title = "Z-score"),
  border = TRUE,
  cell_fun = function(j, i, x, y, w, h, fill) {
    qv <- mat_p_adj[i, j]
    if (!is.finite(qv)) return(invisible(NULL))
    sym <- if (qv < 0.001) "***" else if (qv < 0.01) "**" else if (qv < 0.05) "*" else return(invisible(NULL))
    grid::grid.text(sym, x, y, gp = grid::gpar(fontsize = 9))
  }
)

# Draw with main title separately
ComplexHeatmap::draw(
  ht,
  column_title     = "Neighborhood Enrichment of Prognostic Cell Types",
  column_title_gp  = grid::gpar(fontsize = 14, fontface = "bold"),
  heatmap_legend_side = "right", annotation_legend_side = "right"
)
```


## Prognostic markers

`find_phenotype_markers()` supports the same **marker scopes** on spatial data (cells or spots mapped to spots) as in the single-cell vignette:

- **Cell type agnostic** — `marker_scope = "phenotype_groups"` (**default**): **Most Adverse** vs all other cells with a phenotype label, and **Most Favorable** vs all other cells, **ignoring** cell type. This asks which genes distinguish the prognostic extremes in the whole mixture on the slide.
- **Cell type specific** — `marker_scope = "cell_type_specific"`: for each cell type in metadata (here, CytoSPACE-mapped types), contrasts that type in the adverse or favorable tail vs a reference set controlled by `celltype_contrast`. **Default** `celltype_contrast = "within_cell_type"` compares to other phenotype bins **within the same type** (reduces housekeeping genes repeating across type × tail blocks). **`celltype_contrast = "vs_cohort_rest"`** restores the **original** cohort-wide contrast: (type ∩ tail) vs **every other cell**. With **`celltype_unique_genes = TRUE`** (default), each gene is kept in at most one adverse/favorable row (strongest `avg_log2FC`).

Below, **Step 1–2** use **cell-type-agnostic** markers; **Step 2** draws **`plot_phenotype_markers(..., heatmap_type = "global")`** when **ComplexHeatmap** and **circlize** are installed (same helper as the **single-cell** vignette), with a **pheatmap** fallback otherwise. **Step 3** runs **cell-type-specific** markers and **`plot_phenotype_markers(..., heatmap_type = "cell_type_specific")`** when those packages are available.

### Step 1: Markers for adverse vs. favorable cells (cell type agnostic)

```{r markers-spatial, cache = FALSE}
markers <- NULL
assay_markers <- NULL
## Tail labels on the Seurat object (5th / 95th percentiles via define_phenotype_groups).
cells <- colnames(seurat)
scores_df_markers <- seurat@meta.data[cells, score_col, drop = FALSE]
rownames(scores_df_markers) <- cells
groups_markers <- PhenoMapR::define_phenotype_groups(
  scores_df_markers,
  percentile = 0.05,
  score_columns = score_col
)
group_col <- grep("phenotype_group", names(groups_markers), value = TRUE)[1]

seurat@meta.data[[group_col]] <-
  groups_markers[rownames(seurat@meta.data), group_col, drop = TRUE]

group_vec <- seurat@meta.data[cells, group_col]
group_df <- data.frame(
  cell_id = cells,
  phenotype_group = as.character(group_vec),
  stringsAsFactors = FALSE
)
markers <- PhenoMapR::find_phenotype_markers(
  seurat,
  group_labels = group_df,
  group_column = "phenotype_group",
  cell_id_column = "cell_id",
  marker_scope = "phenotype_groups",
  assay = "Spatial",
  slot = "data",
  # max_cells_per_ident = spatial_max_cells,
  verbose = FALSE
)
```

### Step 2: Heatmap of adverse vs. favorable markers (global phenotype groups)

**`plot_phenotype_markers(..., heatmap_type = "global")`** matches the single-cell vignette: favorable- then adverse-marker rows, column order by PhenoMapR score, and **pheatmap** only if ComplexHeatmap is unavailable or marker tables are incomplete.

```{r heatmap-markers-spatial, fig.align='center', fig.width=10, fig.height=4}
  n_top <- 15L
  expr <- NULL
  assay_use_hm <- if (exists("assay_use") && !is.null(assay_use)) as.character(assay_use)[1] else "Spatial"
  assay_order <- unique(c(
    if (exists("assay_markers") && !is.null(assay_markers)) assay_markers else character(0),
    assay_use_hm, "RNA", "SCT"
  ))
  for (a in assay_order) {
    if (!a %in% names(seurat@assays)) next
    expr <- tryCatch(
      Seurat::GetAssayData(seurat, layer = "data", assay = a),
      error = function(e) tryCatch(Seurat::GetAssayData(seurat, slot = "data", assay = a), error = function(e2) NULL)
    )
    if (!is.null(expr) && nrow(expr) > 0 && ncol(expr) > 0) break
    expr <- tryCatch(
      SeuratObject::LayerData(seurat, layer = "data", assay = a),
      error = function(e) NULL
    )
    if (is.null(expr) || ncol(expr) == 0) {
      expr <- tryCatch(
        SeuratObject::LayerData(seurat, layer = "counts", assay = a),
        error = function(e) tryCatch(SeuratObject::LayerData(seurat, layer = NULL, assay = a), error = function(e2) NULL)
      )
    }
    if (!is.null(expr) && nrow(expr) > 0 && ncol(expr) > 0) break
  }
  if (!is.null(expr) && ncol(expr) > 0) {
    cells_expr <- colnames(expr)
    if (is.null(cells_expr)) cells_expr <- character(0)
    cells_obj <- colnames(seurat)
    cells_meta <- rownames(seurat@meta.data)
    cells_use <- intersect(cells_expr, cells_obj)
    if (length(cells_use) == 0 && length(cells_meta) > 0) {
      cells_use <- intersect(cells_expr, cells_meta)
    }
    if (length(cells_use) == 0 && length(cells_expr) > 0 && any(grepl("_", cells_expr, fixed = TRUE))) {
      stripped <- sub("^[^_]+_", "", cells_expr)
      cells_use <- intersect(stripped, cells_obj)
      if (length(cells_use) == 0) cells_use <- intersect(stripped, cells_meta)
      if (length(cells_use) > 0) {
        expr_cols <- vapply(cells_use, function(c) cells_expr[which(stripped == c)[1]], character(1))
        expr <- expr[, expr_cols, drop = FALSE]
        colnames(expr) <- cells_use
      }
    }
    if (length(cells_use) == 0) {
      message("No overlapping cells between expression and metadata; skipping heatmap. ",
              "expr ncol=", length(cells_expr), ", obj ncol=", length(cells_obj),
              if (length(cells_expr) > 0) paste0("; expr sample: ", head(cells_expr, 2)) else ""
      )
    } else {
      expr_pm <- as.matrix(expr[, cells_use, drop = FALSE])
      meta_pm <- seurat@meta.data[cells_use, , drop = FALSE]
      meta_pm$cell_id_plot <- cells_use
      group_col_hm <- paste0("phenotype_group_", score_col)
      if (!group_col_hm %in% names(meta_pm)) group_col_hm <- group_col

      ct_col_hm <- NULL
      if (exists("spatial_df_celltype_col") && !is.null(spatial_df_celltype_col) &&
          spatial_df_celltype_col %in% names(meta_pm)) {
        ct_col_hm <- spatial_df_celltype_col
      } else if (exists("celltype_col") && !is.null(celltype_col) && celltype_col %in% names(meta_pm)) {
        ct_col_hm <- celltype_col
      }
      if (is.null(ct_col_hm)) {
        meta_pm$celltype_plot <- factor(rep("Cell", nrow(meta_pm)))
        ct_col_hm <- "celltype_plot"
      } else {
        meta_pm[[ct_col_hm]] <- factor(as.character(meta_pm[[ct_col_hm]]))
      }
      pal_ct <- PhenoMapR::get_celltype_palette(levels(meta_pm[[ct_col_hm]]))

      drew_ch <- FALSE
      if (requireNamespace("ComplexHeatmap", quietly = TRUE) && requireNamespace("circlize", quietly = TRUE)) {
        adv_ok <- !is.null(markers$adverse_markers) && nrow(markers$adverse_markers) > 0L
        fav_ok <- !is.null(markers$favorable_markers) && nrow(markers$favorable_markers) > 0L
        if (adv_ok && fav_ok) {
          PhenoMapR::plot_phenotype_markers(
            markers = markers,
            expr_mat = expr_pm,
            meta = meta_pm,
            cell_id_col = "cell_id_plot",
            group_col = group_col_hm,
            score_col = score_col,
            celltype_col = ct_col_hm,
            celltype_palette = pal_ct,
            heatmap_type = "global",
            top_n_markers = 100L,
            n_mark_labels = 15L,
            p_adj_threshold = 0.05,
            column_title = ""
          )
        }
      }
    }
  }
```

### Stacked barplot: adverse vs. favorable by cell type

Each column is a cell type; the height is filled by the number of adverse (5th percentile) or favorable (5th percentile) cells.

```{r stacked-bar-adverse-favorable-by-celltype, fig.width=5, fig.height=5}
if (!is.null(group_col)) {
  ct_col_bar <- NULL
  if (exists("spatial_df_celltype_col") && !is.null(spatial_df_celltype_col) && spatial_df_celltype_col %in% names(seurat@meta.data)) {
    ct_col_bar <- spatial_df_celltype_col
  } else if (exists("celltype_col") && !is.null(celltype_col) && celltype_col %in% names(seurat@meta.data)) {
    ct_col_bar <- celltype_col
  } else {
    for (c in c("CellType", "Celltype..major.lineage.", "cell_type", "celltype", "annotation", "seurat_clusters")) {
      if (c %in% names(seurat@meta.data) && length(unique(na.omit(seurat@meta.data[[c]]))) >= 2) {
        ct_col_bar <- c
        break
      }
    }
  }
  if (!is.null(ct_col_bar)) {
    meta <- seurat@meta.data
    meta$pg <- meta[[group_col]]
    meta$ct <- as.character(meta[[ct_col_bar]])
    meta$ct_ok <- !is.na(meta$ct) & nzchar(trimws(meta$ct))
    idx_extreme <- meta$pg %in% c("Most Adverse", "Most Favorable") & meta$ct_ok
    df_bar <- as.data.frame(table(
      CellType = meta$ct[idx_extreme],
      Prognostic_group = meta$pg[idx_extreme],
      useNA = "no"
    ))
    if (nrow(df_bar) > 0) {
      df_bar$Prognostic_group <- factor(df_bar$Prognostic_group, levels = c("Most Favorable", "Most Adverse"))
      pal_bar <- c(`Most Adverse` = "#B2182B", `Most Favorable` = "#2166AC")
      adverse <- df_bar[df_bar$Prognostic_group == "Most Adverse", c("CellType", "Freq")]
      fav <- df_bar[df_bar$Prognostic_group == "Most Favorable", c("CellType", "Freq")]
      names(adverse)[2] <- "n_adverse"
      names(fav)[2] <- "n_fav"
      df_labels <- merge(adverse, fav, by = "CellType", all = TRUE)
      df_labels$n_adverse[is.na(df_labels$n_adverse)] <- 0
      df_labels$n_fav[is.na(df_labels$n_fav)] <- 0
      df_labels$label <- paste0(df_labels$n_adverse, "/", df_labels$n_fav)
      df_labels$total <- df_labels$n_adverse + df_labels$n_fav
      ct_ord <- levels(reorder(df_bar$CellType, df_bar$Freq, function(x) -sum(x)))
      df_labels$CellType <- factor(df_labels$CellType, levels = ct_ord)
      p_bar <- ggplot(df_bar, aes(x = reorder(.data$CellType, .data$Freq, function(x) -sum(x)), y = .data$Freq, fill = .data$Prognostic_group)) +
        geom_col(position = "stack") +
        # ggchicklet2::geom_chicklet_bar(radius = grid::unit(1.5, "pt"), position = "stack") +
        geom_text(data = df_labels, aes(x = .data$CellType, y = .data$total, label = .data$n_adverse),
                  inherit.aes = FALSE, vjust = -0.3, size = 3.5, color = "#B2182B", hjust = 1.05) +
        geom_text(data = df_labels, aes(x = .data$CellType, y = .data$total, label = paste0("/", .data$n_fav)),
                  inherit.aes = FALSE, vjust = -0.3, size = 3.5, color = "#2166AC", hjust = -0.05) +
        scale_fill_manual(values = pal_bar) +
        labs(x = "Cell type", y = "Number of cells", fill = "Prognostic group",
             title = "Adverse vs. favorable cells (5th percentile) by cell type") +
        theme_minimal(base_size = 14) +
        theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12), legend.position = "top")
      print(p_bar)
    }
  }
}
```
In line with the other Pancreatic cancer vignettes, PhenoMapR identifies ductal cells (and some fibroblasts) as the most associated cell type with an adverse PhenoMapR prognostic score. Alpha, plasma, and beta cells are the most associated with the more favorable prognostic signal. Interestingly, fibroblasts seems to comprise both adverse and favorable phenotypes.


## Summary

- **Part 1 (spots)**: Spot-level object `HT270P1-S1H2Fc2U1Z1Bs1-H2Bs2-Test.hgnc.rds`; **H&E** via `SpatialPlot`; PhenoMapR scoring; **hex** summary of z-scaled scores plus **point** map of 5th-percentile groups (**same `coord_fixed` / limits as H&E**; no jitter).
- **Part 2 (CytoSPACE)**: `HT270P1-S1H2Fc2U1Z1Bs1-H2Bs2-Test_processed.rds`; same PRECOG **Pancreatic** reference at **cell** resolution; score distribution, score by cell type, prognostic groups, and spatial maps (jittered cells per spot).
- **Co-localization**: **[`spatialCooccur`](https://github.com/juninamo/spatialCooccur)** **`nhood_enrichment()`** and **`cooccur_local()`** on Visium coordinates and prognostic groups (Part 2 only; install from GitHub).
- **Markers**: (1) **Cell-type-agnostic** `find_phenotype_markers(..., marker_scope = "phenotype_groups")` with **`plot_phenotype_markers(..., heatmap_type = "global")`** (or **pheatmap** fallback); (2) **Cell-type-specific** `marker_scope = "cell_type_specific"` with **`plot_phenotype_markers(..., heatmap_type = "cell_type_specific")`** when **ComplexHeatmap** is available.

## References

<a id="ref1"></a>
**[1]** Benard, B. A. et al. PRECOG update: an augmented resource of clinical outcome associations with gene expression for adult, pediatric, and immunotherapy cohorts. Nucleic Acids Res. 54, D1579–D1589 (2026).

## Session Info

```{r session-info}
sessionInfo()
```
