---
title: "Scoring spatial transcriptomics data with PhenoMapR"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_collapsed: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, out.width = "100%")

# GitHub Actions / pkgdown often hits exit 143 (SIGTERM) from OOM or timeout on this vignette.
spatial_ci_lite <- identical(Sys.getenv("CI"), "true") ||
  tolower(Sys.getenv("PHENOMAPR_SPATIAL_LITE", "")) %in% c("true", "1", "yes")
spatial_max_cells <- if (spatial_ci_lite) 2000L else 5000L
```

## Overview

This vignette demonstrates using **{PhenoMapR}** on spatial transcriptomics data. We use a preprocessed 10X Visium spatial object (`HT270P1-S1H2Fc2U1Z1Bs1-H2Bs2-Test_processed.rds`) to score spots with the PRECOG **Pancreatic** reference, define adverse vs. favorable prognostic groups, and visualize both score distributions and **where** PhenoMapR scores localize on the tissue image. The workflow mirrors the single-cell vignette but adds spatial plots to show score and prognostic group across the slide.

## Load data

Here, we use a 10X Visium spatial transcriptomics sample from a pancreatic cancer dataset available from [HTAN](https://humantumoratlas.org/). The sample has been preprocessed and has paired single cell RNAseq data mapped to the Visium spots using [**CytoSPACE**](https://www.nature.com/articles/s41587-023-01697-9).

```{r load-setup}
suppressPackageStartupMessages({
  library(PhenoMapR)
  library(Seurat)
  library(ggplot2)
  library(dplyr)
})

knitr::opts_chunk$set(fig.width = 12, out.width = "100%", warning = FALSE)
theme_set(theme_minimal(base_size = 14))

rds_name <- "HT270P1-S1H2Fc2U1Z1Bs1-H2Bs2-Test_processed.rds"
spatial_url <- Sys.getenv("PHENOMAPR_SPATIAL_RDS_URL", unset = "")
if (nzchar(spatial_url)) {
  if (requireNamespace("curl", quietly = TRUE)) {
    curl::curl_download(spatial_url, rds_name, quiet = TRUE)
  } else {
    utils::download.file(spatial_url, rds_name, mode = "wb", quiet = TRUE)
  }
} else {
  suppressPackageStartupMessages(library(googledrive))
  googledrive::drive_deauth()
  googledrive::drive_download(googledrive::as_id("1HM0dBrQnaNsdm5mnq23aaQ2ILofJ0_vj"), rds_name, overwrite = TRUE)
}

  suppressMessages({
    seurat <- PhenoMapR::load_rds_fast(rds_name)
    update_fn <- tryCatch(
      get("UpdateSeuratObject", envir = asNamespace("SeuratObject")),
      error = function(e) get0("UpdateSeuratObject", envir = asNamespace("Seurat"), mode = "function")
    )
    if (is.function(update_fn)) {
      seurat <- tryCatch(update_fn(seurat), error = function(e) seurat)
    }
  })

  if (exists("spatial_ci_lite") && isTRUE(spatial_ci_lite) && ncol(seurat) > 1L) {
    set.seed(1L)
    nkeep <- min(2000L, as.integer(ncol(seurat)))
    if (ncol(seurat) > nkeep) {
      seurat <- subset(seurat, cells = sample(colnames(seurat), nkeep))
      message("spatial_ci_lite=TRUE: using ", ncol(seurat), " cells/spots to reduce CI memory (exit 143).")
    }
  }

  assay_use <- if ("Spatial" %in% names(seurat@assays)) "Spatial" else "RNA"
  n_genes <- nrow(seurat)
  n_spots <- ncol(seurat)
  n_images <- length(seurat@images)
```

## Score sample with PhenoMapR

Apply **{PhenoMapR}** using the built-in PRECOG primary **Pancreatic** reference [[1]](#ref1), then add scores to the Seurat object metadata.

```{r score-spots}
suppressMessages({
  scores_spatial <- PhenoMapR::PhenoMap(
    expression = seurat,
    reference = "precog",
    cancer_type = "Pancreatic",
    assay = assay_use,
    slot = "data",
    verbose = FALSE
  )
  seurat <- PhenoMapR::add_scores_to_seurat(seurat, scores_spatial)
})
score_col <- grep("weighted_sum_score", names(scores_spatial), value = TRUE)[1]
```

## Score distribution

Distribution of PhenoMapR (PRECOG Pancreatic) scores across spots.

```{r score-distribution}
plot_score_distribution(
  seurat@meta.data[[score_col]],
  main = "PRECOG Pancreatic score distribution",
  base_size = 14
)
```

## Score by cell type

Before looking at any spatial information, we plot the **PhenoMapR** score by cell type to see cell types most enriched in the adverse and favorable prognostic groups.

```{r score-by-annotation}
spatial_celltype_pal <- NULL
spatial_celltype_col <- NULL
meta_names <- names(seurat@meta.data)
celltype_col <- "CellType"

df <- seurat@meta.data
n_meta <- nrow(df)
ann_vec <- NULL
if (!is.null(celltype_col)) {
  raw <- df[[celltype_col]]
  if (is.list(raw)) {
    ann_vec <- vapply(raw, function(x) if (is.null(x) || length(x) == 0) NA_character_ else as.character(x)[1], character(1))
  } else {
    ann_vec <- as.vector(raw)
  }
  if (length(ann_vec) != n_meta) ann_vec <- NULL
}

  df$annotation <- factor(ann_vec, exclude = NULL)

  pal <- PhenoMapR::get_celltype_palette(levels(df$annotation))
  print(ggplot(df, aes(
    x = reorder(.data$annotation, .data[[score_col]], FUN = median),
    y = .data[[score_col]],
    fill = .data$annotation
  )) +
    geom_boxplot(outlier.alpha = 0.3) +
    scale_fill_manual(values = pal, name = celltype_col) +
    theme_minimal(base_size = 14) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
      legend.position = "right"
    ) +
    labs(y = "PRECOG Pancreatic score", x = celltype_col, title = "Score by annotation (spatial)"))

```
As see in the other Pancreatic caner vignettes, the ductal cell type is the most associated with the adverse prognostic signal, while the beta and delta cells are most associated with favorable signal. Interestingly, plasma cells show quite a wide range of score and contain some of the most favorably prognostic cells across the sample. 


## Define phenotype groups

Label cells as Most Adverse (top 5%), Most Favorable (bottom 5%), or Other, then attach to metadata for spatial plotting.

```{r prognostic-groups}
scores_df <- seurat@meta.data[, grep("weighted_sum_score", names(seurat@meta.data)), drop = FALSE]
groups <- PhenoMapR::define_phenotype_groups(scores_df, percentile = 0.05)
suppressMessages(seurat <- AddMetaData(seurat, groups))

group_col <- paste0("phenotype_group_", score_col)
if (!group_col %in% names(seurat@meta.data)) group_col <- grep("phenotype_group", names(seurat@meta.data), value = TRUE)[1]
table(seurat@meta.data[[group_col]], useNA = "ifany")
```

## Build spatial locations dataframe for plotting

Here, we use the coordinates information from the spatial seurat object and pair them with the cell level metadata in order to plot the results in spatial contect.

```{r spatial-build-df}
coords <- tryCatch(
  suppressMessages(Seurat::GetTissueCoordinates(seurat)),
  error = function(e) NULL
)
if (is.null(coords) && length(seurat@images) > 0L) {
  img <- seurat@images[[1]]
  coords <- tryCatch(img@coordinates, error = function(e) NULL)
}
# Fallback: look for row/col or imagerow/imagecol in metadata
if (is.null(coords) || nrow(coords) == 0) {
  meta <- seurat@meta.data
  for (pair in list(
    c("row", "col"),
    c("imagerow", "imagecol"),
    c("x", "y"),
    c("tile", "cell")
  )) {
    if (pair[1] %in% names(meta) && pair[2] %in% names(meta)) {
      coords <- meta[, pair, drop = FALSE]
      names(coords) <- c("row", "col")
      break
    }
  }
}
if (!is.null(coords) && nrow(coords) > 0) {
  if (!"row" %in% names(coords)) coords$row <- coords[[grep("row|imagerow|x", names(coords), ignore.case = TRUE)[1]]]
  if (!"col" %in% names(coords)) coords$col <- coords[[grep("col|imagecol|y", names(coords), ignore.case = TRUE)[1]]]
  spatial_df <- as.data.frame(seurat@meta.data)
  spatial_df$cell_id <- rownames(spatial_df)
  # Align coordinates by cell id (coords rownames match cell barcodes)
  ids <- spatial_df$cell_id
  spatial_df$row <- coords[match(ids, rownames(coords)), "row"]
  spatial_df$col <- coords[match(ids, rownames(coords)), "col"]
  if (!is.null(group_col)) spatial_df$prognostic_group <- spatial_df[[group_col]]
  # Cell type for spatial: use spatial_celltype_col if set, else find CellType or any 2--100 level column
  celltype_plot_col <- NULL
  if (!is.null(spatial_celltype_col) && spatial_celltype_col %in% names(spatial_df)) {
    celltype_plot_col <- spatial_celltype_col
  } else if ("CellType" %in% names(spatial_df)) {
    celltype_plot_col <- "CellType"
  } else {
    idx <- grep("celltype|cell_type|annotation", names(spatial_df), ignore.case = TRUE)
    if (length(idx) > 0) celltype_plot_col <- names(spatial_df)[idx[1]]
  }
  if (is.null(celltype_plot_col)) {
    for (c in setdiff(names(spatial_df), c(score_col, group_col, "cell_id", "row", "col", "prognostic_group"))) {
      raw <- spatial_df[[c]]
      if (is.list(raw)) raw <- vapply(raw, function(x) as.character(x)[1], character(1))
      else raw <- as.character(raw)
      if (length(unique(na.omit(raw))) >= 2L && length(unique(na.omit(raw))) <= 100L) {
        celltype_plot_col <- c
        break
      }
    }
  }
  if (!is.null(celltype_plot_col)) {
    spatial_df$celltype_plot <- factor(spatial_df[[celltype_plot_col]], exclude = NULL)
    spatial_df_celltype_col <- celltype_plot_col
    n_lev <- length(levels(spatial_df$celltype_plot))
    if (n_lev >= 2L && (is.null(spatial_celltype_pal) || length(spatial_celltype_pal) != n_lev)) {
      spatial_celltype_pal <- PhenoMapR::get_celltype_palette(levels(spatial_df$celltype_plot))
      spatial_celltype_col <- celltype_plot_col
    }
  } else {
    spatial_df$celltype_plot <- NULL
    spatial_df_celltype_col <- NULL
  }
  spatial_df <- spatial_df %>%
    group_by(.data$row, .data$col) %>%
    mutate(points_per_location = n()) %>%
    ungroup()
  # Scaled PhenoMapR score (z-score) for gradient plot so colors show variation
  if (score_col %in% names(spatial_df)) {
    raw_score <- as.numeric(spatial_df[[score_col]])
    spatial_df$phenomapr_scaled <- as.numeric(scale(raw_score))
  } else {
    spatial_df$phenomapr_scaled <- NA_real_
  }
  # Shared jitter and point size so all three spatial plots match (multiple cells per spot)
  rng_row <- diff(range(spatial_df$row, na.rm = TRUE))
  rng_col <- diff(range(-spatial_df$col, na.rm = TRUE))
  spatial_jitter_w <- max(0.15, if (rng_row > 0) rng_row * 0.025 else 0.15)
  spatial_jitter_h <- max(0.15, if (rng_col > 0) rng_col * 0.025 else 0.15)
  spatial_point_range <- c(0.5, 1.6)
  has_spatial_coords <- TRUE
} else {
  has_spatial_coords <- FALSE
  message("No spatial coordinates found; skipping spatial plots.")
}
```

### Where cell types are

Here, we plot the location of the different cell types across the sample. Points are sized based on the number of cells mapped to each spot (CytoSPACE) and jittered in order to see all points when multiple cells map to the same spot.

```{r spatial-celltype-plot, fig.width=6, fig.height=5.5, out.width=NULL}
if (exists("has_spatial_coords") && has_spatial_coords && !is.null(spatial_df$celltype_plot) &&
    length(levels(spatial_df$celltype_plot)) >= 2L) {
  ct_col <- if (exists("spatial_df_celltype_col") && !is.null(spatial_df_celltype_col)) spatial_df_celltype_col else spatial_celltype_col
  ct_freq <- sort(table(spatial_df$celltype_plot, useNA = "no"), decreasing = TRUE)
  ct_order <- names(ct_freq)
  spatial_df$celltype_zorder <- as.numeric(factor(as.character(spatial_df$celltype_plot), levels = ct_order))
  ct_pal <- if (!is.null(spatial_celltype_pal)) spatial_celltype_pal else PhenoMapR::get_celltype_palette(levels(spatial_df$celltype_plot))
  p3 <- ggplot(spatial_df, aes(x = .data$row, y = -.data$col, color = .data$celltype_plot,
    size = points_per_location, zorder = .data$celltype_zorder)) +
    geom_jitter(alpha = 0.8, width = spatial_jitter_w, height = spatial_jitter_h, shape = 16) +
    ggtitle(paste("Cell type (", ct_col, ")", sep = "")) +
    scale_color_manual(values = ct_pal, name = ct_col, na.value = "grey90") +
    scale_size_continuous(range = spatial_point_range, trans = "reverse", name = "Cells per spot") +
    theme_minimal(base_size = 14) +
    theme(
      plot.title = element_text(hjust = 0.5, size = 14),
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      axis.title = element_blank()
    )
  print(p3)
}
```

### Where raw PhenoMapR scores are

Spatial map of PhenoMapR score (z-scaled for color gradient). Blue = more favorable, red = more adverse.

```{r spatial-score, fig.width=6, fig.height=5.5, out.width=NULL}
if (exists("has_spatial_coords") && has_spatial_coords) {
  score_vals <- spatial_df$phenomapr_scaled
  p1 <- ggplot(spatial_df, aes(x = .data$row, y = -.data$col, color = score_vals,
    size = points_per_location, zorder = score_vals)) +
    geom_jitter(alpha = 0.8, width = spatial_jitter_w, height = spatial_jitter_h, shape = 16) +
    ggtitle("PhenoMapR score (PRECOG Pancreatic, z-scaled)") +
    scale_color_gradientn(
      colors = colorRampPalette(c("#2166AC", "#F7F7F7", "#B2182B"))(100),
      name = "PhenoMapR\nscore (z)",
      limits = c(-2, 2),
      na.value = "grey80"
    ) +
    scale_size_continuous(range = spatial_point_range, trans = "reverse", name = "Cells per spot") +
    theme_minimal(base_size = 14) +
    theme(
      plot.title = element_text(hjust = 0.5, size = 14),
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      axis.title = element_blank()
    )
  print(p1)
}
```

### Where 5th percentile cells are

In order to make the visualization more apparent, we restrict the spatial map of prognostic groups to: top 5% (Most Adverse), bottom 5% (Most Favorable), and the rest (Other).

```{r spatial-group, fig.width=6, fig.height=5.5, out.width=NULL}
if (exists("has_spatial_coords") && has_spatial_coords && !is.null(group_col)) {
  pg <- as.character(spatial_df$prognostic_group)
  df_other <- spatial_df[pg == "Other" | is.na(pg), ]
  df_extreme <- spatial_df[pg %in% c("Most Adverse", "Most Favorable"), ]
  p2 <- ggplot() +
    geom_jitter(data = df_other, aes(x = .data$row, y = -.data$col, color = .data$prognostic_group,
      size = points_per_location), alpha = 0.8, width = spatial_jitter_w, height = spatial_jitter_h, shape = 16) +
    geom_jitter(data = df_extreme, aes(x = .data$row, y = -.data$col, color = .data$prognostic_group,
      size = points_per_location), alpha = 0.8, width = spatial_jitter_w, height = spatial_jitter_h, shape = 16) +
    ggtitle("5th percentile: Most Adverse vs Most Favorable") +
    scale_color_manual(
      values = c(`Most Adverse` = "#B2182B", Other = "#f7f7f7", `Most Favorable` = "#2166AC"),
      name = "Prognostic group",
      na.value = "grey90",
      drop = FALSE
    ) +
    scale_size_continuous(range = spatial_point_range, trans = "reverse", name = "Cells per spot") +
    theme_minimal(base_size = 14) +
    theme(
      plot.title = element_text(hjust = 0.5, size = 14),
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      axis.title = element_blank()
    )
  print(p2)
}
```

It looks like the adverse and favorable cells tend to co-localize independently (adverse with adverse, and favorable with favorable).


## Prognostic markers

`find_phenotype_markers()` supports the same **marker scopes** on spatial data (cells or spots mapped to spots) as in the single-cell vignette:

- **Cell type agnostic** — `marker_scope = "phenotype_groups"` (**default**): **Most Adverse** vs all other cells with a phenotype label, and **Most Favorable** vs all other cells, **ignoring** cell type. This asks which genes distinguish the prognostic extremes in the whole mixture on the slide.
- **Cell type specific** — `marker_scope = "cell_type_specific"`: for each cell type in metadata (here, CytoSPACE-mapped types), contrasts that type in the adverse or favorable tail vs a reference set controlled by `celltype_contrast`. **Default** `celltype_contrast = "within_cell_type"` compares to other phenotype bins **within the same type** (reduces housekeeping genes repeating across type × tail blocks). **`celltype_contrast = "vs_cohort_rest"`** restores the **original** cohort-wide contrast: (type ∩ tail) vs **every other cell**. With **`celltype_unique_genes = TRUE`** (default), each gene is kept in at most one adverse/favorable row (strongest `avg_log2FC`).

Below, **Step 1–2** use **cell-type-agnostic** markers and a **pheatmap**. **Step 3** runs **cell-type-specific** markers and, when **ComplexHeatmap** is installed, draws **`plot_phenotype_markers(..., heatmap_type = "cell_type_specific")`** (as in the **single-cell** package vignette).

### Step 1: Markers for adverse vs. favorable spots (cell type agnostic)

```{r markers-spatial, cache = FALSE}
markers <- NULL
assay_markers <- NULL
if (!is.null(group_col)) {
  cells <- colnames(seurat)
  group_vec <- seurat@meta.data[cells, group_col]
  group_df <- data.frame(
    cell_id = cells,
    phenotype_group = as.character(group_vec),
    stringsAsFactors = FALSE
  )
  for (a in unique(c(assay_use, "RNA", "SCT"))) {
    if (!a %in% names(seurat@assays)) next
    markers <- tryCatch(
      PhenoMapR::find_phenotype_markers(
        seurat,
        group_labels = group_df,
        group_column = "phenotype_group",
        cell_id_column = "cell_id",
        marker_scope = "phenotype_groups",
        assay = a,
        slot = "data",
        max_cells_per_ident = spatial_max_cells,
        verbose = FALSE
      ),
      error = function(e) {
        message("find_phenotype_markers (assay ", a, ") error: ", conditionMessage(e))
        NULL
      }
    )
    if (!is.null(markers) && (nrow(markers$adverse_markers) > 0 || nrow(markers$favorable_markers) > 0)) {
      assay_markers <- a
      break
    }
  }
  if (!is.null(markers)) {
    message("Adverse markers (top 5):")
    print(head(markers$adverse_markers, 5))
    message("Favorable markers (top 5):")
    print(head(markers$favorable_markers, 5))
  } else {
    message("Marker analysis returned no results. Check that Most Adverse and Most Favorable groups have enough cells.")
  }
}
```

### Step 2: Heatmap of adverse vs. favorable markers (global phenotype groups)

```{r heatmap-markers-spatial}
if (exists("markers") && !is.null(markers)) {
  n_top <- if (exists("spatial_ci_lite") && isTRUE(spatial_ci_lite)) 10L else 15L
  expr <- NULL
  assay_order <- unique(c(
    if (exists("assay_markers") && !is.null(assay_markers)) assay_markers else character(0),
    assay_use, "RNA", "SCT"
  ))
  for (a in assay_order) {
    if (!a %in% names(seurat@assays)) next
    expr <- tryCatch(
      Seurat::GetAssayData(seurat, layer = "data", assay = a),
      error = function(e) tryCatch(Seurat::GetAssayData(seurat, slot = "data", assay = a), error = function(e2) NULL)
    )
    if (!is.null(expr) && nrow(expr) > 0 && ncol(expr) > 0) break
    expr <- tryCatch(
      SeuratObject::LayerData(seurat, layer = "data", assay = a),
      error = function(e) NULL
    )
    if (is.null(expr) || ncol(expr) == 0) {
      expr <- tryCatch(
        SeuratObject::LayerData(seurat, layer = "counts", assay = a),
        error = function(e) tryCatch(SeuratObject::LayerData(seurat, layer = NULL, assay = a), error = function(e2) NULL)
      )
    }
    if (!is.null(expr) && nrow(expr) > 0 && ncol(expr) > 0) break
  }
  if (!is.null(expr) && ncol(expr) > 0) {
  adverse_pos <- markers$adverse_markers
  if (nrow(adverse_pos) > 0 && "avg_log2FC" %in% names(adverse_pos)) adverse_pos <- adverse_pos[adverse_pos$avg_log2FC > 0, ]
  favorable_pos <- markers$favorable_markers
  if (nrow(favorable_pos) > 0 && "avg_log2FC" %in% names(favorable_pos)) favorable_pos <- favorable_pos[favorable_pos$avg_log2FC > 0, ]
  pcol <- if ("p_adj" %in% names(markers$adverse_markers)) "p_adj" else if ("p_val_adj" %in% names(markers$adverse_markers)) "p_val_adj" else "p_val"
  top_genes <- unique(c(
    if (nrow(adverse_pos) > 0 && "gene" %in% names(adverse_pos)) head(adverse_pos$gene[order(adverse_pos[[pcol]])], n_top) else character(0),
    if (nrow(favorable_pos) > 0 && "gene" %in% names(favorable_pos)) head(favorable_pos$gene[order(favorable_pos[[pcol]])], n_top) else character(0)
  ))
  top_genes <- top_genes[top_genes %in% rownames(expr)]
  if (length(top_genes) == 0) top_genes <- head(rownames(expr), 20)

  cells_expr <- colnames(expr)
  if (is.null(cells_expr)) cells_expr <- character(0)
  cells_obj <- colnames(seurat)
  cells_meta <- rownames(seurat@meta.data)
  cells_use <- intersect(cells_expr, cells_obj)
  if (length(cells_use) == 0 && length(cells_meta) > 0) {
    cells_use <- intersect(cells_expr, cells_meta)
  }
  if (length(cells_use) == 0 && length(cells_expr) > 0 && any(grepl("_", cells_expr, fixed = TRUE))) {
    stripped <- sub("^[^_]+_", "", cells_expr)
    cells_use <- intersect(stripped, cells_obj)
    if (length(cells_use) == 0) cells_use <- intersect(stripped, cells_meta)
    if (length(cells_use) > 0) {
      expr_cols <- vapply(cells_use, function(c) cells_expr[which(stripped == c)[1]], character(1))
      expr <- expr[, expr_cols, drop = FALSE]
      colnames(expr) <- cells_use
    }
  }
  if (length(cells_use) == 0) {
    message("No overlapping cells between expression and metadata; skipping heatmap. ",
            "expr ncol=", length(cells_expr), ", obj ncol=", length(cells_obj),
            if (length(cells_expr) > 0) paste0("; expr sample: ", head(cells_expr, 2)) else "")
  } else {
  mat <- as.matrix(expr[top_genes, cells_use, drop = FALSE])
  mat_scaled <- t(scale(t(mat)))
  mat_scaled[mat_scaled < -3] <- -3
  mat_scaled[mat_scaled > 3] <- 3
  meta <- seurat@meta.data[cells_use, , drop = FALSE]
  ord <- order(meta[[score_col]])
  mat_scaled <- mat_scaled[, ord, drop = FALSE]
  meta_ord <- meta[ord, , drop = FALSE]

  group_col_heatmap <- paste0("phenotype_group_", score_col)
  if (!group_col_heatmap %in% names(meta_ord)) group_col_heatmap <- group_col

  if (requireNamespace("pheatmap", quietly = TRUE)) {
    ann_col <- data.frame(
      `PhenoMapR Score` = meta_ord[[score_col]],
      `Prognostic group` = factor(meta_ord[[group_col_heatmap]], levels = c("Most Adverse", "Other", "Most Favorable")),
      check.names = FALSE
    )
    ct_col_ann <- if (exists("spatial_df_celltype_col") && !is.null(spatial_df_celltype_col)) spatial_df_celltype_col else if (exists("celltype_col") && !is.null(celltype_col)) celltype_col else NULL
    if (!is.null(ct_col_ann) && ct_col_ann %in% names(meta_ord)) {
      ann_col$`Cell type` <- factor(meta_ord[[ct_col_ann]])
    }
    rownames(ann_col) <- colnames(mat_scaled)
    pal_score <- colorRampPalette(c("#2166AC", "#F7F7F7", "#B2182B"))(100)
    pal_group <- c(`Most Adverse` = "#B2182B", Other = "#f7f7f7", `Most Favorable` = "#2166AC")
    pal_celltype <- if ("Cell type" %in% names(ann_col)) {
      PhenoMapR::get_celltype_palette(levels(ann_col$`Cell type`))
    } else list()
    ann_colors <- list(
      `PhenoMapR Score` = pal_score,
      `Prognostic group` = pal_group
    )
    if (length(pal_celltype) > 0) ann_colors$`Cell type` <- pal_celltype
    heatmap_colors <- if (requireNamespace("paletteer", quietly = TRUE)) {
      colorRampPalette(paletteer::paletteer_d("MexBrewer::Vendedora"))(100)
    } else {
      colorRampPalette(c("#2166AC", "#F7F7F7", "#B2182B"))(100)
    }
    pheatmap::pheatmap(
      mat_scaled,
      scale = "none",
      cluster_cols = FALSE,
      cluster_rows = TRUE,
      show_colnames = FALSE,
      annotation_col = ann_col,
      annotation_colors = ann_colors,
      color = heatmap_colors,
      breaks = seq(-3, 3, length.out = 101),
      main = "Top adverse & favorable marker genes (spatial spots)",
      fontsize = 12,
      fontsize_row = 10,
      treeheight_row = 25
    )
  }
  }
  } else {
    msg <- "Could not retrieve expression data for heatmap"
    if (!is.null(expr)) {
      msg <- paste0(msg, " (expr nrow=", nrow(expr), ", ncol=", ncol(expr), "; try LayerData for Seurat v5)")
    }
    message(msg, ".")
  }
}
```

### Stacked barplot: adverse vs. favorable by cell type

Each column is a cell type; the height is filled by the number of adverse (5th percentile) or favorable (5th percentile) cells.

```{r stacked-bar-adverse-favorable-by-celltype}
if (!is.null(group_col)) {
  ct_col_bar <- NULL
  if (exists("spatial_df_celltype_col") && !is.null(spatial_df_celltype_col) && spatial_df_celltype_col %in% names(seurat@meta.data)) {
    ct_col_bar <- spatial_df_celltype_col
  } else if (exists("celltype_col") && !is.null(celltype_col) && celltype_col %in% names(seurat@meta.data)) {
    ct_col_bar <- celltype_col
  } else {
    for (c in c("CellType", "Celltype..major.lineage.", "cell_type", "celltype", "annotation", "seurat_clusters")) {
      if (c %in% names(seurat@meta.data) && length(unique(na.omit(seurat@meta.data[[c]]))) >= 2) {
        ct_col_bar <- c
        break
      }
    }
  }
  if (!is.null(ct_col_bar)) {
    meta <- seurat@meta.data
    meta$pg <- meta[[group_col]]
    meta$ct <- as.character(meta[[ct_col_bar]])
    meta$ct_ok <- !is.na(meta$ct) & nzchar(trimws(meta$ct))
    idx_extreme <- meta$pg %in% c("Most Adverse", "Most Favorable") & meta$ct_ok
    df_bar <- as.data.frame(table(
      CellType = meta$ct[idx_extreme],
      Prognostic_group = meta$pg[idx_extreme],
      useNA = "no"
    ))
    if (nrow(df_bar) > 0) {
      df_bar$Prognostic_group <- factor(df_bar$Prognostic_group, levels = c("Most Favorable", "Most Adverse"))
      pal_bar <- c(`Most Adverse` = "#B2182B", `Most Favorable` = "#2166AC")
      adverse <- df_bar[df_bar$Prognostic_group == "Most Adverse", c("CellType", "Freq")]
      fav <- df_bar[df_bar$Prognostic_group == "Most Favorable", c("CellType", "Freq")]
      names(adverse)[2] <- "n_adverse"
      names(fav)[2] <- "n_fav"
      df_labels <- merge(adverse, fav, by = "CellType", all = TRUE)
      df_labels$n_adverse[is.na(df_labels$n_adverse)] <- 0
      df_labels$n_fav[is.na(df_labels$n_fav)] <- 0
      df_labels$label <- paste0(df_labels$n_adverse, "/", df_labels$n_fav)
      df_labels$total <- df_labels$n_adverse + df_labels$n_fav
      ct_ord <- levels(reorder(df_bar$CellType, df_bar$Freq, function(x) -sum(x)))
      df_labels$CellType <- factor(df_labels$CellType, levels = ct_ord)
      p_bar <- ggplot(df_bar, aes(x = reorder(.data$CellType, .data$Freq, function(x) -sum(x)), y = .data$Freq, fill = .data$Prognostic_group)) +
        geom_col(position = "stack") +
        geom_text(data = df_labels, aes(x = .data$CellType, y = .data$total, label = .data$n_adverse),
                  inherit.aes = FALSE, vjust = -0.3, size = 3.5, color = "#B2182B", hjust = 1.05) +
        geom_text(data = df_labels, aes(x = .data$CellType, y = .data$total, label = paste0("/", .data$n_fav)),
                  inherit.aes = FALSE, vjust = -0.3, size = 3.5, color = "#2166AC", hjust = -0.05) +
        scale_fill_manual(values = pal_bar) +
        labs(x = "Cell type", y = "Number of cells", fill = "Prognostic group",
             title = "Adverse vs. favorable cells (5th percentile) by cell type") +
        theme_minimal(base_size = 14) +
        theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12), legend.position = "top")
      print(p_bar)
    }
  }
}
```
In line with the other Pancreatic cancer vignettes, PhenoMapR identifies ductal cells (and some fibroblasts) as the most associated cell type with an adverse PhenoMapR prognostic score. Alpha, plasma, and beta cells are the most associated with the more favorable prognostic signal. Interestingly, fibroblasts seems to comprise both adverse and favorable phenotypes.


### Step 3: Cell-type-specific phenotype markers

Here we call `find_phenotype_markers()` with **`marker_scope = "cell_type_specific"`**, using the same cell type column as elsewhere in this vignette. The default **`celltype_contrast = "within_cell_type"`** contrasts each prognostic tail **within** a mapped cell type to the other bins in that type. For the legacy cohort-wide reference (tail vs all other cells), set **`celltype_contrast = "vs_cohort_rest"`** (see `?find_phenotype_markers`).

When both adverse and favorable marker tables are non-empty and **ComplexHeatmap** is installed, **`plot_phenotype_markers(..., heatmap_type = "cell_type_specific")`** matches the single-cell vignette (column order: phenotype bin, cell type, score; row blocks with optional gene labels).

```{r markers-celltype-specific-spatial, cache = FALSE, fig.align='center', fig.width=14, fig.height=10, eval=!spatial_ci_lite}
markers_ct_spatial <- NULL
ct_col_spatial <- NULL
if (!is.null(group_col)) {
  if (exists("spatial_df_celltype_col") && !is.null(spatial_df_celltype_col) &&
      spatial_df_celltype_col %in% names(seurat@meta.data)) {
    ct_col_spatial <- spatial_df_celltype_col
  } else if (!is.null(celltype_col) && celltype_col %in% names(seurat@meta.data)) {
    ct_col_spatial <- celltype_col
  } else {
    for (c in c("CellType", "Celltype..major.lineage.", "cell_type", "celltype", "annotation", "seurat_clusters")) {
      if (c %in% names(seurat@meta.data) && length(unique(na.omit(seurat@meta.data[[c]]))) >= 2) {
        ct_col_spatial <- c
        break
      }
    }
  }
  if (!is.null(ct_col_spatial)) {
    cells_ct <- colnames(seurat)
    group_df_ct <- data.frame(
      cell_id = cells_ct,
      phenotype_group = as.character(seurat@meta.data[cells_ct, group_col]),
      cell_type = as.character(seurat@meta.data[cells_ct, ct_col_spatial]),
      stringsAsFactors = FALSE
    )
    assay_ct <- if (exists("assay_markers") && !is.null(assay_markers)) assay_markers else assay_use
    markers_ct_spatial <- tryCatch(
      PhenoMapR::find_phenotype_markers(
        seurat,
        group_labels = group_df_ct,
        group_column = "phenotype_group",
        cell_id_column = "cell_id",
        cell_type_column = "cell_type",
        marker_scope = "cell_type_specific",
        celltype_contrast = "within_cell_type",
        assay = assay_ct,
        slot = "data",
        max_cells_per_ident = spatial_max_cells,
        verbose = FALSE
      ),
      error = function(e) {
        message("cell_type_specific find_phenotype_markers: ", conditionMessage(e))
        NULL
      }
    )
    if (!is.null(markers_ct_spatial)) {
      message("Cell-type-specific adverse markers (top 5):")
      print(utils::head(markers_ct_spatial$adverse_markers, 5))
      message("Cell-type-specific favorable markers (top 5):")
      print(utils::head(markers_ct_spatial$favorable_markers, 5))
    }
    expr_pm <- NULL
    for (a in unique(c(assay_ct, assay_use, "RNA", "SCT"))) {
      if (!a %in% names(seurat@assays)) next
      expr_pm <- tryCatch(
        as.matrix(Seurat::GetAssayData(seurat, layer = "data", assay = a)),
        error = function(e) tryCatch(
          as.matrix(Seurat::GetAssayData(seurat, slot = "data", assay = a)),
          error = function(e2) NULL
        )
      )
      if (!is.null(expr_pm) && nrow(expr_pm) > 0 && ncol(expr_pm) > 0) break
      expr_pm <- tryCatch(
        as.matrix(SeuratObject::LayerData(seurat, layer = "data", assay = a)),
        error = function(e) NULL
      )
      if (!is.null(expr_pm) && nrow(expr_pm) > 0 && ncol(expr_pm) > 0) break
    }
    if (!is.null(markers_ct_spatial) && !is.null(expr_pm) && ncol(expr_pm) > 0) {
      meta_pm <- as.data.frame(seurat@meta.data[cells_ct, , drop = FALSE])
      meta_pm$cell_id_plot <- cells_ct
      if (all(cells_ct %in% colnames(expr_pm))) {
        expr_pm <- expr_pm[, cells_ct, drop = FALSE]
      }
      pal_ct <- PhenoMapR::get_celltype_palette(levels(factor(meta_pm[[ct_col_spatial]])))
      if (requireNamespace("ComplexHeatmap", quietly = TRUE) && requireNamespace("circlize", quietly = TRUE)) {
        PhenoMapR::plot_phenotype_markers(
          markers = markers_ct_spatial,
          expr_mat = expr_pm,
          meta = meta_pm,
          cell_id_col = "cell_id_plot",
          group_col = group_col,
          score_col = score_col,
          celltype_col = ct_col_spatial,
          celltype_palette = pal_ct,
          heatmap_type = "cell_type_specific",
          top_n_markers = 20L,
          n_mark_labels = 5L,
          p_adj_threshold = 0.05,
          column_title = paste0(
            "Cell-type-specific phenotype markers (spatial; within_cell_type contrast; ",
            "use celltype_contrast = \"vs_cohort_rest\" for cohort-wide DE)"
          )
        )
      } else {
        message("Install packages 'ComplexHeatmap' and 'circlize' to draw the cell-type-specific marker heatmap.")
      }
    }
  } else {
    message("No suitable cell type column in metadata; skipping cell_type_specific markers.")
  }
}
```

## Summary

- **Data**: Processed spatial object (`HT270P1-S1H2Fc2U1Z1Bs1-H2Bs2-Test_processed.rds`).
- **Scoring**: PhenoMapR with PRECOG **Pancreatic**; scores added to Seurat metadata.
- **Plots**: Score distribution, score by annotation (if present), prognostic groups, and **spatial maps** of cell type, continuous score, and prognostic group (points ordered so less frequent groups are drawn on top for visibility).
- **Markers**: (1) **Cell-type-agnostic** `find_phenotype_markers(..., marker_scope = "phenotype_groups")` with a **pheatmap** of top genes; (2) **Cell-type-specific** `marker_scope = "cell_type_specific"` with default **`celltype_contrast = "within_cell_type"`** (or **`"vs_cohort_rest"`** for cohort-wide reference) and **`plot_phenotype_markers(..., heatmap_type = "cell_type_specific")`** when **ComplexHeatmap** is available.

## References

<a id="ref1"></a>
**[1]** Benard, B. A. et al. PRECOG update: an augmented resource of clinical outcome associations with gene expression for adult, pediatric, and immunotherapy cohorts. Nucleic Acids Res. 54, D1579–D1589 (2026).

## Session Info

```{r session-info}
sessionInfo()
```
