---
title: "Draft: Alzheimer's custom reference + single-cell scoring"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_collapsed: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, out.width = "100%")
# Data-dependent chunks below require local files; paths chunk sets eval accordingly.
```

## Purpose

This draft vignette shows how to:

1. Build a custom AD-vs-control reference signature from GEO bulk files:
   - `~/Downloads/GSE39420_series_matrix.txt.gz`
   - `~/Downloads/GPL11532.annot.gz`
2. Score all cells in a Seurat object:
   - `~/Downloads/adsn_seuV5.rds`
3. Visualize score distributions by cell type and generate phenotype marker heatmaps.

This file is intentionally a draft and is not wired into pkgdown navigation.

## Load packages and set input paths

```{r}
suppressPackageStartupMessages({
  library(PhenoMapR)
  library(Seurat)
  library(dplyr)
  library(ggplot2)
  library(ggpubr)
})

sc_rds_path <- file.path(Sys.getenv("HOME"), "Downloads", "adsn_seuV5.rds")
gpl_path <- file.path(Sys.getenv("HOME"), "Downloads", "GPL11532.annot.gz")
series_path <- file.path(Sys.getenv("HOME"), "Downloads", "GSE39420_series_matrix.txt.gz")

draft_data_ok <- file.exists(sc_rds_path) && file.exists(gpl_path) && file.exists(series_path)
knitr::opts_chunk$set(eval = draft_data_ok)
if (!draft_data_ok) {
  message(
    "Draft vignette: local GEO/Seurat files not found under ~/Downloads/; ",
    "skipping execution of data-dependent chunks (e.g. pkgdown/CI)."
  )
}
```

```{r draft-data-note, eval=TRUE, echo=FALSE, results="asis"}
if (!draft_data_ok) {
  cat(
    "_This draft requires local files in `~/Downloads/` (see Purpose). ",
    "Those files are absent here, so code is shown but not executed._\n\n"
  )
}
```

## Helper functions for GEO parsing

```{r helper-functions, eval=TRUE}
read_series_matrix_table <- function(series_matrix_gz) {
  lines <- readLines(gzfile(series_matrix_gz), warn = FALSE)

  i0 <- grep("^!series_matrix_table_begin", lines)
  i1 <- grep("^!series_matrix_table_end", lines)
  if (length(i0) != 1L || length(i1) != 1L || i1 <= i0) {
    stop("Could not find !series_matrix_table_begin / !series_matrix_table_end in series matrix.")
  }

  tab_lines <- lines[(i0 + 1L):(i1 - 1L)]
  expr <- read.delim(
    textConnection(tab_lines),
    header = TRUE,
    sep = "\t",
    check.names = FALSE,
    quote = "\"",
    comment.char = "",
    stringsAsFactors = FALSE
  )
  closeAllConnections()
  expr
}

parse_sample_meta_from_series <- function(series_matrix_gz) {
  lines <- readLines(gzfile(series_matrix_gz), warn = FALSE)
  sample_lines <- grep("^!Sample_", lines, value = TRUE)
  if (length(sample_lines) == 0L) {
    stop("No !Sample_ lines found in series matrix.")
  }

  split_tab <- strsplit(sample_lines, "\t", fixed = TRUE)
  keys <- sub("^!Sample_", "", vapply(split_tab, `[`, character(1), 1L))
  vals <- lapply(split_tab, function(x) gsub("^\"|\"$", "", x[-1L]))

  n <- unique(vapply(vals, length, integer(1)))
  if (length(n) != 1L) {
    stop("Sample metadata rows have inconsistent sample counts.")
  }

  meta <- as.data.frame(setNames(vals, keys), stringsAsFactors = FALSE)
  # If same key appears multiple times (e.g. characteristics_ch1), keep all with suffixes.
  names(meta) <- make.unique(names(meta))
  meta
}

normalize_colname <- function(x) {
  x <- tolower(trimws(as.character(x)))
  gsub("[^a-z0-9]", "", x)
}

pick_first_existing <- function(x, candidates) {
  x_norm <- normalize_colname(x)
  cand_norm <- normalize_colname(candidates)
  m <- match(cand_norm, x_norm)
  m <- m[!is.na(m)]
  if (length(m) == 0L) NA_character_ else x[m[1L]]
}

read_gpl_annotation <- function(gpl_gz) {
  lines <- readLines(gzfile(gpl_gz), warn = FALSE)
  i0 <- grep("^!platform_table_begin", lines)
  i1 <- grep("^!platform_table_end", lines)

  if (length(i0) == 1L && length(i1) == 1L && i1 > i0) {
    tab_lines <- lines[(i0 + 1L):(i1 - 1L)]
    gpl <- read.delim(
      textConnection(tab_lines),
      header = TRUE,
      sep = "\t",
      check.names = FALSE,
      quote = "\"",
      comment.char = "",
      fill = TRUE,
      stringsAsFactors = FALSE
    )
    closeAllConnections()
    return(gpl)
  }

  # Fallback: direct read (some GPL files are already plain tables at top).
  read.delim(
    gzfile(gpl_gz),
    sep = "\t",
    header = TRUE,
    check.names = FALSE,
    quote = "\"",
    comment.char = "!",
    fill = TRUE,
    stringsAsFactors = FALSE
  )
}

collapse_probe_to_gene <- function(expr_probe_by_sample, annot_df) {
  # Probe column
  probe_col <- pick_first_existing(
    names(annot_df),
    c("ID", "ID_REF", "Probe.Set.ID", "ProbeID", "Probe ID", "Platform ID", "SPOT_ID")
  )
  if (is.na(probe_col)) {
    # Last-resort fallback: first column is usually the platform probe ID.
    probe_col <- names(annot_df)[1L]
  }

  # Gene symbol column
  gene_col <- pick_first_existing(
    names(annot_df),
    c("Gene Symbol", "Gene.symbol", "Symbol", "GENE_SYMBOL", "GeneSymbol", "UniGene symbol")
  )
  if (is.na(gene_col)) {
    # Regex fallback for unusual label casing/spacing.
    nm <- names(annot_df)
    nm_norm <- normalize_colname(nm)
    idx <- grep("genesymbol|symbol", nm_norm)
    if (length(idx) > 0L) {
      gene_col <- nm[idx[1L]]
    } else {
      stop(
        "Could not infer gene symbol column in GPL annotation. Columns seen: ",
        paste(names(annot_df), collapse = ", ")
      )
    }
  }

  annot <- annot_df[, c(probe_col, gene_col), drop = FALSE]
  names(annot) <- c("probe", "gene_symbol")
  annot$probe <- as.character(annot$probe)
  annot$gene_symbol <- as.character(annot$gene_symbol)

  # Clean symbols: keep first entry if multi-mapped
  annot$gene_symbol <- trimws(annot$gene_symbol)
  annot$gene_symbol <- sub("///.*$", "", annot$gene_symbol)
  annot$gene_symbol <- sub(";.*$", "", annot$gene_symbol)
  annot$gene_symbol <- sub(",.*$", "", annot$gene_symbol)

  expr_df <- data.frame(
    probe = rownames(expr_probe_by_sample),
    expr_probe_by_sample,
    check.names = FALSE
  )

  merged <- merge(expr_df, annot, by = "probe", all.x = FALSE, all.y = FALSE)
  merged <- merged[!is.na(merged$gene_symbol) & nzchar(merged$gene_symbol), , drop = FALSE]
  if (nrow(merged) == 0L) stop("No probe-to-gene mappings remained after merge.")

  mat <- as.matrix(merged[, setdiff(names(merged), c("probe", "gene_symbol")), drop = FALSE])
  storage.mode(mat) <- "double"

  # Aggregate duplicated symbols by mean expression.
  genes <- merged$gene_symbol
  gene_levels <- unique(genes)
  out <- sapply(gene_levels, function(g) {
    idx <- which(genes == g)
    if (length(idx) == 1L) mat[idx, ] else colMeans(mat[idx, , drop = FALSE], na.rm = TRUE)
  })
  out <- t(out) # genes x samples
  rownames(out) <- gene_levels
  out
}
```

## Build custom reference from GEO bulk data (Control vs Early-onset AD only)

```{r}
# 1) Read expression table from series matrix
series_expr_raw <- read_series_matrix_table(series_path)
stopifnot("ID_REF" %in% names(series_expr_raw))

probe_ids <- as.character(series_expr_raw$ID_REF)
bulk_probe <- as.matrix(series_expr_raw[, setdiff(names(series_expr_raw), "ID_REF"), drop = FALSE])
storage.mode(bulk_probe) <- "double"
rownames(bulk_probe) <- probe_ids

# 2) Read sample metadata from the same series matrix file
sample_meta <- parse_sample_meta_from_series(series_path)
if (!"geo_accession" %in% names(sample_meta)) {
  # Fallback: expression column names are usually GSM IDs
  sample_meta$geo_accession <- colnames(bulk_probe)
}

# 3) Read GPL annotation and map probes -> genes
gpl <- read_gpl_annotation(gpl_path)

bulk_gene <- collapse_probe_to_gene(bulk_probe, gpl)

# 4) Build phenotype groups from sample metadata and retain only:
#    - Control
#    - Early-onset AD
candidate_cols <- names(sample_meta)[
  grepl("characteristics|disease|diagnosis|status|group|phenotype|title|source", names(sample_meta), ignore.case = TRUE)
]
if (length(candidate_cols) == 0L) {
  stop("No candidate phenotype columns found in sample metadata; inspect 'sample_meta'.")
}

meta_text <- apply(sample_meta[, candidate_cols, drop = FALSE], 1, function(x) {
  paste(x, collapse = " | ")
})
meta_text <- tolower(meta_text)

is_ctrl <- grepl("control|normal|non[- ]?demented|healthy", meta_text)
is_early <- grepl("early[- ]?onset|eo[- ]?ad|sporadic", meta_text)


bulk_group <- ifelse(
  is_ctrl, "Control",
  ifelse(is_early, "EarlyAD", NA_character_)
)

pheno <- data.frame(
  sample_id = as.character(sample_meta$geo_accession),
  ad_status = factor(bulk_group, levels = c("EarlyAD", "Control")),
  stringsAsFactors = FALSE
)
pheno <- pheno[!is.na(pheno$ad_status), , drop = FALSE]

# Keep only samples that exist in expression matrix
common <- intersect(colnames(bulk_gene), pheno$sample_id)
bulk_gene <- bulk_gene[, common, drop = FALSE]
pheno <- pheno[match(common, pheno$sample_id), , drop = FALSE]

table(pheno$ad_status)

# 5) Derive custom reference z-scores
ad_reference <- derive_reference_from_bulk(
  bulk_expression = bulk_gene,              # genes x samples
  phenotype = pheno,
  sample_id_column = "sample_id",
  phenotype_column = "ad_status",
  phenotype_type = "binary",
  binary_positive_reference = "first",      # EarlyAD is positive class
  normalize = TRUE,
  hugo_species = "human",
  verbose = TRUE
)

head(ad_reference)
```

## Optional: visualize the custom reference

```{r, fig.height=2, fig.width=7.5}
plot_reference_signature(
  reference = ad_reference,
  n_label = 20,
  row_title = "AD vs Control z-score",
  legend_side = "bottom"
)
```

## Load Seurat object and score all cells

```{r}
seu <- readRDS(sc_rds_path)
if (!inherits(seu, "Seurat")) stop("Expected a Seurat object in adsn_seuV5.rds")

seu@meta.data <- seu@meta.data %>%
  mutate(cellType = recode(cellType,
    "oligo"  = "Oligodendrocyte",
    "astro"  = "Astrocyte",
    "OPC"    = "Oligodendrocyte Precursor Cell",
    "neuron" = "Neuron",
    "endo"   = "Endothelial",
    "mg"     = "Microglia"
  ))

assay_use <- Seurat::DefaultAssay(seu)
message("Using assay: ", assay_use)

# Identify sample and cell-type columns in metadata.
meta0 <- seu[[]]
sample_col <- "patient"
celltype_col <- "cellType"

# Remove Ct-un and AD-un samples, and remove unID/ad doublet cell types.
sample_id <- as.character(meta0[[sample_col]])
celltype_raw <- as.character(meta0[[celltype_col]])

drop_sample <- grepl("^(Ct|AD)-?un$", sample_id, ignore.case = TRUE)
drop_celltype <- grepl("unid|doublet", celltype_raw, ignore.case = TRUE)
cells_keep <- rownames(meta0)[!(drop_sample | drop_celltype)]
seu <- subset(seu, cells = cells_keep)
meta0 <- seu[[]]

# Annotate AD/Ctrl group from sample IDs (AD1-6, Ct1-6).
sample_id <- as.character(meta0[[sample_col]])
disease_group <- ifelse(
  grepl("^AD[1-6]$", sample_id, ignore.case = TRUE), "AD",
  ifelse(grepl("^Ct[1-6]$", sample_id, ignore.case = TRUE), "Ctrl", NA_character_)
)
keep2 <- rownames(meta0)[!is.na(disease_group)]
seu <- subset(seu, cells = keep2)
meta0 <- seu[[]]
meta0$disease_group <- ifelse(
  grepl("^AD[1-6]$", as.character(meta0[[sample_col]]), ignore.case = TRUE),
  "AD",
  "Ctrl"
)
seu$disease_group <- meta0$disease_group
message("Cells retained after filtering: ", ncol(seu))

# Use raw counts for scoring
scores <- PhenoMap(
  expression = seu,
  reference = ad_reference,
  assay = assay_use,
  slot = "counts",
  z_score_cutoff = 2,
  verbose = TRUE
)

score_col <- names(scores)[1]
seu <- add_scores_to_seurat(seu, scores)
score_z_col <- paste0(score_col, "_z")
seu[[score_z_col]] <- as.numeric(scale(seu[[score_col]][, 1]))
head(scores)
```

## UMAP with cell type labels

```{r, fig.width=10, fig.height=4}
meta <- seu[[]]
pal_cells <- get_celltype_palette(meta[[celltype_col]])

if (!"umap" %in% names(seu@reductions)) {
  message("No UMAP reduction found in this Seurat object (expected seu@reductions$umap). Skipping UMAP plot.")
} else {
  p_ct <- DimPlot(
    seu,
    reduction = "umap",
    group.by = celltype_col,
    cols = pal_cells,
    label = FALSE
  ) +
    ggtitle("Alzheimer's dataset: cell types on UMAP") +
    theme_void() +
    theme(legend.position = "right")

  um <- as.data.frame(seu@reductions$umap@cell.embeddings[, 1:2, drop = FALSE])
  colnames(um) <- c("UMAP_1", "UMAP_2")
  um$score_z <- seu@meta.data[[score_z_col]]
  # Plot low-|z| points first, extremes last (on top).
  um <- um[order(abs(um$score_z)), , drop = FALSE]
  # Use asymmetric limits to preserve more red-side dynamic range when negatives are extreme.
  q_hi <- as.numeric(stats::quantile(um$score_z, probs = 0.99, na.rm = TRUE))
  neg_abs <- abs(um$score_z[um$score_z < 0])
  q_lo_abs <- if (length(neg_abs) > 0) as.numeric(stats::quantile(neg_abs, probs = 0.9, na.rm = TRUE)) else 1
  lim_hi <- if (is.finite(q_hi) && q_hi > 0) q_hi else 1
  lim_lo <- if (is.finite(q_lo_abs) && q_lo_abs > 0) -q_lo_abs else -1

  p_score <- ggplot(um, aes(x = UMAP_1, y = UMAP_2, color = score_z)) +
    geom_point(size = 0.25, alpha = 0.9) +
    scale_color_gradient2(
      low = "#2166AC",
      mid = "#F7F7F7",
      high = "#B2182B",
      midpoint = 0,
      # limits = c(lim_lo, lim_hi),
      trans = scales::pseudo_log_trans(sigma = 0.2),
      oob = scales::squish
    ) +
    ggtitle("Alzheimer's dataset: PhenoMapR score (z) on UMAP") +
    theme_void() +
    theme(legend.position = "right")

  if (requireNamespace("patchwork", quietly = TRUE)) {
    suppressPackageStartupMessages(library(patchwork))
    print(p_ct + p_score + patchwork::plot_layout(widths = c(1, 1)))
  } else {
    print(p_ct)
    print(p_score)
  }
}
```

```{r, fig.height=2, fig.width=5.2}
um$Source <- seu@meta.data$disease_group

pal <- c(Ctrl = "#56884A", AD = "#6B4886")

p_source <- ggscatter(um, 
          x = "UMAP_1", 
          y = "UMAP_2",
          size = 1,
          color = "Source",
          palette = pal,
          legend = "right"
) + guides(color = guide_legend(override.aes = list(size = 2))) +
    labs(color = "Sample\nType") +
  theme_void() +
    ggtitle("Sample Type Distribution") + 
    theme(plot.title = element_text(hjust = 0.5))

p_score <- ggscatter(um, 
          x = "UMAP_1", 
          y = "UMAP_2",
          size = 1,
          color = "score_z",
          palette = pal,
          legend = "right"
) +
  guides(color = guide_colorbar(override.aes = list(size = 2))) +
  scale_color_gradient2(
    low      = "#2166AC",
    mid      = "#F7F7F7",
    high     = "#B2182B",
    midpoint = 0,
    trans = scales::pseudo_log_trans(sigma = 0.2)
  ) +
  labs(color = NULL) +
  theme_void() +
    ggtitle("PhenoMapR Score Distribution") 

print(p_source + p_score + patchwork::plot_layout(widths = c(1, 1)))
```

## Inspect score distributions

```{r}
meta <- seu[[]]
meta$Cell <- rownames(meta)
message("Using cell type column: ", celltype_col)

pal_cells <- get_celltype_palette(meta[[celltype_col]])
```

```{r, fig.height=3.5}
# Global distribution
plot_score_distribution(
  scores = seu@meta.data[[score_z_col]],
  main = "AD-reference PhenoMapR score distribution (z-scored)"
)
```

```{r, fig.height=3.5}
# Per-cell-type AD vs Ctrl distributions (side-by-side violins) with per-cell-type
# tests for differences in means.
plot_df <- meta %>%
  dplyr::filter(
    !is.na(.data[[celltype_col]]),
    !is.na(.data[[score_col]]),
    !is.na(.data$disease_group)
  ) %>%
  dplyr::mutate(disease_group = factor(.data$disease_group, levels = c("Ctrl", "AD")))

ct_order <- plot_df %>%
  dplyr::group_by(.data[[celltype_col]]) %>%
  dplyr::summarise(median_score = median(.data[[score_col]], na.rm = TRUE), .groups = "drop") %>%
  dplyr::arrange(median_score) %>%
  dplyr::pull(.data[[celltype_col]])

plot_df[[celltype_col]] <- factor(plot_df[[celltype_col]], levels = ct_order)

# Test AD vs Ctrl per cell type (difference in means via t-test).
if (!requireNamespace("ggpubr", quietly = TRUE)) {
  stop("Please install ggpubr: install.packages('ggpubr')")
}

ggplot(
  plot_df,
  aes(
    x = .data[[celltype_col]],
    y = .data[[score_col]],
    fill = .data$disease_group
  )
) +
  geom_boxplot(outlier.alpha = 0.25,
    # alpha = 0.4,
    position = position_dodge()
  ) +
  ggpubr::stat_compare_means(
    aes(group = .data$disease_group),
    method = "t.test",
    label = "p.signif"
  ) +
  scale_fill_manual(values = c("Ctrl" = "#56884A", "AD" = "#6B4886"), name = "Condition") +
  labs(
    title = "AD-reference PhenoMapR scores by cell type (Ctrl vs AD)",
    x = "Cell type",
    y = score_col
  ) +
  theme_minimal(base_size = 12) +
  theme(
    # axis.text.x = element_text(angle = 60, hjust = 1),
    legend.position = "right"
  ) +
  coord_cartesian(clip = "off")
```

```{r, fig.width=5, fig.height=6}
cell_types <- as.data.frame(seu@reductions$umap@cell.embeddings[, 1:2, drop = FALSE])
cell_types$cellType <- seu@meta.data$cellType 

cell_types <- cell_types %>%
  mutate(cellType = recode(cellType,
    "Oligodendrocyte Precursor Cell" = "Oligodendrocyte\nPrecursor Cell"
  ))

# edit cell colors to be more aligned
pal_cells[1]="#547F44"
pal_cells[2]="#4477AA"
pal_cells[7]="#99C399"
names(pal_cells)[7]="Oligodendrocyte\nPrecursor Cell"

 p1 <- ggscatter(cell_types, 
          x = "UMAP_1", 
          y = "UMAP_2",
          size = 1,
          color = "cellType",
          palette = pal_cells,
          legend = "right"
) + guides(color = guide_legend(override.aes = list(size = 2))) +
    labs(color = "Cell\nType") +
  theme_void() +
    # ggtitle("Sample Type Distribution") + 
    theme(plot.title = element_text(hjust = 0.5))

p2 <- ggplot(plot_df,
               aes(x = cellType, y = scale(weighted_sum_score_ad_status_z), fill = cellType, color = disease_group)) +
    geom_boxplot(outlier.alpha = 0.5, median.linewidth = 0.25, 
                 outlier.fill = NULL, 
                 outlier.color = NULL, outlier.shape = 21) +
    scale_fill_manual(
      values = pal_cells,
      name = "Cell Type"
    ) +
    ggpubr::stat_compare_means(
    aes(group = .data$disease_group),
    method = "t.test",
    label = "p.signif", show.legend = F
  ) +
  scale_color_manual(values = c("Ctrl" = "lightgrey", "AD" = "black"), name = "Source") +
    theme_pubr(base_size = 10) +
theme(
  axis.text.x  = element_blank(),
  axis.ticks.x = element_blank(),
    axis.line.x = element_blank(),
  legend.position = "right",
  strip.text = element_text(size = 12)
) +
guides(
  fill  = "none",   # removes fill legend
  color = guide_legend()  # keeps color legend
) +
    labs(x = NULL, y = "Scaled PhenoMapR score", title = "Score Distribution by Cell Type and Source") +
    theme(plot.title = element_text(hjust = 0.5))
  
      print(p2 / p1 + patchwork::plot_layout(widths = c(1, 1), guides = "collect"))
```



## Define phenotype tails and detect marker genes

```{r}
groups <- define_phenotype_groups(
  scores = scores,
  percentile = 0.05,
  score_columns = score_col
)
group_col <- grep("^phenotype_group_", names(groups), value = TRUE)[1]

meta[[group_col]] <- groups[match(meta$Cell, groups$cell_id), group_col]
table(meta[[group_col]], useNA = "ifany")
```

```{r}
expr_counts <- Seurat::GetAssayData(seu, assay = assay_use, layer = "counts")

meta <- meta %>%
  mutate(
    cellType = case_when(
      cellType == "Oligodendrocyte Precursor Cell" ~ "OPC",
      TRUE ~ cellType
    )
  )

# edit cell colors to be more aligned
pal_cells[7]="#99C399"
names(pal_cells)[7]="OPC"

group_df <- data.frame(
  cell_id = meta$Cell,
  phenotype_group = meta[[group_col]],
  cell_type = as.character(meta[[celltype_col]]),
  stringsAsFactors = FALSE
)

# Global markers
markers_global <- find_phenotype_markers(
  expression = expr_counts,
  group_labels = group_df,
  group_column = "phenotype_group",
  cell_id_column = "cell_id",
  pval_threshold = 0.05,
  max_cells_per_ident = 5000L,
  verbose = TRUE
)

# Cell-type-specific markers
markers_ct <- find_phenotype_markers(
  expression = expr_counts,
  group_labels = group_df,
  group_column = "phenotype_group",
  cell_id_column = "cell_id",
  marker_scope = "cell_type_specific",
  cell_type_column = "cell_type",
  pval_threshold = 0.05,
  max_cells_per_ident = 5000L,
  verbose = TRUE
)
```

## Marker heatmaps

```{r, fig.width=11.5, fig.height=3.2}
plot_phenotype_markers(
  markers = markers_global,
  expr_mat = expr_counts,
  meta = meta,
  cell_id_col = "Cell",
  group_col = group_col,
  score_col = score_col,
  celltype_col = celltype_col,
  celltype_palette = pal_cells,
  heatmap_type = "global",
  top_n_markers = 10L,
  n_mark_labels = 10L,
  p_adj_threshold = 0.05
)
```

```{r, fig.width=11.5, fig.height=3.2}
plot_phenotype_markers(
  markers = markers_ct,
  expr_mat = expr_counts,
  meta = meta,
  cell_id_col = "Cell",
  group_col = group_col,
  score_col = score_col,
  celltype_col = celltype_col,
  celltype_palette = pal_cells,
  heatmap_type = "cell_type_specific",
  top_n_markers = 10L,
  n_mark_labels = 5L,
  p_adj_threshold = 0.05
)
```


## Independent validation cohorts (Scissor Alzheimer’s GEO)

This section mirrors the validation idea used in Scissor’s Alzheimer’s analysis by scoring **independent GEO cohorts** with the bulk-derived reference (`ad_reference`) and testing whether PhenoMapR scores differ between **healthy vs Alzheimer’s** samples (and across **stages** for GSE28146).

**Datasets (GEO):**

- **GSE109887** — independent cohort for Healthy vs AD score separation.
- **GSE28146** — independent cohort with Alzheimer’s staging; we test score shifts across stages.

**Notes:**

- Requires **`GEOquery`** and network access.
- Runs only when this draft is running with local inputs available (`draft_data_ok`) and `ad_reference` exists.

```{r alz-validation-geo, fig.width=7.5, fig.height=3.5, message=TRUE, warning=TRUE}
validation_plot_df <- NULL
validation_tests <- list()

if (!exists("draft_data_ok") || !isTRUE(draft_data_ok) || !exists("ad_reference")) {
  message("Skipping GEO validation: draft_data_ok is FALSE or ad_reference not found.")
} else if (!requireNamespace("GEOquery", quietly = TRUE)) {
  message("Install GEOquery to run GEO validation: BiocManager::install('GEOquery').")
} else {
  score_col_bulk <- paste0("weighted_sum_score_", colnames(ad_reference)[1])
  # Validation palette (similar to Scissor Alzheimer figure): light yellow -> orange -> red
  pal_val <- c(
    Control = "#56884A",
    Incipient = "#e7d4e8",
    Moderate = "#c2a5cf",
    Severe = "#6B4886",
    AD = "#6B4886"
  )

  collapse_to_symbol <- function(eset) {
    expr <- Biobase::exprs(eset)
    probes <- rownames(expr)

    # 1) If features are Ensembl IDs, map ENSG -> SYMBOL via org.Hs.eg.db
    if (any(grepl("^ENSG[0-9]+", probes, ignore.case = TRUE)) &&
      requireNamespace("AnnotationDbi", quietly = TRUE) &&
      requireNamespace("org.Hs.eg.db", quietly = TRUE)) {
      ens_base <- sub("\\\\.[0-9]+$", "", probes)
      sym <- AnnotationDbi::mapIds(
        org.Hs.eg.db,
        keys = ens_base,
        column = "SYMBOL",
        keytype = "ENSEMBL",
        multiVals = "first"
      )
      sym <- as.character(sym)
      ok <- !is.na(sym) & nzchar(sym)
      if (any(ok)) {
        expr2 <- expr[ok, , drop = FALSE]
        sym2 <- sym[ok]
        df <- data.frame(sym = sym2, expr2, check.names = FALSE)
        agg <- stats::aggregate(. ~ sym, data = df, FUN = mean)
        rownames(agg) <- agg$sym
        return(as.matrix(agg[, -1, drop = FALSE]))
      }
    }

    # 2) If fData has a symbol column, use it directly
    fd <- tryCatch(Biobase::fData(eset), error = function(e) NULL)
    if (!is.null(fd) && nrow(fd) > 0) {
      sym_col <- NULL
      for (cand in c(
        "Symbol", "SYMBOL", "Gene Symbol", "GENE_SYMBOL", "Gene.symbol", "gene_assignment",
        "Gene.symbol.", "Gene.Symbol", "GENE", "ILMN_Gene", "ORF"
      )) {
        if (cand %in% colnames(fd)) {
          sym_col <- cand
          break
        }
      }
      if (!is.null(sym_col)) {
        sym <- as.character(fd[[sym_col]])
        ok <- !is.na(sym) & nzchar(sym)
        if (any(ok)) {
          expr2 <- expr[ok, , drop = FALSE]
          sym2 <- sym[ok]
          df <- data.frame(sym = sym2, expr2, check.names = FALSE)
          agg <- stats::aggregate(. ~ sym, data = df, FUN = mean)
          rownames(agg) <- agg$sym
          return(as.matrix(agg[, -1, drop = FALSE]))
        }
      }
    }

    # 3) Fall back: map probe IDs -> gene symbols via the platform (GPL) annotation table
    gpl_id <- tryCatch(as.character(Biobase::annotation(eset)), error = function(e) "")
    if (!nzchar(gpl_id) || !requireNamespace("GEOquery", quietly = TRUE)) return(NULL)

    gpl_obj <- suppressMessages(tryCatch(GEOquery::getGEO(gpl_id), error = function(e) NULL))
    if (is.null(gpl_obj)) return(NULL)

    gpl_tab <- tryCatch(GEOquery::Table(gpl_obj), error = function(e) NULL)
    if (is.null(gpl_tab) || nrow(gpl_tab) == 0) return(NULL)

    pick_col <- function(nms, candidates, regex = NULL) {
      for (c in candidates) if (c %in% nms) return(c)
      if (!is.null(regex)) {
        idx <- grep(regex, nms, ignore.case = TRUE)
        if (length(idx) > 0) return(nms[idx[1]])
      }
      NULL
    }

    probe_col <- pick_col(
      names(gpl_tab),
      c("ID", "ID_REF", "Probe.Set.ID", "ProbeID", "Probe ID", "SPOT_ID"),
      regex = "^id(_ref)?$|probe"
    )
    sym_col <- pick_col(
      names(gpl_tab),
      c("Gene Symbol", "GENE_SYMBOL", "Symbol", "SYMBOL", "ILMN_Gene", "Gene.symbol", "ORF"),
      regex = "gene.*symbol|^symbol$|ilmn_gene|orf"
    )
    if (is.null(probe_col) || is.null(sym_col)) return(NULL)

    map_probe <- as.character(gpl_tab[[probe_col]])
    map_sym <- as.character(gpl_tab[[sym_col]])
    okm <- !is.na(map_probe) & nzchar(map_probe) & !is.na(map_sym) & nzchar(map_sym)
    map_probe <- map_probe[okm]
    map_sym <- map_sym[okm]

    m <- match(probes, map_probe)
    sym <- map_sym[m]
    ok <- !is.na(sym) & nzchar(sym)
    if (!any(ok)) return(NULL)

    expr2 <- expr[ok, , drop = FALSE]
    sym2 <- sym[ok]
    df <- data.frame(sym = sym2, expr2, check.names = FALSE)
    agg <- stats::aggregate(. ~ sym, data = df, FUN = mean)
    rownames(agg) <- agg$sym
    as.matrix(agg[, -1, drop = FALSE])
  }

  score_bulk_mat <- function(expr_gene_by_sample, reference) {
    sc <- PhenoMap(
      expression = expr_gene_by_sample,
      reference = reference,
      z_score_cutoff = 2,
      verbose = FALSE
    )
    sc[, score_col_bulk, drop = FALSE]
  }

  ## --- GSE109887: Healthy vs AD --------------------------------------------
  tryCatch(
    {
      g <- suppressMessages(GEOquery::getGEO("GSE109887", GSEMatrix = TRUE, AnnotGPL = TRUE)[[1]])
      pd <- Biobase::pData(g)
      meta_text <- tolower(apply(pd, 1, function(x) paste(x, collapse = " | ")))
      grp <- ifelse(
        grepl("alzheimer|ad\\b|dementia", meta_text), "AD",
        ifelse(grepl("control|healthy|normal", meta_text), "Control", NA_character_)
      )
      expr_sym <- collapse_to_symbol(g)
      if (is.null(expr_sym)) stop("GSE109887: could not find a gene-symbol column in fData().")

      keep <- !is.na(grp)
      if (sum(keep) < 4) stop("GSE109887: too few labeled samples after parsing metadata.")
      expr_sym <- expr_sym[, keep, drop = FALSE]
      grp <- factor(grp[keep], levels = c("Control", "AD"))

      sc <- score_bulk_mat(expr_sym, ad_reference)
      x <- sc[, 1, drop = TRUE]

      validation_plot_df <- rbind(
        validation_plot_df,
        data.frame(
          cohort = "GSE109887",
          group = grp,
          score = x,
          stringsAsFactors = FALSE
        )
      )
      ok_w <- stats::complete.cases(x, grp)
      if (sum(ok_w) > 0 && length(unique(grp[ok_w])) >= 2) {
        validation_tests[["GSE109887"]] <- stats::wilcox.test(x[ok_w] ~ grp[ok_w])
      }
    },
    error = function(e) message("GSE109887 failed: ", conditionMessage(e))
  )

  ## --- GSE28146: stage / severity -----------------------------------------
  tryCatch(
    {
      g <- suppressMessages(GEOquery::getGEO("GSE28146", GSEMatrix = TRUE, AnnotGPL = TRUE)[[1]])
      pd <- Biobase::pData(g)
      meta_text <- tolower(apply(pd, 1, function(x) paste(x, collapse = " | ")))

      # Attempt to recover an ordered stage variable from metadata text.
      stage <- ifelse(grepl("control|healthy|normal", meta_text), "Control", NA_character_)
      stage[grepl("incipient", meta_text)] <- "Incipient"
      stage[grepl("moderate", meta_text)] <- "Moderate"
      stage[grepl("severe", meta_text)] <- "Severe"

      expr_sym <- collapse_to_symbol(g)
      if (is.null(expr_sym)) stop("GSE28146: could not find a gene-symbol column in fData().")

      keep <- !is.na(stage)
      if (sum(keep) < 4) stop("GSE28146: too few labeled samples after parsing metadata.")
      expr_sym <- expr_sym[, keep, drop = FALSE]
      stage <- factor(stage[keep], levels = c("Control", "Incipient", "Moderate", "Severe"), ordered = TRUE)

      sc <- score_bulk_mat(expr_sym, ad_reference)
      x <- sc[, 1, drop = TRUE]

      validation_plot_df <- rbind(
        validation_plot_df,
        data.frame(
          cohort = "GSE28146",
          group = stage,
          score = x,
          stringsAsFactors = FALSE
        )
      )
      ok_w <- stats::complete.cases(x, stage)
      if (sum(ok_w) > 0 && length(unique(stage[ok_w])) >= 2) {
        validation_tests[["GSE28146"]] <- stats::kruskal.test(x[ok_w] ~ stage[ok_w])
      }
    },
    error = function(e) message("GSE28146 failed: ", conditionMessage(e))
  )

  if (!is.null(validation_plot_df) && nrow(validation_plot_df) > 0) {
    # Scale within cohort to make y-axis comparable across platforms.
    validation_plot_df$score_scaled <- ave(
      validation_plot_df$score,
      validation_plot_df$cohort,
      FUN = function(x) as.numeric(scale(x))
    )

    if (!requireNamespace("ggpubr", quietly = TRUE)) {
      stop("Please install ggpubr to annotate validation p-values: install.packages('ggpubr')")
    }

    validation_plot_df$cohort <- factor(
      validation_plot_df$cohort,
      levels = c("GSE109887", "GSE28146")
    )
    validation_plot_df$group <- factor(
      as.character(validation_plot_df$group),
      levels = c("Control", "AD", "Incipient", "Moderate", "Severe")
    )

    p_val <- ggplot(validation_plot_df, aes(x = group, y = score_scaled, fill = group)) +
      geom_violin(trim = FALSE, alpha = 0.35, color = NA) +
      geom_boxplot(width = 0.18, outlier.alpha = 0.0, alpha = 0.65) +
      geom_jitter(width = 0.08, height = 0, size = 0.9, alpha = 0.85) +
      facet_wrap(~cohort, scales = "free_x", nrow = 1L) +
      ggpubr::stat_compare_means(
        aes(group = validation_plot_df$group),
        method = "wilcox.test",
        # ref.group = "Control",
        comparisons = list(c("Control", "AD")),
        label = "p.signif",
        label.y.npc = "top",
        vjust = 0.5
      ) +
            ggpubr::stat_compare_means(
        aes(group = validation_plot_df$group),
        method = "wilcox.test",
        # ref.group = "Control",
        comparisons = list(c("Control", "Incipient"), c("Control", "Moderate", "Severe"), c("Control", "Severe")),
        label = "p.signif",
        label.y.npc = "top",
        vjust = 0.5
      ) +
      scale_fill_manual(
        values = pal_val[c("Control", "AD", "Incipient", "Moderate", "Severe")],
        drop = FALSE
      ) +
      labs(
        title = "Independent GEO validation",
        x = NULL,
        y = "Scaled PhenoMap score (within cohort)"
      ) +
      theme_bw() +
      theme(legend.position = "none")

    print(p_val)
  } else {
    message("No validation results to plot (check GEO access and phenotype parsing).")
  }
}
```

