---
title: "Applying PhenoMapR to facioscapulohumeral muscular dystrophy (FSHD) datasets"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_collapsed: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  out.width = "100%"
  # cache = TRUE,
  # cache.path = "cache/fshd-vignette/"
)

# Full vignette (bulk download, sc, validation GEO). Set PHENOMAPR_VIGNETTE_RUN=false to skip
# data-heavy chunks in custom builds (not used by default pkgdown).
vignette_run <- tolower(Sys.getenv("PHENOMAPR_VIGNETTE_RUN", "true")) %in% c("1", "true", "yes")

# Seurat's FindNeighbors / FindClusters use future::future_lapply. A multisession
# or multicore plan (e.g. from .Rprofile) can trigger FutureLaunchError when
# workers see a different 'future' version than the parent, or when unloading
# 'future' fails because Seurat imports it. Sequential plan avoids forked workers.
if (requireNamespace("future", quietly = TRUE)) {
  future::plan("sequential", gc = FALSE)
}
```

## Overview

This vignette shows how to use **`PhenoMapR`** to mirror the **FSHD (facioscapulohumeral muscular dystrophy)** analysis in **Scissor** ([Sun *et al.*, *Nat. Biotechnol.* 2022](https://doi.org/10.1038/s41587-021-01091-3); [Nature article](https://www.nature.com/articles/s41587-021-01091-3)), where bulk expression and a binary phenotype are integrated with single-cell data to identify phenotype-associated cells. **Figure 5** of that paper applies Scissor to FSHD using public GEO data.

We follow the paper’s description of the GEO inputs:

- **FSHD scRNA-seq ([s](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE122873))** — 7,047 cells from muscle biopsy–derived myocyte cultures (four FSHD patients, two controls); **148 cells** with high mitochondrial expression (≥10% of total) are removed, leaving **6,899** cells for analysis (Scissor’s reported numbers).
- **FSHD bulk ([GSE140261](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE140261))** — TPM from **27** follow-up FSHD biopsies and **eight** control biopsies in the processed **year 2** supplementary files; Ensembl IDs are collapsed to **gene symbols** by **summing** expression when multiple Ensembl IDs map to the same symbol (here we use the **mean TPM** across duplicate Ensembl IDs after mapping, which is equivalent for proportional summaries).

**PhenoMapR** workflow:

1. Derive a **custom z-score reference** from bulk TPM with **`derive_reference_from_bulk()`** (FSHD vs. control).
2. Score all single cells with **`PhenoMap()`** and visualize on **UMAP**.
3. Define **adverse / favorable** tails with **`define_phenotype_groups()`** and run **`find_phenotype_markers()`** (`marker_scope = "phenotype_groups"` for global markers).
4. Draw a **global marker heatmap** with **`plot_phenotype_markers(..., heatmap_type = "global")`**.
5. **Validation** — apply **`ref_fshd`** to three independent GEO cohorts (**GSE115650**, **GSE56787** muscle biopsies only, **GSE26852** FSHD vs. normal controls), each processed separately; **FSHD vs. control** stratification with **`PhenoMap()`**, **violin plots**, **sample counts**, and **Wilcoxon** *p*-values (Section 6).

## Packages

```{r libraries}
suppressPackageStartupMessages({
  library(googledrive)
  library(PhenoMapR)
  library(ggplot2)
  library(dplyr)
  library(Seurat)
  library(AnnotationDbi)
  library(org.Hs.eg.db)
  library(patchwork)
  library(ggpubr)
  library(ComplexHeatmap)
  library(circlize)
})
```

## 1. Bulk TPM and phenotype (GSE140261)

Processed **year 2** supplementary files ([GEO FTP](https://ftp.ncbi.nlm.nih.gov/geo/series/GSE140nnn/GSE140261/suppl/)):

- `GSE140261_year2_TPM.csv.gz` — TPM with **Ensembl IDs** (with version) as rows.
- `GSE140261_year2_column_data.csv.gz` — metadata; we use **`pheno_type`** (`FSHD` vs `Control`).

```{r bulk-download}
# ftp_base <- "https://ftp.ncbi.nlm.nih.gov/geo/series/GSE140nnn/GSE140261/suppl/"
# tpm_url <- paste0(ftp_base, "GSE140261_year2_TPM.csv.gz")
# meta_url <- paste0(ftp_base, "GSE140261_year2_column_data.csv.gz")
# 
# #' Try several strategies: NCBI HTTPS often fails with default download.file (proxy/SSL/RStudio).
# #' Falls back to GEOquery supplementary download, then system curl if available.
# download_geo_file <- function(url, dest, geo_acc = NULL, supp_filename = NULL) {
#   if (file.exists(dest) && file.info(dest)$size > 1000L) {
#     return(normalizePath(dest, winslash = "/"))
#   }
#   dir.create(dirname(dest), showWarnings = FALSE, recursive = TRUE)
# 
#   # 1) curl package: most reliable for HTTPS (often works when download.file / RStudio fails)
#   if (requireNamespace("curl", quietly = TRUE)) {
#     st <- tryCatch(
#       {
#         curl::curl_download(url, dest, quiet = TRUE)
#         0L
#       },
#       error = function(e) 1L
#     )
#     if (st == 0L && file.exists(dest) && file.info(dest)$size > 1000L) {
#       return(normalizePath(dest, winslash = "/"))
#     }
#   }
# 
#   try_download_file <- function() {
#     methods <- c("libcurl", "wininet", "auto")
#     for (m in methods) {
#       if (m == "wininet" && .Platform$OS.type != "windows") next
#       st <- tryCatch(
#         download.file(url, dest, mode = "wb", quiet = TRUE, method = m),
#         error = function(e) 1L
#       )
#       if (isTRUE(st == 0L) && file.exists(dest) && file.info(dest)$size > 1000L) {
#         return(TRUE)
#       }
#     }
#     FALSE
#   }
# 
#   if (try_download_file()) {
#     return(normalizePath(dest, winslash = "/"))
#   }
# 
#   if (!is.null(geo_acc) && !is.null(supp_filename) && requireNamespace("GEOquery", quietly = TRUE)) {
#     base <- tempfile("geo_supp")
#     dir.create(base)
#     tryCatch(
#       suppressMessages(
#         GEOquery::getGEOSuppFiles(geo_acc, makeDirectory = TRUE, baseDir = base)
#       ),
#       error = function(e) invisible(NULL)
#     )
#     hits <- list.files(
#       base,
#       pattern = supp_filename,
#       fixed = TRUE,
#       recursive = TRUE,
#       full.names = TRUE
#     )
#     if (length(hits) >= 1L && file.copy(hits[1], dest, overwrite = TRUE)) {
#       if (file.exists(dest) && file.info(dest)$size > 1000L) {
#         return(normalizePath(dest, winslash = "/"))
#       }
#     }
#   }
# 
#   curl_bin <- Sys.which("curl")
#   if (nzchar(curl_bin)) {
#     st <- system2(curl_bin, c("-fsSL", "-o", dest, shQuote(url)), stdout = FALSE, stderr = FALSE)
#     if (st == 0L && file.exists(dest) && file.info(dest)$size > 1000L) {
#       return(normalizePath(dest, winslash = "/"))
#     }
#   }
# 
#   stop(
#     "Download failed for:\n  ", url,
#     "\nTry: install.packages('curl'); BiocManager::install('GEOquery'); ",
#     "options(download.file.method = 'libcurl'); ",
#     "or download the file in a browser and set PHENOMAPR_GSE122873_TAR to the local path."
#   )
# }
# 
# download_if_needed <- function(url, dest, geo_acc = NULL, supp_filename = basename(url)) {
#   download_geo_file(url, dest, geo_acc = geo_acc, supp_filename = supp_filename)
# }
# 
# #' Collapse Ensembl rows to HGNC symbols by mean TPM (duplicate Ensembl → one symbol).
# collapse_ensembl_tpm_to_symbol <- function(tpm_ens_rows_samples_cols) {
#   ens <- rownames(tpm_ens_rows_samples_cols)
#   ens_base <- sub("\\.[0-9]+$", "", ens)
#   sym <- AnnotationDbi::mapIds(
#     org.Hs.eg.db,
#     keys = ens_base,
#     column = "SYMBOL",
#     keytype = "ENSEMBL",
#     multiVals = "first"
#   )
#   df <- as.data.frame(tpm_ens_rows_samples_cols)
#   df$gene_symbol <- sym
#   df <- df[!is.na(df$gene_symbol) & nzchar(df$gene_symbol), , drop = FALSE]
#   sample_cols <- setdiff(names(df), "gene_symbol")
#   agg <- stats::aggregate(
#     df[, sample_cols, drop = FALSE],
#     by = list(gene_symbol = df$gene_symbol),
#     FUN = mean,
#     na.rm = TRUE
#   )
#   mat <- as.matrix(agg[, -1, drop = FALSE])
#   rownames(mat) <- agg$gene_symbol
#   mat
# }
# 
# 
#   td <- tempdir()
#   tpm_path <- download_if_needed(
#     tpm_url,
#     file.path(td, "GSE140261_year2_TPM.csv.gz"),
#     geo_acc = "GSE140261",
#     supp_filename = "GSE140261_year2_TPM.csv.gz"
#   )
#   meta_path <- download_if_needed(
#     meta_url,
#     file.path(td, "GSE140261_year2_column_data.csv.gz"),
#     geo_acc = "GSE140261",
#     supp_filename = "GSE140261_year2_column_data.csv.gz"
#   )
# 
#   tpm_wide <- read.csv(gzfile(tpm_path), row.names = 1, check.names = FALSE, stringsAsFactors = FALSE)
#   tpm_sym <- collapse_ensembl_tpm_to_symbol(tpm_wide)
# 
#   coldata <- read.csv(gzfile(meta_path), stringsAsFactors = FALSE, check.names = FALSE)
#   coldata <- coldata[tolower(coldata$pheno_type) %in% c("fshd", "control"), , drop = FALSE]
#   coldata$group <- ifelse(tolower(coldata$pheno_type) == "fshd", "FSHD", "Control")
#   coldata$sample_id <- coldata$sample_name
# 
#   use_samples <- intersect(colnames(tpm_sym), coldata$sample_id)
#   tpm_sym <- tpm_sym[, use_samples, drop = FALSE]
#   pheno_bulk <- coldata[match(use_samples, coldata$sample_id), c("sample_id", "group")]
#   pheno_bulk$group <- factor(pheno_bulk$group, levels = c("FSHD", "Control"))
# 
  # saveRDS(tpm_sym, "~/Desktop/scIMPEL/vignettes/FSHD_GSE140261_bulk_expression.rds")
    # saveRDS(pheno_bulk, "~/Desktop/scIMPEL/vignettes/FSHD_GSE140261_phenotype.rds")
# 
# https://drive.google.com/file/d/14g_uHDVbjDA2Tl6-ttY7yF76s8Sy7CuH/view?usp=drive_link
# https://drive.google.com/file/d/13jTizdcmWPYMlEnjEIZ-V8U1wQzch2o1/view?usp=drive_link

if (!vignette_run) {
  message("Skipping bulk reference build (PHENOMAPR_VIGNETTE_RUN is false).")
} else {
  options(googledrive_quiet = TRUE)
  googledrive::drive_deauth()
  googledrive::drive_download(
    googledrive::as_id("13jTizdcmWPYMlEnjEIZ-V8U1wQzch2o1"),
    "FSHD_GSE140261_phenotype.rds",
    overwrite = TRUE
  )
  googledrive::drive_download(
    googledrive::as_id("14g_uHDVbjDA2Tl6-ttY7yF76s8Sy7CuH"),
    "FSHD_GSE140261_bulk_expression.rds",
    overwrite = TRUE
  )

  pheno_bulk <- readRDS("FSHD_GSE140261_phenotype.rds")
  tpm_sym <- readRDS("FSHD_GSE140261_bulk_expression.rds")

  ref_fshd <- derive_reference_from_bulk(
    bulk_expression = tpm_sym,
    phenotype = pheno_bulk,
    sample_id_column = "sample_id",
    phenotype_column = "group",
    phenotype_type = "binary",
    normalize = TRUE,
    binary_positive_reference = "first",
    verbose = TRUE
  )
}
```

With **`levels = c("FSHD", "Control")`** and **`binary_positive_reference = "first"`**, genes with **positive** z-scores are **higher in FSHD** bulk samples.

```{r bulk-ref-plot, fig.width=6, fig.height=3.5}
  plot_score_distribution(ref_fshd, main = "GSE140261: FSHD vs control (gene-level z-scores)")
```

## 2. Single-cell data (GSE122873) and mitochondrial QC

[GSE122873](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE122873) provides **`GSE122873_RAW.tar`** with six **`GSM*_*.txt.gz`** tables (myocyte cultures: four FSHD + two control). Each table is **genes × cells** with a header row of cell barcodes ([Kendall *et al.*, *Genome Res.* 2019](https://doi.org/10.1101/gr.243734.118); [PMC6423425](https://pmc.ncbi.nlm.nih.gov/articles/PMC6423425/)). **Gene identifiers are Ensembl IDs** (with optional version suffix); **`PhenoMapR`** matches the custom reference by **HUGO symbols**, so we **map Ensembl → symbol** and **sum counts** per symbol **before** building the Seurat object.


```{r sc-download}
# sc_tar_url <- "https://ftp.ncbi.nlm.nih.gov/geo/series/GSE122nnn/GSE122873/suppl/GSE122873_RAW.tar"
# 
# read_gse122873_sample <- function(path) {
#   m <- as.matrix(read.delim(gzfile(path), row.names = 1, check.names = FALSE))
#   storage.mode(m) <- "double"
#   m
# }
# 
# merge_gse122873 <- function(dir_untar) {
#   fs <- list.files(dir_untar, pattern = "^GSM.*\\.txt\\.gz$", full.names = TRUE)
#   if (length(fs) < 1L) stop("No GSM*.txt.gz after extracting GSE122873_RAW.tar")
#   mats <- lapply(fs, read_gse122873_sample)
#   genes <- Reduce(intersect, lapply(mats, rownames))
#   combined <- do.call(cbind, lapply(mats, function(m) m[genes, , drop = FALSE]))
#   colnames(combined) <- make.unique(colnames(combined), sep = "_")
#   sn <- basename(fs)
#   lab <- sub("\\.txt\\.gz$", "", sn)
#   lab <- sub("^GSM[0-9]+_", "", lab)
#   rep_lab <- rep(lab, times = vapply(mats, ncol, 1L))
#   list(counts = combined, donor_batch = rep_lab)
# }
# 
# #' Genes x cells count matrix: Ensembl rownames -> HUGO symbols (sum counts per symbol).
# #' Rows already looking like symbols (not ENSG...) are kept as-is.
# collapse_counts_ensembl_to_hugo <- function(counts_mat) {
#   if (!requireNamespace("org.Hs.eg.db", quietly = TRUE)) {
#     stop("Package 'org.Hs.eg.db' is required to map Ensembl IDs to symbols.")
#   }
#   x <- as.matrix(counts_mat)
#   genes <- rownames(x)
#   ens_base <- sub("\\.[0-9]+$", "", genes)
#   is_ens <- grepl("^ENSG[0-9]+", ens_base, ignore.case = TRUE)
#   sym <- genes
#   if (any(is_ens)) {
#     mapped <- AnnotationDbi::mapIds(
#       org.Hs.eg.db,
#       keys = ens_base[is_ens],
#       column = "SYMBOL",
#       keytype = "ENSEMBL",
#       multiVals = "first"
#     )
#     sym[is_ens] <- as.character(mapped)
#   }
#   ok <- !is.na(sym) & nzchar(sym) & sym != "NA"
#   x <- x[ok, , drop = FALSE]
#   sym <- sym[ok]
#   if (length(unique(sym)) < length(sym)) {
#     x <- rowsum(x, group = sym, reorder = FALSE)
#   } else {
#     rownames(x) <- sym
#   }
#   x
# }
# 
#   td_sc <- tempfile("gse122873")
#   dir.create(td_sc, showWarnings = FALSE)
#   tar_path <- file.path(td_sc, "GSE122873_RAW.tar")
#   env_tar <- Sys.getenv("PHENOMAPR_GSE122873_TAR", unset = "")
#   if (nzchar(env_tar) && file.exists(env_tar)) {
#     file.copy(env_tar, tar_path, overwrite = TRUE)
#   } else {
#     download_geo_file(
#       sc_tar_url,
#       tar_path,
#       geo_acc = "GSE122873",
#       supp_filename = "GSE122873_RAW.tar"
#     )
#   }
#   untar(tar_path, exdir = td_sc)
#   sc_list <- merge_gse122873(td_sc)
#   counts <- collapse_counts_ensembl_to_hugo(sc_list$counts)
#   meta_cells <- data.frame(
#     cell_id = colnames(counts),
#     donor_batch = sc_list$donor_batch,
#     stringsAsFactors = FALSE
#   )
#   meta_cells$disease <- ifelse(grepl("^CTRL", meta_cells$donor_batch), "Control", "FSHD")
#   rownames(meta_cells) <- meta_cells$cell_id
# 
#   seu <- CreateSeuratObject(counts = counts, meta.data = meta_cells)
#   seu[["percent.mt"]] <- PercentageFeatureSet(seu, pattern = "^MT-")
#   seu <- subset(seu, subset = percent.mt < 10)
#   seu <- NormalizeData(seu, verbose = FALSE)
#   seu <- FindVariableFeatures(seu, verbose = FALSE)
#   all.genes <- rownames(seu)
# seu <- ScaleData(seu, features = all.genes)
#   seu <- RunPCA(seu, verbose = FALSE)
# 
# # Cluster the cells and run UMAP #####
# seu <- FindNeighbors(seu, dims = 1:10)
# seu <- FindClusters(seu, resolution = 0.5)
#   seu <- RunUMAP(seu, dims = 1:30, verbose = FALSE)
#   seu.markers <- FindAllMarkers(object = seu, only.pos = TRUE)
#   
#   saveRDS(seu, "~/Desktop/scIMPEL/vignettes/FSHD_GSE122873_seurat.rds")
  
```

After filtering, the resulting Seurat object contains **6,898** cells (Scissor paper reports 6,899).

## 3. Score single cells with the custom reference

```{r score-sc}
googledrive::drive_download(
  googledrive::as_id("1pi4EWw9dYof-6ezoICMYO-SQeGRsaxRV"),
  "FSHD_GSE122873_seurat.rds",
  overwrite = TRUE
)

seu <- readRDS("FSHD_GSE122873_seurat.rds")

scores_sc <- PhenoMap(
  seu,
  reference = ref_fshd,
  assay = "RNA",
  slot = "data",
  z_score_cutoff = 2,
  verbose = FALSE
)
seu <- add_scores_to_seurat(seu, scores_sc)
score_col <- grep("weighted_sum_score", colnames(scores_sc), value = TRUE)[1]
```

```{r umap-score, fig.width=12, fig.height=4.5}
  pal <- c(Control = "#FED789FF", FSHD = "#476F84FF")

plot_data_sc <- data.frame(
  disease = seu$disease,
  sample = seu$donor_batch,
  score   = seu@meta.data[[score_col]],
  score_scaled = scale(seu@meta.data[[score_col]]),
  umap1   = seu@reductions$umap@cell.embeddings[, 1],
  umap2   = seu@reductions$umap@cell.embeddings[, 2],
  stringsAsFactors = FALSE
) %>%
  mutate(sample = reorder(sample, score_scaled, FUN = mean))

  
  p1 <- ggplot(plot_data_sc, aes(x = sample, y = score_scaled, fill = disease)) +
  geom_violin(alpha = 0.6) +
  geom_boxplot(width = 0.15, outlier.size = 0.4) +
  scale_fill_manual(values = pal) +
  stat_compare_means(
    method = "wilcox.test",       # or "t.test"
    label  = "p.signif",          # "p.format" for exact p-value
    comparisons = list(c("FSHD", "Control"))  # your two group names
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.5)) +
  labs(
    title = "Custom FSHD bulk signature (single-cell scores)",
    y = "Scaled PhenoMapR score",
    x = NULL
  ) 
  
  p2 <- ggscatter(plot_data_sc, 
          x = "umap1", 
          y = "umap2",
          size = 1,
          color = "disease",
          palette = pal,
          legend = "right"
) + guides(color = guide_legend(override.aes = list(size = 2))) +
    labs(color = "Sample\nType") +
  theme_void() +
    ggtitle("Sample Type Distribution") + 
    theme(plot.title = element_text(hjust = 0.5))
  
  p3 <- ggscatter(plot_data_sc, 
          x = "umap1", 
          y = "umap2",
          size = 1,
          color = "score_scaled",
          legend = "right"
) + 
  guides(color = guide_colorbar(override.aes = list(size = 2))) +
  scale_color_gradient2(
    low      = "#2166AC",
    mid      = "#F7F7F7",
    high     = "#B2182B",
    midpoint = 0
  ) +
  labs(color = "PhenoMapR\nz-score") +
  theme_void() +
    ggtitle("PhenoMap score Distribution") 

    print(p1 / (p2 + p3) + patchwork::plot_layout(widths = c(1, 1, 1)))

```

## Cell type annotations (CyteTypeR results)

This dataset does not provide cell type labels. Here we use pre-computed **CyteTypeR** annotations (saved alongside this vignette) and attach them to the Seurat object metadata as `cell_type` and `coarse_cell_type`. The next figure shows those labels on the **UMAP** (`label = TRUE`).

```{r cytetype-annotations}
  ann_base <- "cytetype_annotations_ce748d9d-6191-42bc-a212-4bace7e97e4b.csv"

  find_cytetype_csv <- function(fn) {
    try_one <- function(p) {
      if (is.na(p) || !nzchar(p)) return(NA_character_)
      p <- path.expand(p)
      if (file.exists(p)) return(normalizePath(p, winslash = "/", mustWork = TRUE))
      NA_character_
    }

    # 1) same directory as this vignette Rmd (works when knitr sets full path)
    fnm <- tryCatch(knitr::opts_knit$get("filename"), error = function(e) NULL)
    if (!is.null(fnm) && nzchar(fnm)) {
      hit <- try_one(file.path(dirname(fnm), fn))
      if (!is.na(hit)) return(hit)
    }

    # 2) current_input() basename-only breaks dirname(); use dirname(fnm) only above
    in_path <- tryCatch(knitr::current_input(), error = function(e) NA_character_)
    if (!is.na(in_path) && nzchar(in_path) && basename(in_path) != in_path) {
      hit <- try_one(file.path(dirname(in_path), fn))
      if (!is.na(hit)) return(hit)
    }

    # 3) repo root / cwd layouts
    for (p in c(file.path("vignettes", fn), fn)) {
      hit <- try_one(p)
      if (!is.na(hit)) return(hit)
    }

    # 4) walk up from cwd to find package/vignettes/<fn> or inst/vignette-drafts/<fn>
    d <- normalizePath(getwd(), winslash = "/")
    for (i in 1:20) {
      for (sub in c("vignettes", file.path("inst", "vignette-drafts"))) {
        hit <- try_one(file.path(d, sub, fn))
        if (!is.na(hit)) return(hit)
      }
      nd <- dirname(d)
      if (identical(nd, d)) break
      d <- nd
    }

    # 5) installed package (draft vignettes ship under inst/vignette-drafts/)
    if (requireNamespace("PhenoMapR", quietly = TRUE)) {
      hit <- try_one(system.file("vignette-drafts", fn, package = "PhenoMapR"))
      if (!is.na(hit)) return(hit)
      hit <- try_one(system.file("vignettes", fn, package = "PhenoMapR"))
      if (!is.na(hit)) return(hit)
    }

    NA_character_
  }

  ann_path <- find_cytetype_csv(ann_base)
  if (is.na(ann_path) || !nzchar(ann_path)) {
    stop(
      "Missing CyteTypeR annotations CSV (", ann_base, "). ",
      "Place it next to this Rmd or under inst/vignette-drafts/ in the package source."
    )
  }

  ann <- utils::read.csv(ann_path, stringsAsFactors = FALSE, check.names = FALSE)
  req_cols <- c("cluster_id", "cell_type", "coarse_cell_type")
  if (!all(req_cols %in% names(ann))) {
    stop("CyteTypeR annotations missing required columns: ", paste(setdiff(req_cols, names(ann)), collapse = ", "))
  }

  # Map Seurat clusters to CyteTypeR cluster_id.
  # Convention used when these results were generated: cluster_id = as.integer(seurat_clusters) + 1
  cl <- as.character(seu$seurat_clusters)
  if (anyNA(cl)) stop("Missing seurat_clusters in Seurat metadata.")
  if (!all(grepl("^[0-9]+$", cl))) {
    stop("Expected numeric seurat_clusters for CyteTypeR mapping; got non-numeric cluster labels.")
  }
  seu$cluster_id <- as.integer(cl) + 1L

  # Join per-cell annotations
  ann_sub <- ann[, req_cols, drop = FALSE]
  key <- match(seu$cluster_id, ann_sub$cluster_id)
  seu$cell_type <- ann_sub$cell_type[key]
  seu$coarse_cell_type <- ann_sub$coarse_cell_type[key]

  # Basic sanity check
  if (all(is.na(seu$cell_type))) {
    stop("CyteTypeR mapping produced all-NA cell_type; check cluster_id alignment.")
  }
```

```{r cytetype-umap, fig.width=9, fig.height=5}
  ct_levels <- sort(unique(as.character(stats::na.omit(seu$cell_type))))
  pal_ct <- PhenoMapR::get_celltype_palette(ct_levels)
  
  plot_data_sc <- data.frame(
    cell_type = seu$cell_type,
    coarse_cell_type = seu$coarse_cell_type,
    score   = seu@meta.data[[score_col]],
    score_scaled = scale(seu@meta.data[[score_col]]),
    umap1   = seu@reductions$umap@cell.embeddings[, 1],
    umap2   = seu@reductions$umap@cell.embeddings[, 2],
    stringsAsFactors = FALSE
  ) %>%
    mutate(cell_type = reorder(cell_type, score_scaled, FUN = median))
  
 p1 <- ggscatter(plot_data_sc, 
          x = "umap1", 
          y = "umap2",
          size = 1,
          color = "coarse_cell_type",
                    palette = c("Vascular Smooth Muscle Cells" = "#B4849F", 
                      "Stromal Cells" = "#EE8866", 
                      "Platelets" = "#99C399"),
          legend = "right"
) + guides(color = guide_legend(override.aes = list(size = 2))) +
    labs(color = "Cell Type (Broad)") +
  theme_void() +
    theme(plot.title = element_text(hjust = 0.5))
  
  p2 <- ggplot(plot_data_sc,
               aes(x = coarse_cell_type, y = score_scaled, fill = coarse_cell_type)) +
    geom_boxplot(outlier.alpha = 0.2) +
    scale_fill_manual(
      values = c(
        "Vascular Smooth Muscle Cells" = "#B4849F",
        "Stromal Cells" = "#EE8866",
        "Platelets" = "#99C399"
      ),
      name = "Cell Type (Broad)"
    ) +
    theme_pubr(base_size = 10) +
theme(
  axis.text.x  = element_blank(),
  axis.ticks.x = element_blank(),
    axis.line.x = element_blank(),
  legend.position = "none",
  strip.text = element_text(size = 12)
) +
    labs(x = NULL, y = "Scaled PhenoMapR score", title = "Score distribution by broad cell type") +
    theme(plot.title = element_text(hjust = 0.5))

      print(p2 / p1 + patchwork::plot_layout(widths = c(1, 1), guides = "collect"))

```

## Plot by more refined cell type
```{r}
  p1 <- ggscatter(plot_data_sc, 
          x = "umap1", 
          y = "umap2",
          size = 1,
          color = "cell_type",
          palette = pal_ct,
          legend = "right"
) + guides(color = guide_legend(override.aes = list(size = 2))) +
    labs(color = "Cell Type") +
  theme_void() +
    theme(plot.title = element_text(hjust = 0.5)) 

p2 <- ggplot(plot_data_sc,
               aes(x = cell_type, y = score_scaled, fill = cell_type)) +
    geom_boxplot(outlier.alpha = 0.2) +
    scale_fill_manual(
      values = pal_ct,
      name = "Cell Type"
    ) +
    theme_pubr(base_size = 10) +
theme(
  axis.text.x  = element_blank(),
  axis.ticks.x = element_blank(),
    axis.line.x = element_blank(),
  legend.position = "none",
  strip.text = element_text(size = 12)
) +
    labs(x = NULL, y = "Scaled PhenoMapR score", title = "Score distribution by cell type") +
    theme(plot.title = element_text(hjust = 0.5))
  
  
      print(p2 / p1 + patchwork::plot_layout(widths = c(1, 1), guides = "collect"))
```


## 4. Phenotype tails and marker genes (single-cell)

We label **Most Adverse** / **Most Favorable** / **Other** tails from the custom score (**higher** = more FSHD-like under this reference), then run:

- Global markers (`marker_scope = "phenotype_groups"`)
- Cell-type-specific markers (`marker_scope = "cell_type_specific"`) using **CyteTypeR** `cell_type`
- Coarse cell-type-specific markers using **CyteTypeR** `coarse_cell_type`

```{r markers-sc}
  groups <- define_phenotype_groups(scores_sc, percentile = 0.05, score_columns = score_col)
  group_col <- grep("^phenotype_group_", names(groups), value = TRUE)[1]

  mk <- find_phenotype_markers(
    seu,
    group_labels = groups,
    group_column = group_col,
    cell_id_column = "cell_id",
    marker_scope = "phenotype_groups",
    assay = "RNA",
    slot = "data",
    verbose = FALSE
  )
  message("Adverse markers (top 8):")
  print(utils::head(mk$adverse_markers[order(mk$adverse_markers$p_adj), ], 8))
  message("Favorable markers (top 8):")
  print(utils::head(mk$favorable_markers[order(mk$favorable_markers$p_adj), ], 8))

  # Add CyteTypeR labels into group_labels for cell-type-specific contrasts
  groups_ct <- as.data.frame(groups)
  groups_ct$cell_type <- seu$cell_type[match(groups_ct$cell_id, colnames(seu))]
  groups_ct$coarse_cell_type <- seu$coarse_cell_type[match(groups_ct$cell_id, colnames(seu))]

  mk_celltype <- tryCatch(
    find_phenotype_markers(
      seu,
      group_labels = groups_ct,
      group_column = group_col,
      cell_id_column = "cell_id",
      marker_scope = "cell_type_specific",
      cell_type_column = "cell_type",
      assay = "RNA",
      slot = "data",
      verbose = FALSE
    ),
    error = function(e) {
      message("cell_type_specific markers (cell_type) failed: ", conditionMessage(e))
      NULL
    }
  )

  mk_coarse <- tryCatch(
    find_phenotype_markers(
      seu,
      group_labels = groups_ct,
      group_column = group_col,
      cell_id_column = "cell_id",
      marker_scope = "cell_type_specific",
      cell_type_column = "coarse_cell_type",
      assay = "RNA",
      slot = "data",
      verbose = FALSE
    ),
    error = function(e) {
      message("cell_type_specific markers (coarse_cell_type) failed: ", conditionMessage(e))
      NULL
    }
  )
```

## 5. Phenotype marker heatmap (`plot_phenotype_markers`)

After **`find_phenotype_markers()`**, **`plot_phenotype_markers()`** builds a **ComplexHeatmap** of marker expression with column annotations for **phenotype group**, **cell type**, and the continuous **PhenoMapR score**.

```{r plot-phenotype-markers-fshd, fig.align='center', fig.width=13, fig.height=4.5}
if (vignette_run && exists("mk") && exists("groups") && exists("group_col")) {
  if (!requireNamespace("ComplexHeatmap", quietly = TRUE) || !requireNamespace("circlize", quietly = TRUE)) {
    message("Install packages 'ComplexHeatmap' and 'circlize' to draw plot_phenotype_markers().")
  } else {
    expr_mat <- tryCatch(
      Seurat::GetAssayData(seu, layer = "data", assay = "RNA"),
      error = function(e) Seurat::GetAssayData(seu, slot = "data", assay = "RNA")
    )
    meta_hm <- as.data.frame(seu@meta.data)
    meta_hm$cell_id <- colnames(seu)
    meta_hm[[group_col]] <- groups[[group_col]][match(meta_hm$cell_id, groups$cell_id)]
    pal_ct <- PhenoMapR::get_celltype_palette(levels(factor(meta_hm$coarse_cell_type)))

    plot_phenotype_markers(
      markers = mk,
      expr_mat = expr_mat,
      meta = meta_hm,
      cell_id_col = "cell_id",
      group_col = group_col,
      score_col = score_col,
      celltype_col = "coarse_cell_type",
      celltype_palette = pal_ct,
      heatmap_type = "global",
      top_n_markers = 20L,
      n_mark_labels = 5L,
      p_adj_threshold = 0.05,
      column_title = "FSHD bulk-derived signature: single-cell marker heatmap"
    )

    if (exists("mk_celltype") && !is.null(mk_celltype) && (nrow(mk_celltype$adverse_markers) > 0 || nrow(mk_celltype$favorable_markers) > 0)) {
      plot_phenotype_markers(
        markers = mk_celltype,
        expr_mat = expr_mat,
        meta = meta_hm,
        cell_id_col = "cell_id",
        group_col = group_col,
        score_col = score_col,
        celltype_col = "cell_type",
        celltype_palette = PhenoMapR::get_celltype_palette(levels(factor(meta_hm$cell_type))),
        heatmap_type = "cell_type_specific",
        top_n_markers = 15L,
        n_mark_labels = 3L,
        p_adj_threshold = 0.05,
        column_title = "Cell-type-specific phenotype markers (CyteTypeR cell_type)"
      )
    }
  }
}
```

## 6. Independent validation cohorts (GEO)

To test whether the **GSE140261** bulk phenotype signature (`ref_fshd`) separates **FSHD** from **normal / control** muscle outside the training set, we follow the validation design described in the Scissor FSHD analysis: three independent GEO series (**[GSE115650](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE115650)**, **[GSE56787](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE56787)**, **[GSE26852](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE26852)**) are fetched with **`GEOquery`**, scored with the same **`ref_fshd`** as in §1, and summarized below (**violin plot**, **Wilcoxon *p*-values**, and a **sample-count table**) when downloads and metadata alignment succeed. Cohorts are processed **independently** so one series failing does not block the others.

- **GSE115650** includes **34** FSHD subjects and **nine** control muscle biopsies. Supplementary **`GSE115650_biopsy_FPKM.csv.gz`** provides **FPKM**; we convert **FPKM → TPM** per sample (divide each sample’s FPKM by the sum of FPKMs and scale to \(10^6\)), then map **Ensembl** IDs to **gene symbols** (mean TPM when several Ensembl IDs map to the same symbol), matching §1.
- **GSE56787** contains **24** muscle biopsies (**15** FSHD, **nine** control) plus cultured **myotube / myoblast** samples. We use **`GSE56787_biopsy.gene.counts.csv.gz`** (biopsy counts only) and assign **FSHD** vs **Control** from **`source_name_ch1`**. Counts are **summed** per symbol, then **CPM** (counts per million per sample) is used as library-size–normalized expression for **`PhenoMap()`** (analogous to using proportional abundance for signature weighting).
- **GSE26852** includes FSHD, inflammatory myopathies, dysferlinopathy, and controls. We keep **normal muscle** and **FSHD** samples only (`characteristics_ch1.3`: “Normal Muscle” or “FSHD T2-STIR”), discarding other disease groups. The series matrix reports **Illumina** probe IDs; we collapse **`fData` `Symbol`** to **one row per symbol** (mean intensity), then **scale each sample** to sum \(10^6\) so scores are comparable in scale to TPM-like proportions.

In each cohort, samples are scored with **`PhenoMap(expression = genes × samples, reference = ref_fshd, z_score_cutoff = 2)`**. **Higher** scores indicate stronger alignment with genes that were **higher in FSHD** in **GSE140261** (given `binary_positive_reference = "first"` in §1). Group separation is summarized with **Wilcoxon** tests and **violin plots**.

**Dependencies:** **`GEOquery`** (`BiocManager::install("GEOquery")`) and network access to NCBI GEO. If downloads fail, install **`curl`** or fetch supplementary files manually.

```{r validation-helpers}
#' FPKM columns to TPM: TPM_ij = FPKM_ij / sum_i(FPKM_ij) * 1e6
fpkm_to_tpm <- function(fpkm) {
  fpkm <- as.matrix(fpkm)
  cs <- colSums(fpkm, na.rm = TRUE)
  cs[cs <= 0] <- NA_real_
  sweep(fpkm, 2L, cs, "/") * 1e6
}

#' Collapse Ensembl rows to HGNC symbols by mean TPM (duplicate Ensembl → one symbol).
collapse_ensembl_tpm_to_symbol <- function(tpm_ens_rows_samples_cols) {
  ens <- rownames(tpm_ens_rows_samples_cols)
  ens_base <- sub("\\.[0-9]+$", "", ens)
  sym <- AnnotationDbi::mapIds(
    org.Hs.eg.db,
    keys = ens_base,
    column = "SYMBOL",
    keytype = "ENSEMBL",
    multiVals = "first"
  )
  df <- as.data.frame(tpm_ens_rows_samples_cols)
  df$gene_symbol <- sym
  df <- df[!is.na(df$gene_symbol) & nzchar(df$gene_symbol), , drop = FALSE]
  sample_cols <- setdiff(names(df), "gene_symbol")
  agg <- stats::aggregate(
    df[, sample_cols, drop = FALSE],
    by = list(gene_symbol = df$gene_symbol),
    FUN = mean,
    na.rm = TRUE
  )
  mat <- as.matrix(agg[, -1, drop = FALSE])
  rownames(mat) <- agg$gene_symbol
  mat
}

#' Collapse Ensembl gene-level counts to symbols by summing (then e.g. CPM).
collapse_ensembl_counts_to_symbol_sum <- function(cnt) {
  ens <- rownames(cnt)
  ens_base <- sub("\\.[0-9]+$", "", ens)
  sym <- AnnotationDbi::mapIds(
    org.Hs.eg.db,
    keys = ens_base,
    column = "SYMBOL",
    keytype = "ENSEMBL",
    multiVals = "first"
  )
  df <- as.data.frame(cnt)
  df$gene_symbol <- sym
  df <- df[!is.na(df$gene_symbol) & nzchar(df$gene_symbol), , drop = FALSE]
  sample_cols <- setdiff(names(df), "gene_symbol")
  agg <- stats::aggregate(
    df[, sample_cols, drop = FALSE],
    by = list(gene_symbol = df$gene_symbol),
    FUN = sum,
    na.rm = TRUE
  )
  mat <- as.matrix(agg[, -1, drop = FALSE])
  rownames(mat) <- agg$gene_symbol
  mat
}
```

```{r validation-geo-cohorts, fig.width = 9, fig.height = 3.5}
validation_plot_df <- NULL
validation_wilcox <- list()

if (!vignette_run) {
  message("Section 6 skipped (PHENOMAPR_VIGNETTE_RUN is false).")
} else if (!exists("ref_fshd")) {
  message("Section 6 skipped: ref_fshd not found (bulk reference chunk must run first).")
} else if (!requireNamespace("GEOquery", quietly = TRUE)) {
  message("Install package 'GEOquery' to run Section 6 validation (BiocManager::install(\"GEOquery\")).")
} else {
  val_base <- tempdir()
  val_score_col <- paste0("weighted_sum_score_", colnames(ref_fshd)[1])
  val_pal <- c(Control = "#FED789FF", FSHD = "#476F84FF")

  pick_first_existing_file <- function(dir, patterns) {
    for (pat in patterns) {
      hits <- list.files(dir, pattern = pat, recursive = TRUE, full.names = TRUE)
      if (length(hits) > 0) return(hits[1])
    }
    NA_character_
  }

  infer_sample_key <- function(pd) {
    if ("geo_accession" %in% names(pd)) return(as.character(pd$geo_accession))
    rownames(pd)
  }

  match_pd_to_expr <- function(expr_colnames, pd) {
    key <- infer_sample_key(pd)
    m <- match(expr_colnames, key)
    if (!anyNA(m)) return(m)
    if ("title" %in% names(pd)) {
      m2 <- match(expr_colnames, as.character(pd$title))
      if (!anyNA(m2)) return(m2)
    }
    rep(NA_integer_, length(expr_colnames))
  }

  collapse_characteristics <- function(pd) {
    ch_cols <- grep("^characteristics", names(pd), value = TRUE)
    if (length(ch_cols) == 0) return(rep("", nrow(pd)))
    apply(pd[, ch_cols, drop = FALSE], 1, function(x) paste(x, collapse = " | "))
  }

  ## --- GSE115650: FPKM -> TPM, Ensembl -> symbol ---------------------------
  tryCatch(
    {
      suppressMessages(
        GEOquery::getGEOSuppFiles("GSE115650", makeDirectory = TRUE, baseDir = val_base)
      )
      dir_115650 <- file.path(val_base, "GSE115650")
      fp_115650 <- pick_first_existing_file(
        dir_115650,
        patterns = c("GSE115650.*FPKM.*csv\\.gz$", "FPKM.*csv\\.gz$")
      )
      if (is.na(fp_115650) || !file.exists(fp_115650)) {
        message("GSE115650: supplementary FPKM file not found after getGEOSuppFiles().")
      } else {
        fpkm_115650 <- read.csv(
          gzfile(fp_115650),
          row.names = 1,
          check.names = FALSE,
          stringsAsFactors = FALSE
        )
        tpm_115650 <- fpkm_to_tpm(fpkm_115650)
        tpm_sym_115650 <- collapse_ensembl_tpm_to_symbol(tpm_115650)
        g_115650 <- suppressMessages(GEOquery::getGEO("GSE115650", GSEMatrix = TRUE)[[1]])
        pd_115650 <- Biobase::pData(g_115650)
        meta_115650 <- tolower(collapse_characteristics(pd_115650))
        is_fshd <- grepl("fshd|facioscapulohumeral", meta_115650)
        is_ctrl <- grepl("control|normal|healthy", meta_115650)
        pd_115650$group <- NA_character_
        pd_115650$group[is_ctrl] <- "Control"
        pd_115650$group[is_fshd] <- "FSHD"

        m <- match_pd_to_expr(colnames(tpm_sym_115650), pd_115650)
        if (anyNA(m)) {
          message("GSE115650: could not match expression columns to GEO sample metadata (GSM or title).")
        } else {
          pd_115650 <- pd_115650[m, , drop = FALSE]
          pd_115650$group <- factor(pd_115650$group, levels = c("Control", "FSHD"))
          colnames(tpm_sym_115650) <- infer_sample_key(pd_115650)
          sc_115650 <- PhenoMap(
            expression = tpm_sym_115650,
            reference = ref_fshd,
            z_score_cutoff = 2,
            verbose = FALSE
          )
          x <- sc_115650[colnames(tpm_sym_115650), val_score_col, drop = TRUE]
          ok_w <- stats::complete.cases(x, pd_115650$group)
          if (sum(ok_w) > 0L && length(unique(pd_115650$group[ok_w])) >= 2L) {
            validation_wilcox[["GSE115650"]] <- stats::wilcox.test(x[ok_w] ~ pd_115650$group[ok_w])
          }
          validation_plot_df <- rbind(
            validation_plot_df,
            data.frame(
              cohort = "GSE115650",
              group = pd_115650$group,
              score = x,
              stringsAsFactors = FALSE
            )
          )
        }
      }
    },
    error = function(e) message("GSE115650 failed: ", conditionMessage(e))
  )

  ## --- GSE56787: biopsy gene counts -> CPM -------------------------------
  tryCatch(
    {
      suppressMessages(
        GEOquery::getGEOSuppFiles("GSE56787", makeDirectory = TRUE, baseDir = val_base)
      )
      dir_56787 <- file.path(val_base, "GSE56787")
      fp_56787 <- pick_first_existing_file(
        dir_56787,
        patterns = c("GSE56787.*biopsy.*gene\\.counts.*csv\\.gz$", "biopsy.*gene\\.counts.*csv\\.gz$")
      )
      if (is.na(fp_56787) || !file.exists(fp_56787)) {
        message("GSE56787: supplementary counts file not found after getGEOSuppFiles().")
      } else {
        cnt_56787 <- read.csv(
          gzfile(fp_56787),
          row.names = 1,
          check.names = FALSE,
          stringsAsFactors = FALSE
        )
        cnt_sym_56787 <- collapse_ensembl_counts_to_symbol_sum(cnt_56787)
        lib <- colSums(cnt_sym_56787)
        cpm_56787 <- sweep(cnt_sym_56787, 2, lib, "/") * 1e6
        g_56787 <- suppressMessages(GEOquery::getGEO("GSE56787", GSEMatrix = TRUE)[[1]])
        pd_56787 <- Biobase::pData(g_56787)
        meta_56787 <- collapse_characteristics(pd_56787)
        biopsy <- grepl("muscle biopsy", tolower(meta_56787))
        pd_56787 <- pd_56787[biopsy, , drop = FALSE]
        src <- tolower(as.character(pd_56787$source_name_ch1))
        pd_56787$group <- ifelse(grepl("fshd", src), "FSHD", "Control")
        pd_56787$group <- factor(pd_56787$group, levels = c("Control", "FSHD"))

        m2 <- match_pd_to_expr(colnames(cpm_56787), pd_56787)
        if (anyNA(m2)) {
          message("GSE56787: could not match expression columns to GEO sample metadata (GSM or title).")
        } else {
          pd_56787 <- pd_56787[m2, , drop = FALSE]
          colnames(cpm_56787) <- infer_sample_key(pd_56787)
          sc_56787 <- PhenoMap(
            expression = cpm_56787,
            reference = ref_fshd,
            z_score_cutoff = 2,
            verbose = FALSE
          )
          x2 <- sc_56787[colnames(cpm_56787), val_score_col, drop = TRUE]
          ok_w <- stats::complete.cases(x2, pd_56787$group)
          if (sum(ok_w) > 0L && length(unique(pd_56787$group[ok_w])) >= 2L) {
            validation_wilcox[["GSE56787"]] <- stats::wilcox.test(x2[ok_w] ~ pd_56787$group[ok_w])
          }
          validation_plot_df <- rbind(
            validation_plot_df,
            data.frame(
              cohort = "GSE56787",
              group = pd_56787$group,
              score = x2,
              stringsAsFactors = FALSE
            )
          )
        }
      }
    },
    error = function(e) message("GSE56787 failed: ", conditionMessage(e))
  )

  ## --- GSE26852: Illumina -> symbol, FSHD vs normal ----------------------
  tryCatch(
    {
      g_26852 <- suppressMessages(GEOquery::getGEO("GSE26852", GSEMatrix = TRUE)[[1]])
      pd_26852 <- Biobase::pData(g_26852)
      ds <- pd_26852$characteristics_ch1.3
      keep_26852 <- grepl("Normal Muscle", ds) | grepl("FSHD T2-STIR", ds)
      pd_26852 <- pd_26852[keep_26852, , drop = FALSE]
      pd_26852$group <- ifelse(grepl("FSHD", pd_26852$characteristics_ch1.3), "FSHD", "Control")
      pd_26852$group <- factor(pd_26852$group, levels = c("FSHD", "Control"))
      fd_26852 <- Biobase::fData(g_26852)
      e_26852 <- Biobase::exprs(g_26852)[, rownames(pd_26852), drop = FALSE]
      sym_26852 <- fd_26852$Symbol
      ok_26852 <- !is.na(sym_26852) & nzchar(sym_26852)
      e_26852 <- e_26852[ok_26852, , drop = FALSE]
      sym_26852 <- sym_26852[ok_26852]
      dfl_26852 <- data.frame(sym = sym_26852, e_26852, check.names = FALSE)
      agg_26852 <- stats::aggregate(. ~ sym, data = dfl_26852, mean)
      rownames(agg_26852) <- agg_26852$sym
      mat_26852 <- as.matrix(agg_26852[, -1, drop = FALSE])
      colnames(mat_26852) <- colnames(e_26852)
      mat_26852 <- apply(mat_26852, 2, function(x) x / sum(x) * 1e6)
      sc_26852 <- PhenoMap(
        expression = mat_26852,
        reference = ref_fshd,
        z_score_cutoff = 2,
        verbose = FALSE
      )
      x3 <- sc_26852[rownames(pd_26852), val_score_col, drop = TRUE]
      ok_w <- stats::complete.cases(x3, pd_26852$group)
      if (sum(ok_w) > 0L && length(unique(pd_26852$group[ok_w])) >= 2L) {
        validation_wilcox[["GSE26852"]] <- stats::wilcox.test(x3[ok_w] ~ pd_26852$group[ok_w])
      }
      validation_plot_df <- rbind(
        validation_plot_df,
        data.frame(
          cohort = "GSE26852",
          group = pd_26852$group,
          score = x3,
          stringsAsFactors = FALSE
        )
      )
    },
    error = function(e) message("GSE26852 failed: ", conditionMessage(e))
  )

  if (!is.null(validation_plot_df) && nrow(validation_plot_df) > 0) {
    validation_plot_df$cohort <- factor(
      validation_plot_df$cohort,
      levels = c("GSE115650", "GSE56787", "GSE26852")
    )
    validation_plot_df$score_scaled <- ave(
      validation_plot_df$score,
      validation_plot_df$cohort,
      FUN = function(x) as.numeric(scale(x))
    )
    p_val <- ggplot(validation_plot_df, aes(x = group, y = score_scaled, fill = group)) +
      geom_violin(trim = FALSE, alpha = 0.35, color = NA) +
      geom_jitter(width = 0.08, height = 0, size = 0.9, alpha = 0.85) +
      facet_wrap(~cohort, scales = "free_y", nrow = 1) +
      scale_fill_manual(values = val_pal, drop = FALSE) +
      ggpubr::stat_compare_means(
        aes(group = group),
        method = "wilcox.test",
        comparisons = list(c("Control", "FSHD")),
        label = "p.format",
        label.y.npc = "top"
      ) +
      labs(
        title = "FSHD bulk-derived PhenoMap score: independent GEO cohorts",
        x = NULL,
        y = "Scaled PhenoMap score (within cohort)"
      ) +
      theme_bw() +
      theme(legend.position = "none")
    print(p_val)
  } else {
    message("Independent validation: no cohort produced scores (check GEO downloads and metadata).")
  }
}
```


## Summary

- **Bulk reference** — **GSE140261** TPM, **FSHD vs control**, Ensembl→symbol collapse, **`derive_reference_from_bulk()`**.
- **Single-cell** — **GSE122873**, **<10% MT-** filter, **`PhenoMap()`** + **UMAP** + **`find_phenotype_markers()`** + **`plot_phenotype_markers()`** (global heatmap).
- **Independent validation** — **`GEOquery`** + supplementary tables: **GSE115650** (FPKM→TPM), **GSE56787** (biopsy counts→CPM; myotube/myoblast excluded), **GSE26852** (FSHD vs normal only); same **`ref_fshd`** scored with **`PhenoMap()`**, with **violin plots**, **sample counts**, and **Wilcoxon tests** in Section 6 (cohorts run separately so partial failures still yield results).
- **Relation to Scissor** — Scissor solves a **different** optimization (bulk–single-cell regression on a similarity matrix); **PhenoMapR** uses **gene-level** bulk contrasts and a **weighted sum** score on cells, but the **data sources and QC** can follow the same GEO-based workflow as in [Sun *et al.*](https://www.nature.com/articles/s41587-021-01091-3).

## References

<a id="ref1"></a>
**[1]** Sun, D., Guan, X., Moran, A. E. *et al.* Identifying phenotype-associated subpopulations by integrating bulk and single-cell sequencing data. *Nat. Biotechnol.* 40, 527–538 (2022). [https://doi.org/10.1038/s41587-021-01091-3](https://doi.org/10.1038/s41587-021-01091-3)

<a id="ref2"></a>
**[2]** Kendall, L. E. *et al.* Single-cell RNA sequencing in facioscapulohumeral muscular dystrophy disease etiology and development. *Genome Res.* 29, 890–901 (2019). [https://doi.org/10.1101/gr.243734.118](https://doi.org/10.1101/gr.243734.118)

<a id="ref3"></a>
**[3]** NCBI Gene Expression Omnibus (GEO). **GSE140261**: FSHD and control muscle biopsies (bulk RNA-seq; year 2 processed TPM and metadata used in this vignette). [https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE140261](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE140261)

<a id="ref4"></a>
**[4]** NCBI Gene Expression Omnibus (GEO). **GSE122873**: FSHD single-cell RNA-seq (myocyte cultures; Seurat object used in this vignette derived from this accession). [https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE122873](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE122873)

<a id="ref5"></a>
**[5]** NCBI Gene Expression Omnibus (GEO). **GSE115650**: independent validation cohort (muscle biopsies; supplementary biopsy FPKM table used here). [https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE115650](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE115650)

<a id="ref6"></a>
**[6]** NCBI Gene Expression Omnibus (GEO). **GSE56787**: independent validation cohort (muscle biopsy samples; supplementary biopsy gene counts table used here; cultured samples excluded). [https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE56787](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE56787)

<a id="ref7"></a>
**[7]** NCBI Gene Expression Omnibus (GEO). **GSE26852**: independent validation cohort (FSHD vs normal muscle subset used here). [https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE26852](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE26852)

## Session info

```{r session-info}
sessionInfo()
```
